---
title: "Cancer Subtype Discovery with Consensus NMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cancer Subtype Discovery with Consensus NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  eval = NOT_CRAN
)
```

## The Problem: Robust Classification from Noisy Data

Cancer is not a single disease — molecular profiling reveals **subtypes** with
distinct biology, prognosis, and treatment response. NMF has been a workhorse for
cancer subtype discovery since Brunet et al. (2004) showed that it could
distinguish acute leukemia subtypes from gene expression alone.

But a single NMF run is sensitive to initialization. Different random seeds produce
different factorizations, raising the question: **which groupings are robust and
which are artifacts of initialization?**

**Consensus NMF** answers this by running many NMF replicates with different
initializations and measuring how consistently pairs of samples cluster together.
If two samples are always assigned to the same factor across 50 runs, their
co-clustering is robust. If they only co-cluster in half the runs, the grouping
is unstable.

## The Golub Leukemia Dataset

The `golub` dataset is the classic benchmark from Golub et al. (1999) — gene
expression profiles from 38 leukemia patients:

```{r load-golub}
library(RcppML)
library(ggplot2)
library(Matrix)

data(golub)
cancer_type <- attr(golub, "cancer_type")

cat("Dimensions:", nrow(golub), "samples ×", ncol(golub), "genes\n")
cat("Cancer types:", table(cancer_type), "\n")
cat("True rank: 2 (ALL vs AML)\n")
cat("Expected optimal: ~3 (ALL has subtypes)\n")
```

The ground truth has two major classes — Acute Lymphoblastic Leukemia (ALL, 27
patients) and Acute Myeloid Leukemia (AML, 11 patients) — but ALL itself contains
subtypes that may warrant a third factor.

## Consensus NMF for Rank Selection

The cophenetic correlation coefficient measures how well the consensus matrix
preserves hierarchical clustering structure. A high cophenetic correlation indicates
stable cluster assignments across replicates. As $k$ increases, the cophenetic
correlation typically decreases — the rank with the highest cophenetic value
represents the most robust number of subtypes:

```{r consensus-sweep, fig.width=7, fig.height=5}
# Sweep k = 2 to 6, each with 20 replicates
coph_scores <- numeric(5)
consensus_models <- list()

for (k in 2:6) {
  con <- consensus_nmf(golub, k = k, reps = 20, method = "hard",
                       seed = 42, verbose = FALSE)
  coph_scores[k - 1] <- con$cophenetic
  consensus_models[[k - 1]] <- con
  cat(sprintf("k = %d: cophenetic = %.3f\n", k, con$cophenetic))
}

# Plot cophenetic correlation vs rank
coph_df <- data.frame(k = 2:6, cophenetic = coph_scores)
ggplot(coph_df, aes(x = k, y = cophenetic)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = coph_df$k[which.max(coph_df$cophenetic)],
             linetype = "dashed", color = "red") +
  labs(title = "Consensus Cophenetic Correlation vs. Rank",
       subtitle = "Highest cophenetic indicates most robust clustering",
       x = "Number of Factors (k)", y = "Cophenetic Correlation") +
  ylim(0, 1) +
  theme_minimal()
```

### Consensus Matrix at Optimal Rank

```{r consensus-heatmap, fig.width=6, fig.height=5}
# Use the best k
best_idx <- which.max(coph_scores)
best_con <- consensus_models[[best_idx]]
cat("Optimal k:", best_idx + 1, "\n")

# Visualize consensus matrix
plot(best_con, show_clusters = TRUE) +
  ggtitle(paste0("Consensus Matrix (k = ", best_idx + 1, ", 20 replicates)"))
```

A perfect consensus matrix shows solid diagonal blocks (within-cluster agreement
= 1) and off-diagonal zeros (between-cluster agreement = 0). For this classic
leukemia dataset, k=2 achieves near-perfect consensus, confirming the strong
ALL vs. AML separation.

```{r consensus-summary}
summary(best_con)
```

### Comparing Consensus Methods

RcppML offers two consensus methods:

- **"hard"**: Traditional connectivity-based consensus (Brunet et al. 2004). Each
  sample is assigned to its maximum-loading factor.
- **"knn_jaccard"**: KNN-Jaccard consensus. Instead of hard assignments, measures
  how much samples' k-nearest neighbor sets overlap across replicates. More
  robust to factor rotation.

```{r consensus-knn, fig.width=6, fig.height=5}
con_knn <- consensus_nmf(golub, k = best_idx + 1, reps = 20,
                         method = "knn_jaccard", seed = 42, verbose = FALSE)

cat("Hard consensus cophenetic:      ", best_con$cophenetic, "\n")
cat("KNN-Jaccard consensus cophenetic:", con_knn$cophenetic, "\n")

plot(con_knn, show_clusters = TRUE) +
  ggtitle(paste0("KNN-Jaccard Consensus (k = ", best_idx + 1, ")"))
```

### Validation Against Known Labels

```{r validate-clusters}
# Check if consensus clusters match known cancer types
cluster_labels <- best_con$clusters
ct <- table(Cluster = cluster_labels, Cancer = cancer_type)
cat("Cluster vs. Cancer Type:\n")
print(ct)

# Classification accuracy (best matching)
accuracy <- sum(apply(ct, 2, max)) / sum(ct)
cat(sprintf("\nClassification accuracy: %.1f%%\n", accuracy * 100))
if (accuracy < 1.0) {
  cat("Note: any misclassified samples may represent biological ambiguity\n")
  cat("at the molecular level, not necessarily algorithmic failure.\n")
}
```

## The AML ATAC-seq Dataset: Dense Data Analysis

The `aml` dataset provides a different modality — chromatin accessibility
(ATAC-seq) from 135 AML patient samples plus healthy controls. This is a
**dense** matrix:

```{r load-aml}
data(aml)
aml_meta <- attr(aml, "metadata_h")

cat("Dimensions:", nrow(aml), "regions ×", ncol(aml), "samples\n")
cat("Data type:", class(aml)[1], "\n")
cat("Sparsity:", round(sum(aml == 0) / prod(dim(aml)), 3), "\n")
cat("\nSample categories:\n")
print(table(aml_meta$category))
```

Unlike the sparse gene expression data, ATAC-seq produces nearly dense matrices —
most chromatin regions have measurable accessibility in most samples.

### Robust NMF with Huber Loss

Chromatin accessibility data can contain outlier peaks — extremely high values in a
few samples. The **Huber loss** is ideal here: it behaves like MSE for small errors
but like MAE for large errors, reducing the influence of outliers:

$$L_\delta(r) = \begin{cases} \frac{1}{2}r^2 & \text{if } |r| \le \delta \\ \delta(|r| - \frac{1}{2}\delta) & \text{otherwise} \end{cases}$$

```{r aml-huber}
model_aml_mse <- nmf(aml, k = 6, tol = 1e-5, maxit = 200,
                     loss = "mse", seed = 42, verbose = FALSE)
model_aml_huber <- nmf(aml, k = 6, tol = 1e-5, maxit = 200,
                       loss = "huber", huber_delta = 1.0,
                       seed = 42, verbose = FALSE)

compare_nmf(MSE = model_aml_mse, Huber = model_aml_huber) +
  ggtitle("Training Convergence: MSE vs Huber Loss on AML Data")
```

Note: the convergence curves track each model's own objective, so they are not
directly comparable in magnitude. The key comparison is in the resulting factors.

### Aligning and Comparing Models

When comparing two NMF models, factors may be in different orders. The `align()`
function reorders one model's factors to best match another using the Hungarian
algorithm for optimal bipartite matching:

```{r align-models}
model_aml_huber_aligned <- align(model_aml_huber, ref = model_aml_mse)

# Cosine similarity between aligned factors
cos_sim <- cosine(model_aml_mse@w, model_aml_huber_aligned@w)
cat("Cosine similarity between MSE and Huber factors (after alignment):\n")
print(round(cos_sim, 3))
```

Diagonal values near 1.0 indicate that MSE and Huber loss discover similar biological
factors; lower values suggest the factors have diverged, often because Huber down-weights
outlier accessibility peaks that strongly influence the MSE solution.

### Semi-NMF for Signed Chromatin Changes

Chromatin accessibility changes can go in both directions — regions can become
more or less accessible. Semi-NMF allows $W$ to have negative values, capturing
regions that are specifically *closed* in certain subtypes:

```{r semi-nmf-aml}
model_semi <- nmf(aml, k = 6, nonneg = c(FALSE, TRUE),
                  tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)

cat("Standard W range:", range(model_aml_mse@w), "\n")
cat("Semi-NMF W range:", range(model_semi@w), "\n")
cat("Semi-NMF H range:", range(model_semi@h), "(constrained ≥ 0)\n")
```

Negative values in $W$ represent chromatin regions that are specifically *not*
associated with a factor — a richer biological model than pure non-negative
factorization.

## SparsePress: Efficient Data Storage

For large genomic datasets, storage and I/O are practical concerns. RcppML's
**SparsePress** format provides high-compression storage for sparse matrices
using rANS entropy coding, delta encoding, and value prediction.

SparsePress is most effective on truly sparse data. Let's demonstrate with the
PBMC single-cell dataset:

```{r sparsepress, eval = NOT_CRAN && "pbmc3k" %in% data(package = "RcppML")$results[, "Item"]}
data(pbmc3k)
cat("pbmc3k sparsity:", round(1 - nnzero(pbmc3k) / prod(dim(pbmc3k)), 3), "\n")

# Save as SparsePress format
tmpfile <- tempfile(fileext = ".spz")
sp_write(pbmc3k, tmpfile)

# Check file info without reading
info <- sp_info(tmpfile)
cat("\nSparsePress file info:\n")
for (nm in names(info)) {
  cat(sprintf("  %s: %s\n", nm, info[[nm]]))
}

# Read back and verify
pbmc_reload <- sp_read(tmpfile)
cat("\nReloaded dimensions:", dim(pbmc_reload), "\n")
cat("Values match:", all.equal(as.matrix(pbmc3k), as.matrix(pbmc_reload),
                               check.attributes = FALSE), "\n")
```

### In-Memory Compression

For transmission or caching, the internal `sp_compress()` and `sp_decompress()`
functions work with raw byte vectors:

```{r compress-decompress, eval = FALSE}
raw_bytes <- RcppML:::sp_compress(pbmc3k)
cat("Compressed size:", format(length(raw_bytes), big.mark = ","), "bytes\n")
cat("Original R object size:", format(as.numeric(object.size(pbmc3k)), big.mark = ","), "bytes\n")
cat("Compression ratio:", round(as.numeric(object.size(pbmc3k)) / length(raw_bytes), 1), "x\n")

pbmc_from_raw <- RcppML:::sp_decompress(raw_bytes)
cat("Decompressed matches:", all.equal(as.matrix(pbmc3k), as.matrix(pbmc_from_raw),
                                       check.attributes = FALSE), "\n")
```

The compression ratio depends on data sparsity and value distribution — highly
sparse matrices with low-entropy values achieve the best ratios.

## Bipartite Matching: Optimal Factor Correspondence

The Hungarian algorithm (`bipartiteMatch()`) finds the optimal one-to-one assignment
between factors of two models, minimizing total cost. This is used internally by
`align()`, but is also useful directly:

```{r bipartite-match}
# Cost matrix: 1 - cosine similarity between factor pairs
cost <- 1 - abs(cosine(model_aml_mse@w, model_aml_huber@w))
match_result <- bipartiteMatch(cost)

cat("Optimal assignment (0-indexed):", match_result$assignment, "\n")
cat("Total cost:", round(match_result$cost, 4), "\n")
```

## Putting It Together: AML Subtype Analysis

```{r aml-subtypes, fig.width=7, fig.height=5}
# For consensus clustering of *samples*, we transpose so samples are rows
aml_t <- t(aml)
con_aml <- consensus_nmf(aml_t, k = 6, reps = 15, method = "hard",
                         seed = 42, verbose = FALSE)

cat("AML consensus cophenetic:", con_aml$cophenetic, "\n")

# Compare discovered clusters with known subtypes
ct_aml <- table(
  Cluster = con_aml$clusters,
  Subtype = aml_meta$category
)
cat("\nConsensus clusters vs. AML subtypes:\n")
print(ct_aml)
```

```{r aml-consensus-plot, fig.width=6, fig.height=5}
plot(con_aml, show_clusters = TRUE) +
  ggtitle("AML Sample Consensus Matrix (k = 6)")
```

## Key Takeaways

1. **Consensus NMF** provides robustness that single runs cannot: by aggregating
   many random initializations, it identifies groupings that are stable features
   of the data, not artifacts of optimization.

2. **Cophenetic correlation** offers a principled metric for rank selection in
   consensus NMF — complementary to cross-validation.

3. **KNN-Jaccard consensus** is more robust than hard-assignment consensus,
   especially when factors are not well-separated, because it measures
   neighborhood structure rather than discrete assignments.

4. **Huber loss** is ideal for genomic data with outliers — it provides MSE-like
   behavior for typical observations while limiting the influence of extreme values.

5. **Semi-NMF** captures bidirectional changes (increase/decrease in accessibility),
   producing a richer biological model than standard non-negative factorization.

6. **Factor alignment** via the Hungarian algorithm (`align()`, `bipartiteMatch()`)
   enables principled comparison between models — essential when comparing loss
   functions, regularization strategies, or biological conditions.

7. **SparsePress** provides efficient storage for large sparse genomic matrices,
   with compression ratios that depend on sparsity and value entropy.

## References

- Brunet, J.P., Tamayo, P., Golub, T.R., & Mesirov, J.P. (2004). Metagenes and
  molecular pattern discovery using matrix factorization. *PNAS*, 101(12), 4164-4169.

- Golub, T.R., et al. (1999). Molecular classification of cancer: class discovery
  and class prediction by gene expression monitoring. *Science*, 286(5439), 531-537.

- Lee, D.D. & Seung, H.S. (1999). Learning the parts of objects by non-negative
  matrix factorization. *Nature*, 401(6755), 788-791.

## Further Reading

- [Getting started with RcppML](getting-started.html) — Core NMF concepts
- [Cross-validation for rank selection](cross-validation.html) — Detailed CV guide
- [Single-cell genomics](single-cell-genomics.html) — Cell type discovery with NMF
- [Image decomposition](image-decomposition.html) — Parts-based representations
- [Ecological biodiversity](ecological-biodiversity.html) — Community ecology with NMF
