---
title: "Clustering, Consensus, and Classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering, Consensus, and Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## Why NMF for Clustering?

NMF is inherently a soft clustering method. Each column of H gives a sample's "membership" across k factors, and taking the argmax over rows yields a hard cluster assignment. However, NMF is non-convex — different random initializations produce different solutions and potentially different clusters. **Consensus NMF** addresses this by running many replicates, tracking which samples consistently co-cluster, and building a robust consensus matrix.

Beyond flat clustering, RcppML provides **divisive hierarchical clustering** (`dclust`) via recursive rank-2 NMF splits. The core idea is that **rank-2 NMF is binary clustering**: each split partitions samples into two groups based on which of the two factors dominates. `dclust` applies this recursively, building a full hierarchy — each node in the tree is a `bipartition()` that finds the best binary split. Classification from embeddings leverages the low-dimensional H representation for supervised tasks.

## API Reference

### Consensus NMF

```r
consensus_nmf(data, k, reps = 50, method = c("hard", "knn_jaccard"),
              knn = 10, seed = NULL, threads = 0, verbose = FALSE, ...)
```

- `data` — input matrix (features × samples) or `.spz` path
- `k` — factorization rank (number of clusters)
- `reps` — number of NMF replicates (more = more stable; use ≤ 30 for speed)
- `method` — `"hard"` assigns clusters by dominant factor; `"knn_jaccard"` uses k-NN Jaccard overlap (more robust to ambiguity)

Returns a `consensus_nmf` object with `$consensus` (n × n co-clustering matrix), `$clusters` (assignments), `$cophenetic` (stability metric), and `$models` (all fitted models).

Use `plot()` for a consensus heatmap and `summary()` for cluster statistics.

### Divisive Clustering

```r
bipartition(data, tol = 1e-5, nonneg = TRUE, ...)
dclust(A, min_samples, min_dist = 0, tol = 1e-5, maxit = 100,
       nonneg = TRUE, seed = NULL, threads = 0, verbose = FALSE)
```

`bipartition()` performs a single rank-2 NMF split. `dclust()` recursively bipartitions until clusters are smaller than `min_samples` or have modularity below `min_dist`.

**Important**: `dclust()` returns 0-indexed `$samples`. Add 1 for R-style indexing.

### Factor Matching

```r
bipartiteMatch(x)
```

Hungarian algorithm for optimal 1:1 factor correspondence given a **cost** (dissimilarity) matrix. Returns 0-indexed `$assignment` and total `$cost`. If using cosine similarity, convert to cost first: `bipartiteMatch(1 - cosine_sim)`.

### Classification from Embeddings

```r
classify_embedding(embedding, labels, test_fraction = 0.2, k = 5L, seed = NULL)
classify_logistic(embedding, labels, test_fraction = 0.2, seed = NULL)
```

Both take a samples × features embedding matrix (e.g., `t(model@h)`) and class labels. Return an `fn_classifier_eval` object with `$accuracy`, `$macro_f1`, `$confusion`, and per-class metrics.

## Theory

The **consensus matrix** $C$ is defined element-wise: $C_{ij}$ is the fraction of replicates where samples $i$ and $j$ are assigned to the same cluster. Perfect clustering produces a binary consensus (0 or 1); noisy or unstable clustering yields intermediate values.

The **cophenetic correlation** measures how well hierarchical clustering of the consensus matrix preserves pairwise distances. Higher values indicate more stable clustering.

**Divisive clustering** recursively applies rank-2 NMF. Each split finds the best binary partition; the process stops when clusters are too small or too homogeneous (low modularity).

**Classification from H**: Columns of H are k-dimensional embeddings of samples. Any classifier (k-NN, logistic regression) applied to these embeddings can separate classes, assuming they are captured by the NMF factors.

## Example 1: Cancer Subtype Discovery (Golub Leukemia)

The Golub leukemia dataset contains 38 bone marrow samples — 27 ALL and 11 AML — measured across 5,000 genes. Can unsupervised NMF discover these known subtypes?

```{r golub-consensus}
data(golub)
labels <- attr(golub, "cancer_type")

# Consensus NMF: samples (rows) x genes (columns)
cons <- consensus_nmf(golub, k = 2, reps = 20, seed = 42, verbose = FALSE)
```

```{r golub-heatmap, fig.width = 6, fig.height = 5}
plot(cons, show_clusters = TRUE) +
  ggtitle("Consensus matrix: Golub leukemia (k = 2, 20 replicates)")
```

The consensus heatmap reveals two tightly co-clustered groups. Let's check how well they correspond to the known ALL/AML labels.

```{r golub-confusion}
# Confusion matrix: consensus clusters vs. true labels
conf <- table(Cluster = paste0("Cluster ", cons$clusters), Cancer = labels)
knitr::kable(conf, caption = "Consensus cluster assignments vs. true cancer type (ALL/AML)")
```

> **Reading the confusion matrix**: Each row is a consensus cluster, each column a true cancer type. A well-separated clustering aligns each row with predominantly one column. Here, Cluster 1 captures nearly all ALL samples and Cluster 2 captures most AML samples. The overall **purity** (fraction of samples assigned to their majority class) is `r round(sum(apply(conf, 1, max)) / sum(conf), 3)`.

```{r golub-metrics}
summary_df <- data.frame(
  Metric = c("Cophenetic correlation", "Cluster 1 size", "Cluster 2 size"),
  Value = c(round(cons$cophenetic, 4),
            sum(cons$clusters == 1),
            sum(cons$clusters == 2))
)
knitr::kable(summary_df, caption = "Consensus NMF summary statistics")
```

Consensus NMF with k = 2 cleanly separates the two leukemia subtypes. The cophenetic correlation confirms high clustering stability across replicates.

### Going Deeper: k = 5 Reveals Substructure

What happens when we push beyond the known two classes? With k = 5, consensus NMF discovers finer-grained substructure within both ALL and AML.

```{r golub-k5}
cons5 <- consensus_nmf(golub, k = 5, reps = 20, seed = 42, verbose = FALSE)
```

```{r golub-k5-heatmap, fig.width = 6, fig.height = 5}
plot(cons5, show_clusters = TRUE) +
  ggtitle("Consensus matrix: Golub leukemia (k = 5, 20 replicates)")
```

```{r golub-k5-confusion}
conf5 <- table(Cluster = paste0("C", cons5$clusters), Cancer = labels)
knitr::kable(conf5, caption = "k = 5 consensus clusters vs. cancer type: substructure within ALL and AML")
```

At k = 5, the consensus matrix reveals that ALL and AML each contain subpopulations — ALL splits into 2–3 subgroups and AML into 2. A cophenetic correlation of `r round(cons5$cophenetic, 3)` confirms stable cluster boundaries even at this finer resolution. This substructure is biologically meaningful: ALL includes B-cell and T-cell subtypes, and AML harbors distinct genetic programs.

## Example 2: Hierarchical Clustering with dclust (AML Methylation)

The AML dataset contains 824 differentially methylated regions (DMRs) with beta-value measurements across 135 samples with known subtype annotations (AML and matched controls at MEP, GMP, and L-MPP progenitor stages). Divisive clustering discovers a hierarchy of subtypes by recursively applying rank-2 NMF splits.

```{r dclust-aml}
data(aml)
meta <- attr(aml, "metadata_h")

clusters <- dclust(aml, min_samples = 10, min_dist = 0.01, seed = 42)

# Build cluster composition table
cluster_info <- do.call(rbind, lapply(clusters, function(cl) {
  sample_idx <- cl$samples + 1L  # 0-indexed to 1-indexed
  categories <- meta$category[sample_idx]
  top_cats <- sort(table(categories), decreasing = TRUE)
  data.frame(
    Cluster = cl$id,
    Size = cl$size,
    `Top Category` = names(top_cats)[1],
    `Top Count` = as.integer(top_cats[1]),
    `Composition` = paste(paste0(names(top_cats)[1:min(3, length(top_cats))], ":",
                                  as.integer(top_cats[1:min(3, length(top_cats))])),
                          collapse = ", "),
    check.names = FALSE
  )
}))

knitr::kable(head(cluster_info, 8), caption = "Divisive clustering of AML methylation data")
```

Divisive clustering discovers a hierarchy of subtypes, with early splits separating biologically distinct cell populations. Each cluster's composition reflects known AML subtype structure.

### Splitting Hierarchy

Cluster labels encode the splitting path as binary strings: each character records whether a node went left ("0") or right ("1"). The root split (depth 0) produces children "0" and "1". Splits at depth 1 subdivide those further: "0" → "00" (left) and "01" (right), and so on. The string length minus one gives the split depth at which each cluster was created.

```{r dclust-tree, fig.height = 6}
plot(clusters, labels = meta$category,
     main = "Divisive Hierarchy with Subtype Composition")
```

The top panel shows the true splitting tree with branch points corresponding to individual `bipartition()` calls. The bottom panel shows the composition of each leaf cluster as a stacked bar, aligned with the tree above. The root split (depth 0) separates biologically distinct cell populations, while deeper splits resolve finer substructure within AML subtypes.

## Example 3: Classification from NMF Embeddings (AML Subtypes)

AML methylation data provides a realistic multi-class classification challenge: 6 subtypes with imbalanced class sizes. The NMF embedding compresses 824 DMRs into k factors; we classify subtypes from this low-dimensional representation.

```{r classify-comparison}
# Fit NMF to AML data (features x samples)
model <- nmf(aml, k = 8, seed = 42, maxit = 100)
embedding <- t(model@h)  # 135 samples x 8 factors
int_labels <- as.integer(factor(meta$category)) - 1L  # 0-indexed

# 20% held-out test set
set.seed(42)
n <- nrow(embedding)
test_idx <- sort(sample(n, floor(n * 0.2)))

knn_eval <- classify_embedding(embedding, int_labels, test_idx = test_idx, k = 5, seed = 42)
log_eval <- classify_logistic(embedding, int_labels, test_idx = test_idx, seed = 42)

comp_table <- data.frame(
  Method = c("k-NN (k=5)", "Logistic regression"),
  Accuracy = c(knn_eval$accuracy, log_eval$accuracy),
  `Macro F1` = c(knn_eval$macro_f1, log_eval$macro_f1),
  check.names = FALSE
)
knitr::kable(comp_table, digits = 3,
             caption = "Classification from NMF embeddings (k = 8, AML subtypes)")
```

```{r classify-confusion}
# k-NN confusion matrix
conf_knn <- knn_eval$confusion
cat_levels <- levels(factor(meta$category))
if (nrow(conf_knn) == length(cat_levels)) {
  rownames(conf_knn) <- cat_levels
  colnames(conf_knn) <- cat_levels
}
knitr::kable(conf_knn, caption = "k-NN confusion matrix (AML subtypes)")
```

Unlike the two-class Golub problem (where NMF achieves near-perfect separation with any classifier), AML's 6 subtypes with imbalanced class sizes are harder to classify from 8 factors. The confusion matrix shows which subtypes are well-resolved and which are confused.

### UMAP of Factor Embeddings

```{r umap-classify, fig.height = 5, eval = requireNamespace("uwot", quietly = TRUE)}
library(uwot)
set.seed(42)
umap_coords <- umap(embedding, n_neighbors = 15, min_dist = 0.3, n_components = 2)

umap_df <- data.frame(
  UMAP1 = umap_coords[, 1],
  UMAP2 = umap_coords[, 2],
  Subtype = meta$category,
  Set = ifelse(seq_len(n) %in% test_idx, "Test", "Train")
)

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Subtype, shape = Set)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(Train = 16, Test = 4)) +
  labs(title = "UMAP of NMF Embeddings (AML Subtypes)",
       subtitle = "8-factor NMF embedding colored by subtype, shaped by train/test split") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The UMAP projection reveals how well separated the subtypes are in the 8-dimensional NMF embedding space. Tight, isolated groups indicate subtypes that will classify well; overlapping groups indicate subtypes that share methylation structure.

## What's Next

- *See the [Factor Graphs](factor-graphs.html) vignette for improving embeddings with `refine()`.*
- *See the [Cross-Validation](cross-validation.html) vignette for choosing the optimal k for clustering.*
- *See the [NMF Fundamentals](nmf-fundamentals.html) vignette for the core NMF API.*
