---
title: "Clustering, Consensus, and Classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering, Consensus, and Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## Why NMF for Clustering?

NMF is inherently a soft clustering method. Each column of H gives a sample's "membership" across k factors, and taking the argmax over rows yields a hard cluster assignment. However, NMF is non-convex — different random initializations produce different solutions and potentially different clusters. **Consensus NMF** addresses this by running many replicates, tracking which samples consistently co-cluster, and building a robust consensus matrix.

Beyond flat clustering, RcppML provides **divisive hierarchical clustering** (`dclust`) via recursive rank-2 NMF splits — each node in the hierarchy is a `bipartition()` that finds the best binary split of the data using rank-2 NMF. Classification from embeddings leverages the low-dimensional H representation for supervised tasks.

## API Reference

### Consensus NMF

```r
consensus_nmf(data, k, reps = 50, method = c("hard", "knn_jaccard"),
              knn = 10, seed = NULL, threads = 0, verbose = FALSE, ...)
```

- `data` — input matrix (features × samples) or `.spz` path
- `k` — factorization rank (number of clusters)
- `reps` — number of NMF replicates (more = more stable; use ≤ 30 for speed)
- `method` — `"hard"` assigns clusters by dominant factor; `"knn_jaccard"` uses k-NN Jaccard overlap (more robust to ambiguity)

Returns a `consensus_nmf` object with `$consensus` (n × n co-clustering matrix), `$clusters` (assignments), `$cophenetic` (stability metric), and `$models` (all fitted models).

Use `plot()` for a consensus heatmap and `summary()` for cluster statistics.

### Divisive Clustering

```r
bipartition(data, tol = 1e-5, nonneg = TRUE, ...)
dclust(A, min_samples, min_dist = 0, tol = 1e-5, maxit = 100,
       nonneg = TRUE, seed = NULL, threads = 0, verbose = FALSE)
```

`bipartition()` performs a single rank-2 NMF split. `dclust()` recursively bipartitions until clusters are smaller than `min_samples` or have modularity below `min_dist`.

**Important**: `dclust()` returns 0-indexed `$samples`. Add 1 for R-style indexing.

### Factor Matching

```r
bipartiteMatch(x)
```

Hungarian algorithm for optimal 1:1 factor correspondence given a **cost** (dissimilarity) matrix. Returns 0-indexed `$assignment` and total `$cost`. If using cosine similarity, convert to cost first: `bipartiteMatch(1 - cosine_sim)`.

### Classification from Embeddings

```r
classify_embedding(embedding, labels, test_fraction = 0.2, k = 5L, seed = NULL)
classify_logistic(embedding, labels, test_fraction = 0.2, seed = NULL)
```

Both take a samples × features embedding matrix (e.g., `t(model@h)`) and class labels. Return an `fn_classifier_eval` object with `$accuracy`, `$macro_f1`, `$confusion`, and per-class metrics.

## Theory

The **consensus matrix** $C$ is defined element-wise: $C_{ij}$ is the fraction of replicates where samples $i$ and $j$ are assigned to the same cluster. Perfect clustering produces a binary consensus (0 or 1); noisy or unstable clustering yields intermediate values.

The **cophenetic correlation** measures how well hierarchical clustering of the consensus matrix preserves pairwise distances. Higher values indicate more stable clustering.

**Divisive clustering** recursively applies rank-2 NMF. Each split finds the best binary partition; the process stops when clusters are too small or too homogeneous (low modularity).

**Classification from H**: Columns of H are k-dimensional embeddings of samples. Any classifier (k-NN, logistic regression) applied to these embeddings can separate classes, assuming they are captured by the NMF factors.

## Example 1: Cancer Subtype Discovery (Golub Leukemia)

The Golub leukemia dataset contains 38 bone marrow samples — 27 ALL and 11 AML — measured across 5,000 genes. Can unsupervised NMF discover these known subtypes?

```{r golub-consensus}
data(golub)
labels <- attr(golub, "cancer_type")

# Consensus NMF: samples (rows) x genes (columns)
cons <- consensus_nmf(golub, k = 2, reps = 20, seed = 42, verbose = FALSE)
```

```{r golub-heatmap, fig.width = 6, fig.height = 5}
plot(cons, show_clusters = TRUE) +
  ggtitle("Consensus matrix: Golub leukemia (k = 2, 20 replicates)")
```

The consensus heatmap reveals two tightly co-clustered groups. Let's check how well they correspond to the known ALL/AML labels.

```{r golub-confusion}
# Confusion matrix: consensus clusters vs. true labels
conf <- table(Cluster = paste0("Cluster ", cons$clusters), Cancer = labels)
knitr::kable(conf, caption = "Consensus cluster assignments vs. true cancer type (ALL/AML)")
```

```{r golub-metrics}
summary_df <- data.frame(
  Metric = c("Cophenetic correlation", "Cluster 1 size", "Cluster 2 size"),
  Value = c(round(cons$cophenetic, 4),
            sum(cons$clusters == 1),
            sum(cons$clusters == 2))
)
knitr::kable(summary_df, caption = "Consensus NMF summary statistics")
```

Consensus NMF with k = 2 cleanly separates the two leukemia subtypes. The cophenetic correlation confirms high clustering stability across replicates.

## Example 2: Hierarchical Clustering with dclust (AML Chromatin)

The AML dataset contains 824 chromatin accessibility regions (ATAC-seq) across 135 samples with known subtype annotations. Divisive clustering discovers a hierarchy of subtypes by recursively applying rank-2 NMF splits.

```{r dclust-aml}
data(aml)
meta <- attr(aml, "metadata_h")

clusters <- dclust(aml, min_samples = 10, min_dist = 0.01, seed = 42)

# Build cluster composition table
cluster_info <- do.call(rbind, lapply(clusters, function(cl) {
  sample_idx <- cl$samples + 1L  # 0-indexed to 1-indexed
  categories <- meta$category[sample_idx]
  top_cats <- sort(table(categories), decreasing = TRUE)
  data.frame(
    Cluster = cl$id,
    Size = cl$size,
    `Top Category` = names(top_cats)[1],
    `Top Count` = as.integer(top_cats[1]),
    `Composition` = paste(paste0(names(top_cats)[1:min(3, length(top_cats))], ":",
                                  as.integer(top_cats[1:min(3, length(top_cats))])),
                          collapse = ", "),
    check.names = FALSE
  )
}))

knitr::kable(head(cluster_info, 8), caption = "Divisive clustering of AML chromatin data")
```

Divisive clustering discovers a hierarchy of subtypes, with early splits separating biologically distinct cell populations. Each cluster's composition reflects known AML subtype structure.

## Example 3: Comparing Classification Methods

Using the Golub NMF model, we compare classification accuracy from k-NN and logistic regression applied to the H embedding.

```{r classify-comparison}
# Fit NMF to get embeddings
model <- nmf(t(golub), k = 3, seed = 42, maxit = 100)
embedding <- t(model@h)  # 38 samples x 3 factors
int_labels <- as.integer(labels) - 1L  # 0-indexed

# Use same test split for fair comparison
set.seed(42)
n <- nrow(embedding)
test_idx <- sort(sample(n, floor(n * 0.3)))

knn_eval <- classify_embedding(embedding, int_labels, test_idx = test_idx, k = 3, seed = 42)
log_eval <- classify_logistic(embedding, int_labels, test_idx = test_idx, seed = 42)

comp_table <- data.frame(
  Method = c("k-NN (k=3)", "Logistic regression"),
  Accuracy = c(knn_eval$accuracy, log_eval$accuracy),
  `Macro F1` = c(knn_eval$macro_f1, log_eval$macro_f1),
  check.names = FALSE
)
knitr::kable(comp_table, digits = 3,
             caption = "Classification from NMF embeddings (k = 3)")
```

Both classifiers achieve high accuracy from just 3 NMF factors, confirming that the low-dimensional embedding captures the ALL/AML distinction effectively. The NMF representation compresses 5,000 genes into a handful of components without losing discriminative power.

## What's Next

- *See the [Factor Graphs](factor-graphs.html) vignette for semi-supervised NMF with `guide_classifier` to improve embeddings with labels.*
- *See the [Cross-Validation](cross-validation.html) vignette for choosing the optimal k for clustering.*
- *See the [NMF Fundamentals](nmf-fundamentals.html) vignette for the core API.*
