---
title: "Cross-Validation for NMF Rank Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-Validation for NMF Rank Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Why Cross-Validation?

The factorization rank $k$ is the most important hyperparameter in NMF. Too few factors underfit (miss real structure), too many overfit (model noise). Cross-validation provides a principled, data-driven way to select $k$.

RcppML uses **speckled holdout**: a random subset of matrix entries are held out during training, and the model's ability to predict these unseen entries is measured as the test loss. The rank that minimizes test loss is the best choice.

## Basic Cross-Validation

Pass a vector of ranks to `k` and set `test_fraction`:

```{r}
library(RcppML)
library(Matrix)

# Simulate data with 5 true factors
sim <- simulateNMF(nrow = 200, ncol = 100, k = 5, noise = 0.1, seed = 42)
A <- sim$A

# Cross-validate ranks 2 through 10
# Use generous patience so higher-rank models have time to converge
cv <- nmf(A, k = 2:10, test_fraction = 0.1, cv_seed = 1,
          maxit = 200, patience = 20, verbose = FALSE)
cv
```

The result is an `nmfCrossValidate` data.frame with columns:

- `k`: factorization rank
- `rep`: replicate index
- `test_mse`: test loss (MSE on held-out entries)
- `train_mse`: training loss
- `best_iter`: iteration at which test loss was minimized
- `total_iter`: total iterations run

### Selecting the Optimal Rank

```{r}
# Find rank with lowest mean test loss
mean_test <- tapply(cv$test_mse, cv$k, mean)
optimal_k <- as.integer(names(which.min(mean_test)))
cat("Optimal rank:", optimal_k, "\n")
cat("True rank:", 5, "\n")
```

### Visualizing Results

```{r fig.width = 7, fig.height = 5}
plot(cv)
```

The plot shows test loss vs. rank, with a dashed line marking the optimal $k$. For synthetic data with well-separated factors, the test loss should decrease as $k$ approaches the true rank, then level off or increase. In practice, the shape depends on the data, noise level, and convergence settings.

## Multi-Seed Replicates

Different random seeds produce different holdout masks and initializations. Using multiple replicates gives more robust rank estimates:

```{r}
# 3 replicates with different CV seeds
cv_multi <- nmf(A, k = 2:10, test_fraction = 0.1, cv_seed = 1:3,
                maxit = 200, patience = 20, verbose = FALSE)

# Each seed appears as a separate replicate
table(cv_multi$rep)
```

```{r fig.width = 7, fig.height = 5}
plot(cv_multi)
```

With multiple replicates, the plot shows individual points for each replicate and a line connecting the means. This reveals the variance across seeds and gives confidence in the rank selection.

## Key Parameters

### `test_fraction`

Controls what fraction of non-zero entries are held out:

- **0.05–0.10**: Standard choice. Enough test entries for stable estimates while preserving training signal.
- **Higher values** (e.g., 0.2): More test entries but less training data; useful for very large matrices.
- **Very low values** (e.g., 0.01): May give noisy test loss estimates.

```{r eval = FALSE}
# Example: different holdout fractions
cv_5pct  <- nmf(A, k = 2:8, test_fraction = 0.05, cv_seed = 1, verbose = FALSE)
cv_10pct <- nmf(A, k = 2:8, test_fraction = 0.10, cv_seed = 1, verbose = FALSE)
cv_20pct <- nmf(A, k = 2:8, test_fraction = 0.20, cv_seed = 1, verbose = FALSE)
```

### `patience`

Early stopping: if test loss doesn't improve for `patience` consecutive iterations, training stops. This prevents wasted computation on overfitting iterations. Setting patience too low can terminate training before the model has fully converged, which may cause higher-rank models to appear worse than they are. A value of 15-20 is recommended:

```{r eval = FALSE}
# Default patience = 5. Increase for better convergence:
cv_patient <- nmf(A, k = 2:10, test_fraction = 0.1, cv_seed = 1,
                  patience = 20, verbose = FALSE)
```

### `mask_zeros`

Controls whether zero entries are included in the test set:

- `mask_zeros = FALSE` (default): all entries (zeros and non-zeros) can be in the test set. Measures full matrix reconstruction.
- `mask_zeros = TRUE`: only non-zero entries are in the test set. More appropriate for recommendation-style data where zeros mean "unobserved" rather than "zero."

```{r eval = FALSE}
# For recommendation data (e.g., MovieLens ratings)
cv_sparse <- nmf(ratings_matrix, k = 2:10, test_fraction = 0.1,
                 mask_zeros = TRUE, cv_seed = 1, verbose = FALSE)
```

## Automatic Rank Selection

Use `k = "auto"` to let RcppML search for the optimal rank automatically:

```{r eval = FALSE}
# Auto rank selection within range [2, 50]
model <- nmf(A, k = "auto", test_fraction = 0.1, cv_seed = 1, verbose = TRUE)
# Returns an nmf model at the best rank found
```

You can customize the search range with `cv_k_range`:

```{r eval = FALSE}
# Search between ranks 2 and 20
model <- nmf(A, k = "auto", test_fraction = 0.1, cv_k_range = c(2, 20),
             cv_seed = 1, verbose = TRUE)
```

## Combining CV with Regularization

Cross-validation works with all regularization options:

```{r}
# CV with L1 regularization
cv_L1 <- nmf(A, k = 2:10, test_fraction = 0.1, cv_seed = 1,
             L1 = c(0.01, 0.01), maxit = 200, patience = 20, verbose = FALSE)

# Compare optimal rank with and without regularization
best_k_plain <- as.integer(names(which.min(tapply(cv$test_mse, cv$k, mean))))
best_k_L1 <- as.integer(names(which.min(tapply(cv_L1$test_mse, cv_L1$k, mean))))
cat("Optimal k (no reg):", best_k_plain, "\n")
cat("Optimal k (L1=0.01):", best_k_L1, "\n")
```

## Fitting the Final Model

After selecting the optimal rank, fit a final model on **all** data (no holdout):

```{r}
# Final model at the selected rank
final_model <- nmf(A, k = optimal_k, seed = 42, verbose = FALSE)
final_model
```

## Practical Recommendations

1. **Start with `test_fraction = 0.1` and `cv_seed = 1:3`** for a quick initial scan.
2. **Increase replicates** (e.g., `cv_seed = 1:10`) for publication-quality rank selection.
3. **Use `mask_zeros = TRUE`** for recommendation or count data where zeros are missing values.
4. **Check `best_iter`**: if it equals `total_iter`, increase `maxit`—the model may not have converged.
5. **Look for the "elbow"**: the optimal rank is where test loss levels off, not necessarily the absolute minimum (which can be noisy for high ranks).

## Summary

| Parameter | Default | Purpose |
|:----------|:--------|:--------|
| `k` | (required) | Vector of ranks to evaluate (e.g., `2:10`) |
| `test_fraction` | 0 | Fraction of entries held out (set > 0 to enable CV) |
| `cv_seed` | `NULL` | Seed(s) for holdout mask; vector for replicates |
| `patience` | 5 | Early stopping patience (iterations); consider 15-20 |
| `mask_zeros` | `FALSE` | If TRUE, only non-zeros in test set |
| `track_train_loss` | `TRUE` | Also compute training loss |
