---
title: "Statistical Distributions for NMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Distributions for NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## Motivation

Standard NMF minimizes mean squared error, implicitly assuming Gaussian noise with constant variance. This assumption fails for many real datasets: count data (gene expression, species surveys) has variance proportional to the mean; heavy-tailed data has extreme outliers that dominate the MSE objective; and some data has excess zeros far beyond what any single distribution predicts (e.g., single-cell RNA-seq dropout).

RcppML unifies distribution-specific NMF via Iteratively Reweighted Least Squares (IRLS): at each NMF iteration, the least squares subproblem is re-weighted to match the chosen distribution's variance function. The result is better fit, more interpretable factors, and statistically principled modeling.

## API Reference

### Distribution Selection in `nmf()`

Use the `loss` parameter to specify the error distribution:

```r
nmf(data, k, loss = "gp", ...)
```

| Distribution | `loss =` | $V(\mu)$ | Use Case |
|---|---|---|---|
| Gaussian | `"mse"` | constant | Dense continuous data |
| Generalized Poisson | `"gp"` | $\mu + \theta\mu^2$ | Overdispersed counts |
| Negative Binomial | `"nb"` | $\mu + \mu^2/r$ | Standard count data |
| Gamma | `"gamma"` | $\mu^2$ | Positive continuous data |
| Inverse Gaussian | `"inverse_gaussian"` | $\mu^3$ | Heavy right tails |
| Tweedie | `"tweedie"` | $\mu^p$ | Hybrid count/continuous |

### Dispersion Control

The `dispersion` parameter controls how dispersion is estimated:

| Value | Description |
|-------|-------------|
| `"per_row"` | One dispersion parameter per feature (default) |
| `"per_col"` | One per sample |
| `"global"` | Single global dispersion |
| `"none"` | No dispersion estimation |

### Zero-Inflation

For data with excess zeros beyond what the chosen distribution predicts:

```r
nmf(data, k, loss = "gp", zi = "row", ...)
```

| `zi =` | Description |
|--------|-------------|
| `"none"` | No zero-inflation modeling (default) |
| `"row"` | Per-row (per-feature) zero-inflation probability |
| `"col"` | Per-column (per-sample) zero-inflation probability |

### Diagnostic Functions

- `auto_nmf_distribution(data, k, distributions, criterion)` — fit multiple distributions, compare via AIC/BIC
- `score_test_distribution(data, model, powers)` — score test for variance power without refitting
- `diagnose_zero_inflation(data, model, threshold)` — test for excess zeros
- `diagnose_dispersion(data, model)` — recommend dispersion granularity

## Theory

### Variance-Mean Relationship

Each distribution assumes a specific relationship between the variance and the mean: $V(\mu) = \mu^p$. Gaussian (p=0) has constant variance. Poisson-family (p=1) has variance proportional to mean. Gamma (p=2) has variance proportional to mean squared. The correct assumption determines how residuals are weighted — high-mean entries get downweighted for count data, matching the natural heteroscedasticity.

### IRLS

At each NMF iteration, IRLS computes weights $w_{ij} = 1/V(\hat{\mu}_{ij})$ and solves the weighted NNLS problem. This iterative reweighting converges to the maximum likelihood estimate for the chosen distribution.

### Dispersion

Dispersion controls how much the observed variance deviates from the base variance function. For Negative Binomial, the dispersion parameter $\theta$ controls overdispersion: $V(\mu) = \mu + \mu^2/\theta$. For Generalized Poisson, $\phi$ controls extra-Poisson variation.

The three dispersion modes control granularity:

- **`per_row`** (default): Each feature (row) gets its own dispersion $\phi_i$. Appropriate when features have different noise levels — e.g., highly-expressed genes have different variance structure than lowly-expressed genes.
- **`per_col`**: Each sample (column) gets its own dispersion $\phi_j$. Appropriate when samples have different quality — e.g., cells with different sequencing depth or capture efficiency.
- **`global`**: A single scalar dispersion $\phi$ for the entire matrix. Appropriate for homogeneous data where all features and samples have similar noise characteristics.

**Guidance**: For gene expression data, `per_row` is typically best (genes vary enormously in expression level and noise). For batch-variable data, `per_col` captures sample-level quality differences. Use `diagnose_dispersion()` to choose empirically.

### Zero-Inflation

The ZI mixture model decomposes each observation as: $P(X=0) = \pi + (1-\pi) \cdot f(0|\mu)$. The EM algorithm alternates between estimating zero-inflation probabilities $\pi$ and updating NMF factors — capturing dropout or structural zeros that the base distribution cannot explain.

### Recommended Workflow

When modeling complex data, build up complexity incrementally rather than estimating everything simultaneously:

1. **Select distribution**: Use `auto_nmf_distribution()` or `score_test_distribution()` to find the best variance function.
2. **Diagnose dispersion**: Use `diagnose_dispersion()` to determine whether per-row, per-col, or global dispersion is needed.
3. **Test for zero-inflation**: Use `diagnose_zero_inflation()` to check for excess zeros.
4. **Combine**: Fit the final model with the chosen distribution, dispersion mode, and ZI setting.

### Information Criteria

Distribution comparison uses standard model selection criteria:

- **NLL** (Negative Log-Likelihood): The negative log-likelihood of the data under the fitted model. Lower = better fit.
- **AIC** (Akaike Information Criterion): $\text{AIC} = 2 \cdot \text{NLL} + 2k$, where $k$ is the number of parameters. Penalizes model complexity lightly.
- **BIC** (Bayesian Information Criterion): $\text{BIC} = 2 \cdot \text{NLL} + k \log(N)$, where $N$ is the number of observations. Penalizes complexity more strongly than AIC, preferring simpler models. **BIC is generally preferred** for distribution selection.

Note: NLL and AIC/BIC values can be negative — this reflects the scale of the log-likelihood, not an error.

## Worked Examples

### Example 1: Distribution Auto-Selection on Count Data

The `hawaiibirds` dataset contains species frequency counts from Hawaiian bird surveys — overdispersed count data where Gaussian assumptions are inappropriate.

```{r auto-selection}
data(hawaiibirds)
result <- auto_nmf_distribution(hawaiibirds, k = 8,
                                 distributions = c("mse", "gp", "nb"),
                                 criterion = "bic", seed = 42,
                                 maxit = 30)
```

```{r auto-selection-table}
comp <- result$comparison
comp_display <- data.frame(
  Distribution = comp$distribution,
  NLL = round(comp$nll, 1),
  df = comp$df,
  AIC = round(comp$aic, 1),
  BIC = round(comp$bic, 1),
  Selected = ifelse(comp$selected, "***", "")
)
knitr::kable(
  comp_display,
  caption = paste0("Distribution comparison on hawaiibirds (BIC criterion). Best: ", result$best, ".")
)
```

```{r auto-selection-plot, fig.height = 3.5}
ggplot(comp, aes(x = distribution, y = bic, fill = selected)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "steelblue"), guide = "none") +
  labs(title = "BIC Comparison Across Distributions (Hawaiian Birds)",
       x = "Distribution", y = "BIC (lower is better)") +
  theme_minimal()
```

BIC selects the best-fitting distribution for the Hawaiian bird count data. The BIC-selected distribution best captures the variance-mean relationship in this ecological count data.

```{r dispersion-diagnostic}
# Diagnose dispersion granularity for the best model
best_model <- result$models[[result$best]]
disp_diag <- diagnose_dispersion(hawaiibirds, best_model)

disp_df <- data.frame(
  Property = c("Recommended mode", "Global φ", "Row CV", "Col CV"),
  Value = c(disp_diag$mode, round(disp_diag$global_phi, 4),
            round(disp_diag$row_cv, 4), round(disp_diag$col_cv, 4))
)
knitr::kable(disp_df, caption = "Dispersion diagnostics for hawaiibirds")
```

### Example 2: Score Test Diagnostics

The score test evaluates the variance-power family without refitting — a fast diagnostic to determine which distribution matches the data's variance structure.

```{r score-test}
model_base <- nmf(hawaiibirds, k = 8, seed = 42, tol = 1e-3, maxit = 30)
scores <- score_test_distribution(hawaiibirds, model_base)
```

```{r score-test-table}
score_df <- scores$scores
score_df$T_stat <- round(score_df$T_stat, 4)
score_df$abs_T <- round(score_df$abs_T, 4)
knitr::kable(
  score_df,
  caption = paste0("Score test results. Best power: p = ", scores$best_power,
                    " (", scores$best_distribution, ").")
)
```

The power with the smallest $|T|$ best matches the observed variance-mean relationship. The score test evaluates the variance power family $V(\mu) = \mu^p$:

- **Power 0** (Gaussian): constant variance — if rejected (large $|T|$), the data has heteroscedastic noise.
- **Power 1** (Poisson/GP): variance proportional to mean — appropriate for count data.
- **Power 2** (Gamma): variance proportional to mean-squared — appropriate for heavy-tailed continuous data.

A negative T-statistic at a given power suggests the model over-estimates variance relative to the data at that power. A positive T-statistic suggests under-estimation. The best power is the one where the T-statistic is closest to zero — the variance assumption matches reality.

```{r nb-diagnostic}
if (!is.null(scores$nb_diagnostic)) {
  nb_msg <- if (scores$nb_diagnostic$overdispersed) {
    "Substantial overdispersion detected (T_NB > 0.1). NB or GP may be preferable to Poisson."
  } else {
    "No strong overdispersion beyond Poisson detected."
  }
}
```

`r if (!is.null(scores$nb_diagnostic)) paste0("**NB overdispersion diagnostic**: T_NB = ", round(scores$nb_diagnostic$T_NB, 4), ". ", nb_msg)`

### Example 3: Zero-Inflation Detection and Modeling

Single-cell RNA-seq data has extreme sparsity with "dropout" zeros beyond what any count distribution predicts. We use `pbmc3k`, a variance-selected subset (8,000 genes × 500 cells) of the full 10x Genomics PBMC 3k dataset.

Dropout is a pervasive artifact of scRNA-seq: low-expression genes often fail to be captured during library preparation, producing excess zeros that no count distribution alone can explain.

```{r zi-setup}
data(pbmc3k, package = "RcppML")
tmp <- tempfile(fileext = ".spz")
writeBin(pbmc3k, tmp)
counts <- st_read(tmp)

sparsity_pct <- round(100 * (1 - Matrix::nnzero(counts) / prod(dim(counts))), 1)
cat("Dimensions:", nrow(counts), "×", ncol(counts), "\n")
```

The dataset has `r sparsity_pct`% zeros — far more than any single count distribution can explain.

```{r zi-diagnostic}
model_gp <- nmf(counts, k = 8, loss = "gp", seed = 42, tol = 1e-3, maxit = 30)
zi_diag <- diagnose_zero_inflation(counts, model_gp)
```

```{r zi-table}
zi_summary <- data.frame(
  Metric = c("Excess Zero Rate", "Zero-Inflation Detected", "Recommended ZI Mode"),
  Value = c(
    round(zi_diag$excess_zero_rate, 4),
    as.character(zi_diag$has_zi),
    zi_diag$zi_mode
  )
)
knitr::kable(zi_summary, caption = "Zero-inflation diagnostics on pbmc3k (8,000 genes × 500 cells).")
```

```{r zi-refit}
if (zi_diag$has_zi && zi_diag$zi_mode != "none") {
  model_zi <- nmf(counts, k = 8, loss = "gp", zi = zi_diag$zi_mode,
                  seed = 42, tol = 1e-3, maxit = 30)
  
  loss_gp <- evaluate(model_gp, counts, loss = "mse")
  loss_zi <- evaluate(model_zi, counts, loss = "mse")
  improvement <- round(100 * (1 - loss_zi / loss_gp), 1)
  
  loss_comp <- data.frame(
    Model = c("GP", paste0("GP + ZI (", zi_diag$zi_mode, ")")),
    `MSE` = round(c(loss_gp, loss_zi), 4),
    check.names = FALSE
  )
  knitr::kable(loss_comp, caption = "Reconstruction error: GP vs. GP with zero-inflation.")
}
```

`r if (exists("improvement")) paste0("Adding ", zi_diag$zi_mode, "-wise zero-inflation reduces reconstruction error by ", improvement, "%, confirming that the excess zeros in single-cell data are structural (gene dropout) rather than sampling noise.")`

`r if (exists("improvement") && improvement <= 0) "Note: The improvement is minimal on this subset, suggesting the GP distribution already handles the zero pattern well at this scale."`

## Next Steps

- **Rank selection with distributions**: Cross-validate with `loss = "gp"` or `"nb"`. See the [Cross-Validation](cross-validation.html) vignette.
- **Factor interpretation**: Combine distribution-aware NMF with consensus clustering. See the Clustering vignette.
- **Core NMF mechanics**: For Gaussian/MSE NMF fundamentals, see [NMF Fundamentals](nmf-fundamentals.html).
