---
title: "Distribution API: Statistical Modeling for NMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distribution API: Statistical Modeling for NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = FALSE
)
```

## Overview

Standard NMF minimizes **mean squared error** (MSE), which assumes Gaussian
noise — that every entry has the same variance regardless of magnitude. This
assumption is wrong for most biological and count data: a gene with mean
expression 1,000 has far more variance than one with mean 10.

RcppML's Distribution API lets you match the **variance-mean relationship** to
your data by choosing from a family of distributions. This is implemented
through Iteratively Reweighted Least Squares (IRLS), which adaptively
reweights the standard NMF solver so that high-variance entries get less
influence and low-variance entries get more.

The API is organized around three orthogonal dimensions:

| Dimension | Controls | Parameters |
|-----------|----------|------------|
| **Distribution** | Variance-mean relationship $V(\mu)$ | `loss` |
| **Zero-inflation** | Structural dropout modeling | `zero_inflation` |
| **Robustness** | Outlier downweighting | `robust` |

These compose multiplicatively: the final IRLS weight for entry $(i,j)$ is

$$w_{ij} = w_{\text{dist}}(\mu_{ij}, \theta) \times w_{\text{robust}}(r^P_{ij})$$

where $r^P_{ij} = (y_{ij} - \mu_{ij}) / \sqrt{V(\mu_{ij})}$ is the Pearson residual.

## Available Distributions

Each distribution defines a variance function $V(\mu)$ that determines
how noise scales with the predicted value:

| Distribution | `loss=` | $V(\mu)$ | IRLS Weight | Best For |
|---|---|---|---|---|
| Gaussian | `"mse"` | 1 | 1 | Dense continuous (images, spectra) |
| Poisson | `"gp"` + `dispersion="none"` | $\mu$ | $1/\mu$ | Equidispersed counts |
| Gen. Poisson | `"gp"` | $\mu(1+\theta)^2$ | Fisher info | Overdispersed counts |
| Neg. Binomial | `"nb"` | $\mu + \mu^2/r$ | $r/(\mu(r+\mu))$ | scRNA-seq, high overdispersion |
| Gamma | `"gamma"` | $\mu^2$ | $1/\mu^2$ | Positive continuous, CV constant |
| Inv. Gaussian | `"inverse_gaussian"` | $\mu^3$ | $1/\mu^3$ | Heavy-tailed positive |
| Tweedie | `"tweedie"` | $\mu^p$ | $1/\mu^p$ | Custom power $p$ |

### Choosing a Distribution

**Rule of thumb**: Plot the relationship between the mean and variance of your
features (rows). The slope on a log-log scale tells you the power $p$:

```{r choose-distribution, purl = FALSE}
library(RcppML)
library(Matrix)

# Load example data
data(hawaiibirds)

# Compute per-row mean and variance
row_means <- rowMeans(hawaiibirds)
row_vars  <- apply(as.matrix(hawaiibirds), 1, var)

# Log-log plot: slope ≈ p tells you the distribution
plot(log10(row_means + 1), log10(row_vars + 1),
     xlab = "log10(mean)", ylab = "log10(variance)",
     main = "Variance-Mean Relationship", pch = 16, cex = 0.4)

# Slope interpretation:
# p ≈ 0 → Gaussian (flat)
# p ≈ 1 → Poisson/GP (linear)
# p ≈ 2 → Gamma/NB (quadratic)
# p ≈ 3 → Inverse Gaussian (cubic)
```

Or let RcppML decide automatically:

```{r auto-select, purl = FALSE}
# Automatic distribution selection via score test
model <- nmf(data, k = 10, distribution = "auto")
```

## Auto-Selection

Setting `distribution = "auto"` runs a fast score test that does not require
refitting:

1. Fit a quick MSE baseline (≤ 20 iterations)
2. Compute the score test statistic $T_p$ for each power $p \in \{0,1,2,3\}$:
   $$T_p = \frac{1}{N} \sum_{ij} \left( \frac{r_{ij}^2}{\mu_{ij}^p} - 1 \right)$$
3. Select $p^* = \arg\min_p |T_p|$, which maps to the best-fitting distribution
4. For integer data, test for NB overdispersion:
   $T_{\text{NB}} = \text{mean}\!\left( (r_{ij}^2 - \mu_{ij}) / \mu_{ij}^2 \right)$

You can also compare distributions explicitly using AIC/BIC:

```{r auto-compare, purl = FALSE}
result <- auto_nmf_distribution(data, k = 10,
                                 distributions = c("mse", "gp", "nb"),
                                 criterion = "bic", verbose = TRUE)
print(result$comparison)
cat("Best distribution:", result$best, "\n")
```

## Dispersion Parameters

For GP, NB, Gamma, and Inverse Gaussian, there is a **dispersion parameter**
that controls the variance-to-mean ratio. RcppML estimates this parameter
during fitting.

### Dispersion Granularity

The `dispersion` parameter controls the resolution of dispersion estimation:

| `dispersion=` | Meaning | Use When |
|---|---|---|
| `"none"` | No dispersion ($\theta = 0$, pure Poisson) | Equidispersed counts |
| `"global"` | One $\theta$ for entire matrix | Homogeneous overdispersion |
| `"per_row"` (default) | $\theta_i$ per row/feature | Genes have different dispersions |
| `"per_col"` | $\theta_j$ per column/sample | Samples have different quality |

```{r dispersion, purl = FALSE}
# GP with per-row dispersion (default)
model <- nmf(data, k = 10, loss = "gp", dispersion = "per_row")

# Pure Poisson (no dispersion)
model <- nmf(data, k = 10, loss = "gp", dispersion = "none")
```

### GP Dispersion Parameters

| Parameter | Default | Description |
|---|---|---|
| `theta_init` | 0.1 | Starting value for dispersion |
| `theta_max` | 5.0 | Cap to prevent instability |
| `theta_min` | 0.0 | Floor (0 = Poisson) |


These go in the `distribution_config` list:

```{r theta-config, purl = FALSE}
model <- nmf(data, k = 10, loss = "gp",
             distribution_config = list(
               theta_init = 0.05,
               theta_max  = 3.0,

             ))
```

### NB Dispersion Parameters

| Parameter | Default | Description |
|---|---|---|
| `nb_size_init` | 10.0 | Initial inverse-dispersion ($r$) |
| `nb_size_max` | 1e6 | Cap (large $r$ → Poisson) |
| `nb_size_min` | 0.01 | Floor |

```{r nb-config, purl = FALSE}
model <- nmf(data, k = 10, loss = "nb",
             distribution_config = list(
               nb_size_init = 5.0
             ))
```

## Zero-Inflation

Many datasets have **excess zeros** beyond what any count distribution predicts.
RcppML models this as a two-component mixture:

$$P(y_{ij} = 0) = \pi_{ij} + (1 - \pi_{ij}) \cdot P_{\text{dist}}(0 \mid \mu_{ij})$$

where $\pi_{ij}$ is the probability that the zero is **structural** (dropout)
rather than a true biological zero-count.

### Zero-Inflation Modes

| `zero_inflation=` | Model | Parameters | Use When |
|---|---|---|---|
| `"none"` | No ZI | — | Dense data, low sparsity |
| `"row"` | $\pi_i$ per row | $m$ params | Gene-level dropout (e.g., detection limits) |
| `"col"` | $\pi_j$ per column | $n$ params | Sample-level quality (e.g., sequencing depth) |
| `"twoway"` | $\pi_{ij} = 1 - (1-\pi_i)(1-\pi_j)$ | $m + n$ params | Both gene and sample dropout |
| `"auto"` | Diagnose from data | — | Unsure |

```{r zero-inflation, purl = FALSE}
# Gene-level dropout (common for scRNA-seq)
model <- nmf(data, k = 10, loss = "gp", zero_inflation = "row")

# Automatic: diagnoses excess zero pattern
model <- nmf(data, k = 10, loss = "gp", zero_inflation = "auto")
```

Auto-detection compares observed vs expected zeros (under Poisson) per row and
column, choosing the granularity that captures the most excess zero variance.

## Robust Estimation

Outliers can dominate MSE-based NMF. The `robust` parameter applies Huber
downweighting to the **Pearson residuals** — entries that are far from their
prediction (in units of the distribution's standard deviation) get reduced
influence:

$$w_{\text{robust}}(r^P) = \min\!\left(1, \; \frac{\delta}{|r^P|}\right)$$

| `robust=` | Huber $\delta$ | Effect |
|---|---|---|
| `FALSE` (default) | — | No robustness |
| `TRUE` | 1.345 | 95% efficiency under Gaussian, protects against outliers |
| `"mae"` | 1e-4 | Near-MAE behavior (extreme downweighting) |
| Numeric | Custom | Fine-tuned delta |

Robustness composes with any distribution. For example, robust GP downweights
outlier counts that deviate far from the Poisson prediction:

```{r robust, purl = FALSE}
# Robust GP: downweight outlier counts
model <- nmf(data, k = 10, loss = "gp", robust = TRUE)

# Custom Huber delta
model <- nmf(data, k = 10, loss = "gp", robust = 2.0)
```

## IRLS Tuning

The IRLS outer loop re-solves each column with updated weights. The defaults
work well for most data, but you can tune:

| Parameter | Default | Description |
|---|---|---|
| `irls_max_iter` | 5 | Max IRLS re-weighting iterations per column |
| `irls_tol` | 1e-4 | Convergence: max relative change in solution |

```{r irls-tune, purl = FALSE}
# More IRLS iterations for better convergence
model <- nmf(data, k = 10, loss = "gp",
             irls_max_iter = 10, irls_tol = 1e-6)
```

Increasing `irls_max_iter` improves statistical accuracy at the cost of compute
time (each IRLS iteration is roughly as expensive as one CD solve).

## Complete Example: scRNA-seq Analysis

```{r complete-example, purl = FALSE}
library(RcppML)
library(Matrix)

# Load sparse count matrix (genes × cells)
data(pbmc3k)

# Step 1: Auto-select distribution
result <- auto_nmf_distribution(pbmc3k, k = 10, verbose = TRUE)
# Typical output: GP or NB selected (count data)

# Step 2: Full model with selected distribution + ZI + robustness
model <- nmf(pbmc3k, k = 10,
             loss = result$best,       # e.g., "gp"
             dispersion = "per_row",   # gene-specific dispersion
             zero_inflation = "auto",  # detect dropout pattern
             robust = TRUE,            # protect against outlier counts
             maxit = 100,
             verbose = 1)

# Step 3: Cross-validate rank
cv <- nmf(pbmc3k, k = 2:20,
          loss = result$best,
          test_fraction = 0.05,
          verbose = 1)
plot(cv)
```

## When to Use What

| Data Type | Recommended `loss` | `dispersion` | `zero_inflation` | `robust` |
|---|---|---|---|---|
| Dense continuous (images) | `"mse"` | — | `"none"` | `FALSE` |
| Dense positive (spectra) | `"gamma"` | `"global"` | `"none"` | `TRUE` |
| Sparse counts (scRNA-seq) | `"gp"` or `"nb"` | `"per_row"` | `"row"` or `"auto"` | `TRUE` |
| Ratings (MovieLens) | `"mse"` | — | `"none"` | `TRUE` |
| Ecological abundance | `"gp"` | `"per_row"` | `"twoway"` | `FALSE` |
| Text (word counts) | `"gp"` | `"none"` | `"none"` | `FALSE` |

## Mathematical Details

### IRLS Algorithm

For each column $j$ and IRLS iteration:

1. **Reconstruct**: $\hat{\mu}_j = W^T h_j$ (predicted column)
2. **Weight**: $w_i = V(\hat{\mu}_i)^{-1}$ (distribution-specific)
3. **Weighted Gram**: $G_w = W^T \operatorname{diag}(w) W$ (reweighted normal equations)
4. **Weighted RHS**: $b_w = W^T \operatorname{diag}(w) a_j$
5. **Solve**: $h_j \leftarrow \text{NNLS}(G_w, b_w)$ via coordinate descent
6. **Check convergence**: $\max_i |h_j^{(\text{new})}[i] - h_j^{(\text{old})}[i]| / (|h_j^{(\text{old})}[i]| + \epsilon) < \texttt{irls\_tol}$

The Gram computation is $O(mk^2)$ per column per iteration. For $k \ll m$
this is efficient, and the typically 3–5 IRLS iterations add modest overhead
relative to the standard CD solve.

### GP Dispersion Estimation

After each NMF iteration, the GP dispersion parameter $\theta_i$ (per row $i$)
is updated via moment matching on the Pearson residuals:

$$\hat{\theta}_i = \frac{1}{|\mathcal{J}_i|} \sum_{j \in \mathcal{J}_i} \frac{(y_{ij} - \mu_{ij})^2 - \mu_{ij}}{\mu_{ij}^2}$$

where $\mathcal{J}_i$ is the set of columns with nonzero entries in row $i$.


### Score Test for Auto-Selection

The variance function score statistic for power $p$ is:

$$T_p = \frac{1}{N} \sum_{i,j} \left( \frac{(y_{ij} - \hat{\mu}_{ij})^2}{\hat{\mu}_{ij}^p} - 1 \right)$$

Under the true variance model $V(\mu) = \mu^p$, the standardized residuals
$r_{ij}^2 / \mu_{ij}^p$ have mean 1, so $T_p \approx 0$. The power minimizing
$|T_p|$ identifies the best-fitting variance function without refitting.

## Simulation and Recovery

To verify that the distribution model recovers known ground truth:

```{r simulation-recovery, eval = FALSE, purl = FALSE}
# Generate Poisson-distributed data from known factors
set.seed(42)
W_true <- matrix(runif(100 * 5), 100, 5)
H_true <- matrix(runif(5 * 80), 5, 80)
mu <- W_true %*% H_true
A <- matrix(rpois(length(mu), lambda = as.vector(mu)), nrow = 100)

# Recover with GP loss (appropriate for Poisson data)
model <- nmf(A, k = 5, loss = "gp", seed = 42, maxit = 200, verbose = FALSE)
cat("Reconstruction error:", evaluate(A, model), "\n")
```
