---
title: "Ecological Biodiversity Patterns in Hawaiian Bird Communities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ecological Biodiversity Patterns in Hawaiian Bird Communities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  eval = NOT_CRAN
)
```

## Community Ecology and Matrix Factorization

Ecologists often survey species across spatial sampling units, producing a
**sites × species** matrix where each entry records the abundance or frequency of
a species at a site. A fundamental question is: **what latent ecological
communities structure this data?**

NMF decomposes the species-by-site matrix into:

$$A \approx W \cdot \text{diag}(d) \cdot H$$

- $W$ (species × k): **community profiles** — which species co-occur in each
  latent community
- $H$ (k × sites): **site memberships** — the strength of each community at
  each geographic location
- $d$: relative prevalence of each community

This is ecologically meaningful because:

1. **Non-negativity** reflects that species abundances cannot be negative.
2. **Additivity** models the idea that a site's fauna is a mixture of overlapping
   communities (e.g., forest birds + wetland birds + invasive generalists).
3. **Soft clustering** allows a site to belong to multiple communities with
   different weights, capturing ecological gradients.

## Hawaiian Bird Survey Data

The `hawaiibirds` dataset records bird species frequencies across survey grid cells
in the Hawaiian archipelago — an exceptional natural laboratory for studying island
biogeography:

```{r load-data}
library(RcppML)
library(ggplot2)
library(Matrix)

data(hawaiibirds)
meta_h <- attr(hawaiibirds, "metadata_h")  # site metadata
meta_w <- attr(hawaiibirds, "metadata_w")  # species metadata

cat("Dimensions:", nrow(hawaiibirds), "species ×", ncol(hawaiibirds), "sites\n")
cat("Sparsity:", round(1 - nnzero(hawaiibirds) / prod(dim(hawaiibirds)), 3), "\n")
```

### Island Distribution

```{r island-table}
island_counts <- table(meta_h$island)
cat("Sites per island:\n")
print(island_counts)
```

### Species Characteristics

```{r species-table}
cat("\nSpecies by status:\n")
print(table(meta_w$status))
cat("\nSpecies by type:\n")
print(table(meta_w$type))
```

Hawaii hosts both native species (many endemic and threatened) and introduced
species (brought by humans). Understanding how these groups structure bird
communities is critical for conservation.

### Geographic Distribution

```{r geo-map, fig.width=7, fig.height=5}
ggplot(meta_h, aes(x = lng, y = lat, color = island)) +
  geom_point(size = 0.8, alpha = 0.7) +
  labs(title = "Survey Grid Locations Across the Hawaiian Archipelago",
       x = "Longitude", y = "Latitude", color = "Island") +
  theme_minimal() +
  coord_fixed(ratio = 1.2)
```

## Selecting the Number of Communities

Cross-validation identifies how many latent communities best explain the data
without overfitting:

```{r cross-validation, fig.width=7, fig.height=4}
cv <- nmf(hawaiibirds, k = 2:15, test_fraction = 0.2,
          tol = 1e-5, maxit = 100, verbose = FALSE)
plot(cv) + ggtitle("Cross-Validation: Optimal Number of Bird Communities")
```

```{r optimal-k}
best_k <- cv$k[which.min(cv$test_mse)]
cat("Optimal rank:", best_k, "\n")
```

## Standard NMF: Discovering Bird Communities

```{r fit-nmf}
model <- nmf(hawaiibirds, k = best_k, L1 = c(0.1, 0),
             tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)
model
```

### Community Composition by Island

Each factor represents a latent bird community. The `summary()` method shows how
these communities distribute across islands:

```{r summary-island, fig.width=7, fig.height=5}
nmf_summary <- summary(model, group_by = meta_h$island)
plot(nmf_summary) +
  ggtitle("Bird Community Representation by Island") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Island-specific communities reflect Hawaii's biogeographic reality: each island has
a distinct elevation profile, habitat diversity, and invasion history, producing
unique assemblages.

### Native vs. Introduced Species in Each Community

```{r native-introduced, fig.width=7, fig.height=5}
# Summarize by species status
W <- model@w

# Compute mean loading per status for each factor
status_list <- split(seq_len(nrow(W)), meta_w$status)
status_means <- sapply(status_list, function(idx) colMeans(W[idx, , drop = FALSE]))

# Build tidy data frame for plotting
status_long <- data.frame(
  Factor = rep(paste0("F", 1:best_k), ncol(status_means)),
  Status = rep(colnames(status_means), each = best_k),
  Loading = as.vector(status_means)
)

ggplot(status_long, aes(x = Factor, y = Loading, fill = Status)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("introduced" = "coral", "native" = "forestgreen")) +
  labs(title = "Native vs. Introduced Species Dominance per Community",
       y = "Mean Factor Loading", x = "Community Factor") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Communities dominated by native species likely represent remnant natural habitats,
while communities dominated by introduced species indicate areas of high human
impact.

## Spatial Mapping of Communities

A powerful way to interpret NMF results is to visualize the $H$ matrix
geographically — mapping where each community is most strongly represented:

```{r spatial-maps, fig.width=7, fig.height=7}
# Prepare spatial data
H <- model@h
n_factors <- min(best_k, 6)

plot_list <- list()
for (j in 1:n_factors) {
  df <- data.frame(
    lng = meta_h$lng,
    lat = meta_h$lat,
    loading = H[j, ],
    island = meta_h$island
  )
  p <- ggplot(df, aes(x = lng, y = lat, color = loading)) +
    geom_point(size = 0.6, alpha = 0.8) +
    scale_color_gradient(low = "gray90", high = "darkblue") +
    labs(title = paste("Community", j), color = "H") +
    theme_minimal() +
    coord_fixed(ratio = 1.2) +
    theme(legend.position = "bottom",
          axis.title = element_blank(),
          axis.text = element_blank(),
          plot.title = element_text(size = 10))
  plot_list[[j]] <- p
}

# Arrange in grid using base R
gridExtra_available <- requireNamespace("patchwork", quietly = TRUE)
if (gridExtra_available) {
  library(patchwork)
  combined <- Reduce(`+`, plot_list) + plot_layout(ncol = 3)
  print(combined)
} else {
  # Fallback: print individual plots
  for (p in plot_list[1:min(4, n_factors)]) print(p)
}
```

Each panel maps one community across the archipelago. Geographic clustering of
community loadings reveals habitat-linked spatial structure.

## Graph-Regularized NMF: Incorporating Spatial Information

Standard NMF treats each site independently, but nearby sites share ecological
conditions. **Graph Laplacian regularization** adds a penalty that encourages
geographically adjacent sites to have similar community memberships.

We construct a spatial adjacency graph using a distance threshold:

```{r graph-construction}
# Build spatial adjacency graph from coordinates
coords <- cbind(meta_h$lng, meta_h$lat)
n_sites <- nrow(coords)

# Compute pairwise distances (using a subset for efficiency)
# Connect sites within ~0.1 degrees (~11 km)
threshold <- 0.1

# Build sparse adjacency via distance threshold
i_vec <- integer(0)
j_vec <- integer(0)

# Efficient: compare within islands only
for (isl in levels(meta_h$island)) {
  idx <- which(meta_h$island == isl)
  if (length(idx) < 2) next
  d <- as.matrix(dist(coords[idx, ]))
  pairs <- which(d > 0 & d < threshold, arr.ind = TRUE)
  if (nrow(pairs) > 0) {
    i_vec <- c(i_vec, idx[pairs[, 1]])
    j_vec <- c(j_vec, idx[pairs[, 2]])
  }
}

# Create symmetric adjacency matrix
if (length(i_vec) > 0) {
  graph_H <- sparseMatrix(i = i_vec, j = j_vec, x = 1,
                          dims = c(n_sites, n_sites),
                          dimnames = list(NULL, NULL))
  # Symmetrize
  graph_H <- (graph_H + t(graph_H)) / 2
  graph_H <- as(graph_H, "dgCMatrix")
  
  cat("Spatial graph:", nrow(graph_H), "nodes,",
      nnzero(graph_H) / 2, "edges\n")
  cat("Mean degree:", mean(rowSums(graph_H > 0)), "\n")
}
```

```{r graph-nmf}
if (exists("graph_H") && nnzero(graph_H) > 0) {
  model_graph <- nmf(hawaiibirds, k = best_k, L1 = c(0.1, 0),
                     graph_H = graph_H, graph_lambda = c(0, 0.5),
                     tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)
  
  cat("Standard NMF MSE:", evaluate(model, hawaiibirds), "\n")
  cat("Graph NMF MSE:   ", evaluate(model_graph, hawaiibirds), "\n")
}
```

Graph regularization produces smoother spatial patterns — communities transition
gradually across geographic space rather than jumping sharply between adjacent sites.

## Biplot: Species × Community Relationships

A biplot visualizes how species relate to two community axes, colored by their
native/introduced status:

```{r biplot, fig.width=7, fig.height=5}
biplot(model, factors = c(1, 2), matrix = "w",
       group_by = meta_w$status) +
  ggtitle("Species Biplot: Community 1 vs Community 2") +
  scale_color_manual(values = c("introduced" = "coral", "native" = "forestgreen"))
```

Species that cluster together in the biplot tend to co-occur at the same sites —
revealing ecological associations.

## Hierarchical Community Structure with Bipartition and dclust

NMF identifies overlapping communities, but sometimes we want a **hierarchical**
decomposition. The `bipartition()` function splits sites into two groups using
rank-2 NMF:

```{r bipartition}
bp <- bipartition(hawaiibirds, seed = 42)
cat("Bipartition split:", bp$size1, "vs", bp$size2, "sites\n")
cat("Separation distance:", round(bp$dist, 4), "\n")
```

For deeper hierarchy, `dclust()` recursively bipartitions until a stopping
criterion is met:

```{r dclust, fig.width=7, fig.height=4}
dc <- dclust(hawaiibirds, min_samples = 20, seed = 42)

if (length(dc$id) > 0) {
  cat("Divisive clusters found:", length(dc$id), "\n")
  cat("Cluster sizes:", sapply(dc$samples, length), "\n")
  
  # Map clusters to islands
  cluster_labels <- rep(NA, ncol(hawaiibirds))
  for (i in seq_along(dc$samples)) {
    cluster_labels[dc$samples[[i]] + 1] <- i  # 0-indexed
  }
  
  # Cross-tabulate with islands
  if (!all(is.na(cluster_labels))) {
    ct <- table(Cluster = cluster_labels, Island = meta_h$island)
    cat("\nCluster × Island cross-tabulation:\n")
    print(ct)
  }
} else {
  cat("dclust did not find stable subclusters with min_samples = 20.\n")
  cat("The bipartition splitting criterion was not met, suggesting\n")
  cat("the data lacks strong hierarchical structure at this resolution.\n")
  cat("Try adjusting min_samples or min_dist for different granularity.\n")
}
```

If clusters align strongly with islands, it confirms that geography is the dominant
axis of community variation — consistent with island biogeography theory.

## Transposed Analysis: Site Perspective

NMF factors can be interpreted from either perspective. Using `t()`, we can flip
to analyze how sites group in species space:

```{r transpose}
model_t <- t(model)
cat("Transposed model:\n")
cat("  W:", nrow(model_t@w), "×", ncol(model_t@w), "(now sites × k)\n")
cat("  H:", nrow(model_t@h), "×", ncol(model_t@h), "(now k × species)\n")
```

```{r sort-subset}
# Sort factors by importance
model_sorted <- sort(model)
cat("\nFactor importance (d, sorted):\n")
print(round(model_sorted@d, 3))

# Extract top 3 most important factors
model_top3 <- model_sorted[1:3]
cat("\nSubset to top 3 factors:", dim(model_top3), "\n")
```

## Top Species per Community

```{r top-species}
for (j in 1:min(best_k, 6)) {
  top_idx <- order(model@w[, j], decreasing = TRUE)[1:8]
  species_names <- meta_w$species[top_idx]
  status <- meta_w$status[top_idx]
  type <- meta_w$type[top_idx]
  cat(sprintf("\n--- Community %d ---\n", j))
  for (i in seq_along(top_idx)) {
    cat(sprintf("  %s (%s, %s): %.3f\n",
                species_names[i], status[i], type[i], model@w[top_idx[i], j]))
  }
}
```

## Key Takeaways

1. **NMF reveals ecological communities** — latent groupings of co-occurring species
   that map to distinct habitats and geographic regions.

2. **Island biogeography** is confirmed: communities are strongly structured by
   island identity, consistent with the theory that oceanic barriers limit species
   dispersal.

3. **Native vs. introduced species** occupy different factors, reflecting contrasting
   ecological strategies: native species in remnant habitats, introduced species in
   disturbed areas.

4. **Graph Laplacian regularization** incorporates spatial proximity, producing
   ecologically realistic community maps where nearby sites have similar compositions.

5. **Hierarchical analysis** via `bipartition()` and `dclust()` can reveal nested
   community structure — island-level divisions may contain finer habitat-level
   subdivisions when the splitting criterion is met.

6. **Geographic visualization** of $H$ loadings transforms abstract factor values
   into spatially interpretable ecological patterns.

## Further Reading

- [Getting started with RcppML](getting-started.html) — Core NMF concepts
- [Cross-validation for rank selection](cross-validation.html) — Detailed CV guide
- [Building a recommendation engine](recommendation-systems.html) — NMF for sparse data
- [Cancer subtype discovery](cancer-subtype-discovery.html) — Consensus NMF for robust groupings
