---
title: "Factorization Graphs, Guides, and Semi-Supervised Learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factorization Graphs, Guides, and Semi-Supervised Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
library(Matrix)
```

## Why Factorization Graphs?

Real-world problems rarely fit the pattern of "one matrix, one factorization." Multi-modal data (e.g., gene expression + chromatin accessibility) sharing the same samples demands a shared latent space. Prior knowledge — class labels, reference patterns — should steer factors rather than being ignored.

RcppML's `factor_net()` treats factorization as a **directed graph** of composable blocks. Combined with **guides** that inject domain knowledge directly into the optimization, it enables workflows no other NMF package offers: classification-augmented embeddings, semi-supervised learning, multi-modal integration, and automated hyperparameter search.

## API Reference

### Graph Construction

| Function | Purpose |
|----------|---------|
| `factor_input(data, name)` | Input node wrapping a matrix or .spz path |
| `nmf_layer(input, k, ...)` | NMF factorization layer |
| `factor_shared(...)` | Multi-modal: shared H across multiple inputs |
| `factor_concat(...)` | Concatenate H factors from parallel branches |
| `factor_add(...)` | Element-wise H addition (skip connection) |
| `factor_condition(input, Z)` | Condition on external metadata |

### Graph Configuration and Execution

- `factor_config(maxit, tol, loss, test_fraction, ...)` — global settings
- `W(L1, guide, ...)` / `H(L1, guide, ...)` — per-factor overrides for regularization and guides
- `factor_net(inputs, output, config)` — compile the graph
- `fit(net)` — execute; returns a `factor_net_result` with per-layer access via `result$layer_name`

### Guides (Semi-Supervised Steering)

Guides add a soft penalty that biases factors toward a target structure:
$$\min_{h \geq 0} \|Gh - b\|^2 + \lambda\|h - t\|^2$$
where $t$ is the guide target, recomputed each iteration.

| Guide | Target | Use Case |
|-------|--------|----------|
| `guide_classifier(labels, lambda)` | Per-class centroids | Semi-supervised clustering |
| `guide_external(target, lambda)` | Fixed matrix | Transfer learning |
| `guide_callback(fn, lambda)` | Dynamic via `fn(factor, iter)` | Custom strategies |
| `guide_reference(layer_name, lambda)` | Another layer's output | Cross-layer coupling |

For `guide_classifier`: labels must be **0-indexed integers**; use `-1` for unlabeled samples.

### Hyperparameter Search

```r
cross_validate_graph(inputs, layer_fn, params, config, reps = 3,
                     strategy = c("grid", "random"), seed = 42L)
```

- `layer_fn` — function `function(p)` that builds output layer from parameter list `p`
- `params` — named list of vectors to search (e.g., `list(k = c(3, 5, 10))`)
- Returns ranked parameter combinations with mean test loss

## Theory

Each `factor_net` layer performs a factorization. Layers can share factors (multi-modal), be stacked (deep), or branch (multi-task). The engine optimizes all layers jointly.

**Guide mechanics**: `guide_classifier` computes the current centroid of each class from the H matrix every iteration. This biases factors toward class structure without overriding data-driven learning. As `lambda` increases, factors shift from purely data-driven to purely label-driven.

**Semi-supervised**: Labels of `-1` are ignored by the guide. Labeled samples steer the embedding; unlabeled samples follow the data alone.

## Example 1: Graph NMF Equivalence to `nmf()`

The simplest factorization graph is a single input feeding a single NMF layer — identical to calling `nmf()` directly.

```{r graph-equivalence}
data(aml)

# Direct nmf()
direct <- nmf(aml, k = 6, seed = 42, maxit = 100)

# Equivalent factor_net
inp <- factor_input(aml, name = "chromatin")
out <- nmf_layer(inp, k = 6, name = "nmf1")
config <- factor_config(maxit = 100, seed = 42)
net <- factor_net(inp, out, config)
graph_result <- fit(net)

# Compute MSE from graph result factors (loss field reports total, not mean)
W_gr <- graph_result$nmf1$W
H_gr <- graph_result$nmf1$H
d_gr <- graph_result$nmf1$d
mse_graph <- mean((as.matrix(aml) - W_gr %*% diag(d_gr) %*% H_gr)^2)

equiv_table <- data.frame(
  Method = c("nmf()", "factor_net"),
  MSE = c(evaluate(direct, aml), mse_graph),
  Iterations = c(direct@misc$iter, graph_result$nmf1$iterations),
  check.names = FALSE
)
knitr::kable(equiv_table, digits = 5,
             caption = "factor_net produces equivalent results to plain nmf()")
```

The graph API produces identical results. Its value emerges with multi-modal, guided, and hierarchical architectures.

## Example 2: Classification-Augmented Embeddings (Golub)

Guides steer NMF factors toward class structure, producing embeddings that are jointly optimized for reconstruction *and* classification. The trade-off between these objectives becomes visible as guidance strength increases.

```{r guided-golub}
data(golub)
labels_factor <- attr(golub, "cancer_type")
labels <- as.integer(labels_factor) - 1L  # 0 = ALL, 1 = AML

# Lambda sweep: from unsupervised to strongly guided
lambdas <- c(0, 0.1, 0.5, 1.0, 5.0)
models <- lapply(lambdas, function(lam) {
  if (lam == 0) {
    nmf(t(golub), k = 3, seed = 42, maxit = 100)
  } else {
    nmf(t(golub), k = 3, guides = guide_classifier(labels, lambda = lam),
        seed = 42, maxit = 100)
  }
})

# Reconstruction MSE for each
mse_vals <- sapply(models, function(m) evaluate(m, t(golub)))

# Classification accuracy (kNN embedding classifier)
set.seed(42)
test_idx <- sort(sample(38, 11))
acc_vals <- sapply(models, function(m) {
  classify_embedding(t(m@h), labels, test_idx = test_idx, k = 3)$accuracy
})

comp <- data.frame(
  Lambda = lambdas,
  Accuracy = acc_vals,
  `Reconstruction MSE` = mse_vals,
  `MSE Increase (%)` = round(100 * (mse_vals / mse_vals[1] - 1), 1),
  check.names = FALSE
)
knitr::kable(comp, digits = c(1, 3, 0, 1),
             caption = "Guided NMF: reconstruction-classification trade-off across lambda")
```

```{r guided-scatter, fig.width = 7, fig.height = 3.5}
# Scatter plot: Factor 1 vs Factor 2 for unsupervised vs. strongly guided
scatter_data <- rbind(
  data.frame(F1 = models[[1]]@h[1, ], F2 = models[[1]]@h[2, ],
             Cancer = labels_factor, Model = "Unsupervised (λ = 0)"),
  data.frame(F1 = models[[4]]@h[1, ], F2 = models[[4]]@h[2, ],
             Cancer = labels_factor, Model = "Guided (λ = 1.0)")
)

ggplot(scatter_data, aes(x = F1, y = F2, color = Cancer)) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~ Model, scales = "free") +
  theme_minimal() +
  labs(x = "Factor 1", y = "Factor 2",
       title = "NMF embeddings: unsupervised vs. guided")
```

On clean data like Golub, even unsupervised NMF separates the cancer types well. The value of guidance is visible in the factor *structure* — guided factors align more tightly with class boundaries, which matters for downstream interpretation and for noisier datasets where unsupervised separation would fail. The trade-off table shows that strong guidance ($\lambda = 5$) degrades reconstruction substantially.

## Example 3: Semi-Supervised — Partial Labels

When only a fraction of samples have labels, guided NMF propagates label structure through the shared low-dimensional space.

```{r semi-supervised}
# Mask 50% of labels as unlabeled (-1)
partial_labels <- labels
set.seed(42)
partial_labels[sample(length(labels), length(labels) %/% 2)] <- -1L

model_semi <- nmf(t(golub), k = 3,
                  guides = guide_classifier(partial_labels, lambda = 1.0),
                  seed = 42, maxit = 100)

eval_semi <- classify_embedding(t(model_semi@h), labels, test_idx = test_idx, k = 3)
eval_unsup <- classify_embedding(t(models[[1]]@h), labels, test_idx = test_idx, k = 3)
eval_guided <- classify_embedding(t(models[[4]]@h), labels, test_idx = test_idx, k = 3)

semi_table <- data.frame(
  Model = c("Unsupervised", "Semi-supervised (50% labeled)", "Fully guided"),
  Accuracy = c(eval_unsup$accuracy, eval_semi$accuracy, eval_guided$accuracy),
  `Macro F1` = c(eval_unsup$macro_f1, eval_semi$macro_f1, eval_guided$macro_f1),
  check.names = FALSE
)
knitr::kable(semi_table, digits = 3,
             caption = "Semi-supervised NMF with partial labels")
```

Even with only half the samples labeled, guided NMF achieves classification accuracy close to the fully supervised case, demonstrating effective semi-supervised learning from the shared latent space.

## Example 4: Multi-Modal Shared Factorization

When two data modalities share the same samples, `factor_shared()` discovers a common latent space that explains both simultaneously.

```{r multi-modal}
# Simulate two views of the same 100 samples
sim1 <- simulateNMF(200, 100, k = 5, noise = 0.3, seed = 42)
sim2 <- simulateNMF(150, 100, k = 5, noise = 0.3, seed = 43)

inp1 <- factor_input(sim1$A, name = "view1")
inp2 <- factor_input(sim2$A, name = "view2")
shared_inp <- factor_shared(inp1, inp2)
shared <- nmf_layer(shared_inp, k = 5, name = "shared")
config_mm <- factor_config(maxit = 100, seed = 42)
net_mm <- factor_net(list(inp1, inp2), shared, config_mm)
result_mm <- fit(net_mm)

# Compare learned H to true H from each view
H_learned <- result_mm$shared$H
cos_sim <- function(a, b) {
  sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))
}

# Match factors using bipartiteMatch on cosine distance
cost_mat <- matrix(0, 5, 5)
for (i in 1:5) {
  for (j in 1:5) {
    cost_mat[i, j] <- 1 - cos_sim(H_learned[i, ], sim1$h[j, ])
  }
}
match_result <- bipartiteMatch(cost_mat)
matched_sims <- sapply(1:5, function(i) {
  j <- match_result$assignment[i] + 1L  # 0-indexed to 1-indexed
  cos_sim(H_learned[i, ], sim1$h[j, ])
})

modal_table <- data.frame(
  `Learned Factor` = 1:5,
  `Matched True Factor` = match_result$assignment + 1L,
  `Cosine Similarity` = matched_sims,
  check.names = FALSE
)
knitr::kable(modal_table, digits = 3,
             caption = "Shared factorization recovers common latent structure")
```

Shared factorization discovers the common latent structure across two feature spaces, matching the true generating factors with high cosine similarity.

## Example 5: Hyperparameter Search with Cross-Validation

`cross_validate_graph()` automates grid search over rank and regularization strength.

```{r cv-graph}
data(aml)
inp_cv <- factor_input(aml, name = "aml")
layer_fn <- function(p) nmf_layer(inp_cv, k = p$k, L1 = c(p$L1, 0), name = "nmf_cv")
params <- list(k = c(3, 5, 8), L1 = c(0, 0.01, 0.1))
config_cv <- factor_config(test_fraction = 0.1, cv_seed = 42, maxit = 50)

cv <- cross_validate_graph(inp_cv, layer_fn, params, config_cv,
                           reps = 2, seed = 42, verbose = FALSE)

top5 <- head(cv$summary, 5)
knitr::kable(top5, digits = 5,
             caption = "Top 5 parameter combinations by mean test loss")
```

Grid search over 9 parameter combinations identifies the optimal rank and L1 strength, balancing reconstruction accuracy against overfitting.

## What's Next

- *See the [Clustering](clustering.html) vignette for consensus NMF and unsupervised clustering.*
- *See the [Cross-Validation](cross-validation.html) vignette for rank selection without the graph API.*
- *See the [NMF Fundamentals](nmf-fundamentals.html) vignette for the core NMF API.*
