---
title: "Factorization Graphs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factorization Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
library(Matrix)
```

## Why Factorization Graphs?

`nmf()` decomposes one matrix into two factors. That covers a lot of ground --- but real problems often involve richer structure:

- **Multi-modal data**: Two matrices sharing the same samples should share a single latent space, not be factorized independently.
- **Hierarchical decomposition**: Coarse global patterns and fine local patterns live at different ranks --- stacking NMF layers captures both.
- **Guided factorization**: Prior knowledge (class labels, reference patterns) should steer the optimization, not be bolted on afterward.
- **Systematic tuning**: Rank, regularization, and loss function interact in nonobvious ways --- you need a principled way to search over them together.

RcppML's `factor_net()` addresses all of these by treating factorization as a **directed graph** of composable blocks. You wire together inputs, layers, and merge operations, then `fit()` the assembled network.

## Building Blocks

```{r building-blocks-table, echo = FALSE, results = "asis"}
blocks <- data.frame(
  Function = c("`factor_input(data)`",
               "`nmf_layer(input, k)`",
               "`svd_layer(input, k)`",
               "`factor_shared(...)`",
               "`factor_concat(...)`",
               "`factor_add(...)`",
               "`factor_condition(input, Z)`"),
  Role = c("Wraps a dense matrix, sparse matrix, or `.spz` file path as a graph input",
           "NMF layer: *A ≈ W diag(d) H*, non-negative",
           "SVD/PCA layer: signed, optional centering/scaling",
           "Row-binds multiple inputs; one NMF shares *H* across all",
           "Stacks *H* matrices from parallel branches (expands rank)",
           "Element-wise *H* addition (skip connection, same rank)",
           "Appends external metadata *Z* as extra input rows"),
  stringsAsFactors = FALSE
)
knitr::kable(blocks, escape = FALSE)
```

Every graph is compiled with `factor_net(inputs, output, config)` and executed with `fit()`. The result is a `factor_net_result` whose named elements give per-layer access to W, H, d, loss, and convergence info:

```{r usage-pattern, eval = FALSE}
inp <- factor_input(A, name = "data")
out <- nmf_layer(inp, k = 8, name = "my_nmf")
net <- factor_net(inp, out, factor_config(maxit = 100))
result <- fit(net)

result$my_nmf$W   # m x k
result$my_nmf$H   # k x n
result$my_nmf$d   # length-k scaling vector
```

Per-factor overrides for regularization, constraints, and target regularization are set through `W()` and `H()`:

```{r per-factor-example, eval = FALSE}
# Sparse W, with target regularization on H
out <- nmf_layer(inp, k = 8,
                 W = W(L1 = 0.1),
                 H = H(target = T_mat, target_lambda = 0.3))
```

## Example 1: Graph ≡ Plain NMF

The simplest graph is one input feeding one NMF layer --- identical to calling `nmf()`.

```{r equivalence}
data(aml)

# Plain nmf()
set.seed(42)
direct <- nmf(aml, k = 6, seed = 42, maxit = 100)

# Same thing as a factor graph
inp <- factor_input(aml, name = "methylation")
out <- nmf_layer(inp, k = 6, name = "nmf1")
net <- factor_net(inp, out, factor_config(maxit = 100, seed = 42))
graph_result <- fit(net)

# Compare
W_gr <- graph_result$nmf1$W
H_gr <- graph_result$nmf1$H
d_gr <- graph_result$nmf1$d
recon_graph <- W_gr %*% diag(d_gr) %*% H_gr
mse_graph <- mean((as.matrix(aml) - recon_graph)^2)

knitr::kable(
  data.frame(
    Method = c("nmf()", "factor_net"),
    MSE = c(evaluate(direct, aml), mse_graph),
    Iterations = c(direct@misc$iter, graph_result$nmf1$iterations),
    check.names = FALSE
  ),
  digits = 6, caption = "Graph API matches plain nmf() exactly"
)
```

The value of the graph API appears the moment you need something a single `nmf()` call cannot express.

## Example 2: Multi-Modal Shared Factorization

When two matrices share the same $n$ samples, `factor_shared()` finds a common latent space. It vertically concatenates the inputs and fits one NMF:

$$\min_{W_1, W_2, H \geq 0} \left\| \begin{bmatrix} A_1 \\ A_2 \end{bmatrix} - \begin{bmatrix} W_1 \\ W_2 \end{bmatrix} \text{diag}(d)\,H \right\|_F^2$$

The shared $H$ represents structure common to both views. Each $W_p$ captures view-specific loadings.

### Bird Communities + Geography

The `hawaiibirds` dataset records 183 bird species across 1,183 survey grid cells on 8 Hawaiian islands. Each cell also has geographic coordinates. We'll discover factors that simultaneously explain *which species live where* and *where those places are*.

```{r shared-setup}
data(hawaiibirds)
meta_h <- attr(hawaiibirds, "metadata_h")

# View 1: species counts (183 x 1183, sparse)
A_species <- hawaiibirds

# View 2: geographic coordinates (2 x 1183, dense)
A_geo <- rbind(lat = scale(meta_h$lat)[, 1],
               lng = scale(meta_h$lng)[, 1])
# Match magnitudes: geography as ~10% of total signal
geo_weight <- 0.1 * sum(abs(A_species)) / sum(abs(A_geo))
A_geo_weighted <- as(A_geo * geo_weight, "dgCMatrix")
```

```{r shared-fit}
inp_species <- factor_input(A_species, name = "species")
inp_geo     <- factor_input(A_geo_weighted, name = "geography")
shared_inp  <- factor_shared(inp_species, inp_geo)
layer       <- nmf_layer(shared_inp, k = 8, name = "joint")
config      <- factor_config(maxit = 100, seed = 42)
net         <- factor_net(list(inp_species, inp_geo), layer, config)
result      <- fit(net)
```

The shared $H$ assigns each grid cell to a mixture of 8 factors that must explain both bird communities *and* geographic position:

```{r shared-heatmap, fig.height = 5}
H_shared <- result$joint$H
island <- factor(meta_h$island)

# Mean factor activation per island
activation <- sapply(levels(island), function(isl) {
  rowMeans(H_shared[, island == isl, drop = FALSE])
})

# Normalize rows for display
rmax <- apply(activation, 1, max)
rmax[rmax == 0] <- 1
activation_norm <- activation / rmax

hm_df <- expand.grid(Factor = paste0("F", 1:8), Island = levels(island))
hm_df$Activation <- as.vector(activation_norm)

ggplot(hm_df, aes(x = Island, y = Factor, fill = Activation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Shared Factor Activation by Island",
       subtitle = "Each factor explains both species composition and geography",
       fill = "Relative\nActivation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The geographic constraint encourages spatial coherence: factors that activate on neighboring islands share high activation, while geographically distant islands get distinct factors.

### Comparison: Shared vs. Separate Factorizations

How much does the geographic view change the embedding? We can measure agreement between the shared embedding and a species-only NMF by correlating their $H$ factors.

```{r shared-vs-separate, fig.height = 4}
model_species_only <- nmf(A_species, k = 8, seed = 42, maxit = 100)
H_species_only <- model_species_only@h

# Best absolute correlation between each shared factor and any species-only factor
cross_cor <- cor(t(H_shared), t(H_species_only))
best_match <- apply(abs(cross_cor), 1, max)

cor_df <- data.frame(Factor = factor(1:8), Correlation = best_match)
ggplot(cor_df, aes(x = Factor, y = Correlation)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "gray50") +
  ylim(0, 1) +
  labs(title = "Shared vs. Species-Only: Per-Factor Agreement",
       subtitle = "Factors below the dashed line were reshaped by geographic information",
       y = "Best |r| with species-only factor") +
  theme_minimal()
```

Factors with high correlation ($r > 0.9$) look essentially the same with or without geography --- they capture species patterns visible from counts alone. Factors with lower correlation have been genuinely reshaped by the spatial constraint, surfacing structure that species-only NMF misses.

## Example 3: Deep NMF

Stacking NMF layers extracts structure at multiple granularities. The first layer produces a $k_1$-factor embedding; the second factorizes *that embedding* into $k_2$ meta-factors:

$$A \approx W_1\,\text{diag}(d_1)\,H_1, \qquad H_1 \approx W_2\,\text{diag}(d_2)\,H_2$$

$H_2$ captures coarse global structure; $H_1$ captures fine-grained detail.

### Coarse → Fine in AML

The `aml` dataset has 824 DNA methylation features across 135 samples from 6 categories (3 AML subtypes + 3 control cell types). A two-layer architecture first extracts 10 fine factors, then compresses them to 3 meta-factors.

```{r deep-setup}
data(aml)
meta_aml <- attr(aml, "metadata_h")

inp    <- factor_input(aml, name = "methylation")
layer1 <- nmf_layer(inp, k = 10, name = "fine")
layer2 <- nmf_layer(layer1, k = 3, name = "coarse")
net    <- factor_net(inp, layer2, factor_config(maxit = 100, seed = 42))
deep   <- fit(net)

H_fine   <- deep$fine$H       # 10 x 135 (fine factors x samples)
# For the coarse layer, the per-sample embedding is in W (135 x 3)
# because the second layer factorizes H1: samples become "features"
W_coarse <- deep$coarse$W     # 135 x 3 (samples x coarse factors)
```

```{r deep-barplot, fig.height = 5}
categories <- meta_aml$category
cat_short <- gsub("Control ", "Ctrl-", gsub("AML ", "AML-", categories))

coarse_assign <- factor(paste0("C", apply(W_coarse, 1, which.max)))
fine_assign   <- factor(paste0("F", apply(H_fine, 2, which.max)))

comp_df <- rbind(
  data.frame(Factor = coarse_assign, Category = cat_short, Level = "Coarse (k=3)"),
  data.frame(Factor = fine_assign,   Category = cat_short, Level = "Fine (k=10)")
)

ggplot(comp_df, aes(x = Factor, fill = Category)) +
  geom_bar(width = 0.7) +
  facet_wrap(~ Level, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Deep NMF: Coarse vs. Fine Factor Assignments",
       subtitle = "Coarse factors separate AML/Control; fine factors resolve subtypes",
       y = "Samples") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

The 3 coarse meta-factors capture the broad AML-vs-control distinction. The 10 fine factors resolve subtypes --- MEP vs. GMP vs. L-MPP within each group. This hierarchical view is not available from a single `nmf()` call at either rank.

### The Connecting Matrix

The second layer's $H_2$ matrix (3 × 10) reveals how fine factors compose into coarse ones:

```{r deep-connection, fig.height = 3.5}
H2 <- deep$coarse$H  # 3 x 10
d2 <- deep$coarse$d

# Normalize for visualization
H2_scaled <- sweep(H2, 1, d2, "*")
h_max <- max(abs(H2_scaled))
H2_norm <- H2_scaled / h_max

conn_df <- expand.grid(Coarse = paste0("C", 1:3), Fine = paste0("F", 1:10))
conn_df$Weight <- as.vector(H2_norm)

ggplot(conn_df, aes(x = Fine, y = Coarse, fill = Weight)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  labs(title = "Deep NMF: How Fine Factors Compose into Coarse Ones",
       subtitle = "H₂ matrix connecting layer 2 (coarse) to layer 1 (fine) factors",
       fill = "Relative\nWeight") +
  theme_minimal()
```

Each coarse factor draws from a distinct subset of fine factors --- this is the interpretable bridge between resolutions.

## Example 4: Target Regularization in Graphs

The [Guided NMF](guided-nmf.html) vignette covers target regularization in detail. Within a factor graph, you apply it through `W()` and `H()` per-factor config objects:

```{r target-reg}
data(hawaiibirds)
meta_h <- attr(hawaiibirds, "metadata_h")

# Fit a baseline model, then derive targets from island labels
baseline <- nmf(hawaiibirds, k = 8, seed = 42, maxit = 50)
target <- compute_target(baseline@h, factor(meta_h$island))

# Graph with target regularization on H
inp <- factor_input(hawaiibirds, name = "birds")
guided <- nmf_layer(inp, k = 8, name = "guided",
                    H = H(target = target, target_lambda = 0.3))
net <- factor_net(inp, guided, factor_config(maxit = 100, seed = 42))
guided_result <- fit(net)
```

The target biases $H$ toward island-specific centroids during optimization, producing an embedding with better class separability than unsupervised NMF. Use positive `target_lambda` for **enrichment** (enhance structure) and negative for **removal** (suppress label-correlated variation, useful for batch correction).

```{r target-compare, fig.height = 3.5}
H_unsup  <- baseline@h
H_guided <- guided_result$guided$H
island   <- factor(meta_h$island)

# Measure class separability: ratio of between-class to within-class variance
separability <- function(H, labels) {
  grand_mean <- rowMeans(H)
  levels_l <- levels(labels)
  k <- nrow(H)
  between_var <- 0
  within_var  <- 0
  for (lev in levels_l) {
    idx <- which(labels == lev)
    n_l <- length(idx)
    class_mean <- rowMeans(H[, idx, drop = FALSE])
    between_var <- between_var + n_l * sum((class_mean - grand_mean)^2)
    within_var  <- within_var + sum((H[, idx] - class_mean)^2)
  }
  between_var / max(within_var, .Machine$double.eps)
}

sep_df <- data.frame(
  Method = c("Unsupervised", "Target-regularized (λ=0.3)"),
  Separability = c(separability(H_unsup, island),
                   separability(H_guided, island))
)
knitr::kable(sep_df, digits = 3,
             caption = "Between/within class variance ratio (higher = better separated)")
```

See [Guided NMF](guided-nmf.html) for `refine()`, `compute_target()`, and classification benchmarks.

## Example 5: Cross-Validation over Graph Parameters

`cross_validate_graph()` automates hyperparameter search. You provide a function that builds the output layer from a parameter list, and a grid of values to search:

```{r cv-graph}
data(aml)
inp <- factor_input(aml, name = "aml")

layer_fn <- function(p) nmf_layer(inp, k = p$k, L1 = p$L1, name = "cv")
params <- list(k = c(3, 5, 8, 12), L1 = c(0, 0.01, 0.1))

config <- factor_config(test_fraction = 0.1, cv_seed = 42, maxit = 50)
cv <- cross_validate_graph(inp, layer_fn, params, config,
                           reps = 2, seed = 42, verbose = FALSE)

knitr::kable(head(cv$summary, 6), digits = 5,
             caption = "Top parameter combinations by mean test loss")
```

```{r cv-plot, fig.height = 4}
cv_df <- cv$summary
cv_df$L1 <- factor(cv_df$L1)
cv_df$k <- as.numeric(cv_df$k)

ggplot(cv_df, aes(x = k, y = mean_test_loss, color = L1, group = L1)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = mean_test_loss - se_test_loss,
                    ymax = mean_test_loss + se_test_loss),
                width = 0.3, linewidth = 0.5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Cross-Validation: Rank × L1 Grid Search",
       x = "Rank (k)", y = "Mean Test Loss ± SE", color = "L1") +
  theme_minimal()
```

This works with any graph architecture --- cross-validate the second-layer rank of a deep NMF, the regularization strength of a shared factorization, or any other parameter exposed through `layer_fn`.

## Composable Architectures

The real power is composability. Each block is a node; wire them freely:

**Shared + guided** --- two modalities with label guidance:

```{r compose-shared-guided, eval = FALSE}
inp1 <- factor_input(A1, name = "rna")
inp2 <- factor_input(A2, name = "protein")
layer <- nmf_layer(factor_shared(inp1, inp2), k = 10, name = "joint",
                   H = H(target = T_mat, target_lambda = 0.3))
net <- factor_net(list(inp1, inp2), layer, factor_config(maxit = 100))
```

**Deep + CV** --- search over second-layer rank:

```{r compose-deep-cv, eval = FALSE}
layer1 <- nmf_layer(inp, k = 20, name = "fine")
cv <- cross_validate_graph(
  inp,
  layer_fn = function(p) nmf_layer(layer1, k = p$k, name = "coarse"),
  params = list(k = c(3, 5, 8)),
  config = factor_config(test_fraction = 0.1, maxit = 50)
)
```

**Branching + concatenation** --- parallel factorizations at different ranks, merged:

```{r compose-branch, eval = FALSE}
branch_lo <- nmf_layer(inp, k = 3, name = "coarse")
branch_hi <- nmf_layer(inp, k = 10, name = "fine")
merged <- nmf_layer(factor_concat(branch_lo, branch_hi), k = 5, name = "merged")
net <- factor_net(inp, merged, factor_config(maxit = 50))
```

## Summary

| Need | Graph tool | Example |
|------|-----------|---------|
| One matrix, one NMF | `factor_input` → `nmf_layer` | [Example 1](#example-1-graph--plain-nmf) |
| Multi-modal (shared samples) | `factor_shared()` | [Example 2](#example-2-multi-modal-shared-factorization) |
| Hierarchical decomposition | Stacked `nmf_layer()` calls | [Example 3](#example-3-deep-nmf) |
| Label-guided factors | `H(target=..., target_lambda=...)` | [Example 4](#example-4-target-regularization-in-graphs) |
| Hyperparameter search | `cross_validate_graph()` | [Example 5](#example-5-cross-validation-over-graph-parameters) |
| Custom architectures | `factor_concat()`, `factor_add()` | [Composable Architectures](#composable-architectures) |

## What's Next

- [Guided NMF](guided-nmf.html) --- target regularization, `refine()`, and `compute_target()` in depth
- [Cross-Validation](cross-validation.html) --- rank selection with the simpler `nmf(..., test_fraction=...)` interface
- [NMF Fundamentals](nmf-fundamentals.html) --- the core `nmf()` API
- [Clustering](clustering.html) --- consensus NMF and hierarchical divisive clustering
