---
title: "Composable Factorization Graphs with factor_net"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composable Factorization Graphs with factor_net}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

`factor_net` is a composable graph API for building complex factorization
pipelines from simple, reusable building blocks. Instead of a monolithic
function call, you describe a **factorization graph** — inputs, layers, and
connections — and then `fit()` it.

This approach cleanly supports:

- **Multi-modal**: Shared factorizations across multiple data matrices
- **Deep NMF**: Stacked factorization layers with progressive dimensionality reduction
- **Cross-validation**: Speckled holdout within the graph framework
- **Streaming**: Out-of-core factorization of SparsePress (.spz) files
- **Regularization**: L1, L2, graph Laplacian, guides — all composable per-factor

## Single-Layer NMF

The simplest graph is a single input connected to a single NMF layer — equivalent to calling `nmf()` directly:

```{r}
library(RcppML)
library(Matrix)

set.seed(42)
A <- rsparsematrix(200, 100, 0.2)
A <- abs(A)  # make non-negative

# Build the graph
inp <- factor_input(A, "expression")
L1  <- inp |> nmf_layer(k = 8)
net <- factor_net(inputs = inp, output = L1,
                  config = factor_config(maxit = 100, tol = 1e-4, seed = 42))

# Fit
result <- fit(net)
result
```

Access factors via `$`:

```{r}
dim(result$L1$W)    # features x k
dim(result$L1$H)    # k x samples
length(result$L1$d) # scaling vector
```

## Per-Factor Configuration

Each layer can have independent settings for the W and H factors using `W()` and `H()`:

```{r}
# Sparse W, regularised H
L1 <- inp |> nmf_layer(
  k = 8,
  W = W(L1 = 0.1),             # LASSO on W
  H = H(L1 = 0.01, L2 = 0.05) # Elastic net on H
)
net <- factor_net(inputs = inp, output = L1,
                  config = factor_config(maxit = 50, seed = 42))
result <- fit(net)
```

Layer-level defaults (`L1 = 0.01`) apply to both factors unless overridden by a `W()` or `H()` block.

## Multi-Modal Shared Factorization

For data from multiple modalities measured on the same samples (e.g., RNA + ATAC), use `factor_shared()` to share the H factor:

```{r eval = FALSE}
# Two modalities, same 100 samples
A1 <- abs(rsparsematrix(150, 100, 0.15))  # RNA (150 genes)
A2 <- abs(rsparsematrix(80, 100, 0.2))    # ATAC (80 peaks)

inp1 <- factor_input(A1, "rna")
inp2 <- factor_input(A2, "atac")

shared <- factor_shared(inp1, inp2)
L1 <- shared |> nmf_layer(k = 6)

net <- factor_net(
  inputs = list(inp1, inp2), output = L1,
  config = factor_config(maxit = 100, seed = 42))
result <- fit(net)

# Each modality gets its own W; H is shared
dim(result$L1$W$rna)   # 150 x 6
dim(result$L1$W$atac)  # 80 x 6
dim(result$L1$H)       # 6 x 100  (shared)
```

This is equivalent to `nmf(rbind(A1, A2), k = 6)` but gives you named, split W matrices without manual bookkeeping.

## Deep NMF

Stack NMF layers to learn a hierarchy of representations:

```{r}
inp <- factor_input(A, "expression")
enc <- inp |> nmf_layer(k = 15, name = "encoder")
bot <- enc |> nmf_layer(k = 4,  name = "bottleneck")

net <- factor_net(
  inputs = inp, output = bot,
  config = factor_config(maxit = 30, tol = 1e-6, seed = 42))
result <- fit(net)

result$n_layers                    # 2
dim(result$encoder$W)              # 200 x 15
dim(result$bottleneck$W)           # 100 x 4
```

The first layer factorizes $A \approx W_1 \, \text{diag}(d_1) \, H_1$, then the second
factorizes $H_1^T \approx W_2 \, \text{diag}(d_2) \, H_2$.

### Mixed SVD + NMF

Use `svd_layer()` for a linear pre-processing step (PCA), then refine with NMF:

```{r}
inp <- factor_input(A, "expression")
pca <- inp |> svd_layer(k = 20, name = "pca")
nmf <- pca |> nmf_layer(k = 5, name = "nmf")

net <- factor_net(inputs = inp, output = nmf,
                  config = factor_config(maxit = 30, seed = 42))
result <- fit(net)

# SVD allows negative entries; NMF is non-negative
any(result$pca$W < 0)
all(result$nmf$W >= 0)
```

## Graph Laplacian Regularization

Incorporate prior knowledge about feature relationships via a graph Laplacian:

```{r}
# Feature similarity graph (chain Laplacian for illustration)
L_graph <- bandSparse(200, 200, k = c(0, 1),
                      diagonals = list(rep(2, 200), rep(-1, 199)),
                      symmetric = TRUE)
L_graph <- as(L_graph, "dgCMatrix")

inp <- factor_input(A, "expression")
L1 <- inp |> nmf_layer(k = 8, W = W(graph = L_graph, graph_lambda = 0.5))
net <- factor_net(inputs = inp, output = L1,
                  config = factor_config(maxit = 50, seed = 42))
result <- fit(net)
```

The penalty $\lambda \cdot \text{tr}(W^T L W)$ encourages smooth factors over connected features.

## Cross-Validation

Enable speckled holdout directly in the graph config:

```{r}
inp <- factor_input(A, "expression")
L1  <- inp |> nmf_layer(k = 8)
net <- factor_net(
  inputs = inp, output = L1,
  config = factor_config(
    maxit = 100, tol = 1e-4, seed = 42,
    test_fraction = 0.1,  # 10% speckled holdout
    cv_seed = 99L,
    mask_zeros = FALSE,
    patience = 5L))
result <- fit(net)

result$layers$L1$test_loss   # test MSE
result$layers$L1$loss        # training MSE
```

### Rank Selection with `cross_validate_graph()`

Search over hyperparameter grids:

```{r}
inp <- factor_input(A, "expression")
cv <- cross_validate_graph(
  inputs = inp,
  layer_fn = function(p) inp |> nmf_layer(k = p$k),
  params = list(k = c(3, 5, 8, 12)),
  config = factor_config(maxit = 50, seed = 42, test_fraction = 0.1),
  reps = 3, seed = 42, verbose = FALSE)

cv$summary        # mean/sd test loss per rank
cv$best_params    # optimal k
```

Multi-parameter search:

```{r}
cv2 <- cross_validate_graph(
  inputs = inp,
  layer_fn = function(p) inp |> nmf_layer(k = p$k, L1 = p$L1),
  params = list(k = c(5, 10), L1 = c(0, 0.01, 0.05)),
  config = factor_config(maxit = 30, seed = 42, test_fraction = 0.1),
  reps = 2, seed = 42, verbose = FALSE)

cv2$summary
```

## Streaming Out-of-Core NMF

For matrices too large to fit in memory, save to SparsePress format and pass the file path:

```{r eval = FALSE}
# Save large matrix to .spz
sp_write(big_matrix, "data.spz")

# Stream factorization — matrix never loaded into RAM
inp <- factor_input("data.spz", "large_data")
L1  <- inp |> nmf_layer(k = 10)
net <- factor_net(inputs = inp, output = L1,
                  config = factor_config(maxit = 100, seed = 42))
result <- fit(net)
```

## Prediction on New Data

Project new samples through a fitted graph:

```{r}
inp <- factor_input(A, "expression")
L1  <- inp |> nmf_layer(k = 8)
net <- factor_net(inputs = inp, output = L1,
                  config = factor_config(maxit = 100, tol = 1e-4, seed = 42))
result <- fit(net)

# New data with same features
A_new <- abs(rsparsematrix(200, 20, 0.2))
H_new <- predict(result, A_new)
dim(H_new)  # 8 x 20
```

## Training Logger

Monitor convergence during fitting:

```{r}
logger <- training_logger(log_loss = TRUE, log_norms = TRUE)

inp <- factor_input(A, "expression")
enc <- inp |> nmf_layer(k = 10, name = "enc")
bot <- enc |> nmf_layer(k = 4, name = "bot")
net <- factor_net(inputs = inp, output = bot,
                  config = factor_config(maxit = 20, tol = 1e-8, seed = 42))
result <- fit(net, logger = logger)

log_df <- as.data.frame(result$logger)
head(log_df)
```

## Comparison with `nmf()`

| Feature | `nmf()` | `factor_net` |
|---------|---------|--------------|
| Single-layer NMF | Direct | Via graph (identical results) |
| Multi-modal | `nmf(list(...))` | `factor_shared()` (explicit) |
| Deep NMF | Not supported | Chain layers with `|>` |
| Cross-validation | `nmf(k = 2:15)` | `cross_validate_graph()` |
| Streaming | `nmf("file.spz")` | `factor_input("file.spz")` |
| Per-factor config | `L1 = c(w, h)` | `W(L1=...), H(L1=...)` |
| Graph regularization | Yes | Yes, via `W(graph=...)` |
| Guides | Not in `nmf()` | Via `guide_*()` constructors |
| Training logging | Not available | `training_logger()` |

For single-layer NMF with default settings, `nmf()` is the simpler choice. Use `factor_net` when you need composability, deep architectures, or fine-grained control.
