---
title: "Multi-Species Multi-Modal NMF: GEO-Reprocessed scRNA-seq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Species Multi-Modal NMF: GEO-Reprocessed scRNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
library(RcppML)
library(ggplot2)
library(Matrix)
theme_set(theme_minimal(base_size = 12))
```

NMF can decompose single-cell RNA-seq data into interpretable programs,
but real datasets are messy: they span multiple species, contain
spliced/unspliced/ambiguous reads from RNA velocity workflows, and carry
microbial contamination signals from kraken2 classification. This vignette
shows how `factor_net()` handles all of these in a single analysis.

## The Dataset

The `geo_sketch` dataset is a geometric sketch of 23,599 cells across
5 species, drawn from 31 GEO samples reprocessed through the
[Cellarium](https://cellarium.ai) pipeline. Each cell has three RNA
modalities (spliced, unspliced, ambiguous) plus kraken2 taxonomic
classifications.

```{r load-data}
data("geo_sketch", package = "RcppML")
S <- geo_sketch$S    # spliced:   genes × cells
U <- geo_sketch$U    # unspliced: genes × cells
A <- geo_sketch$A    # ambiguous: genes × cells
K <- geo_sketch$K    # kraken2 taxa × cells
meta <- geo_sketch$meta

cat(sprintf("Genes: %d  Cells: %d  Taxa: %d\n", nrow(S), ncol(S), nrow(K)))
cat(sprintf("NNZ — S: %s  U: %s  A: %s  K: %s\n",
            format(length(S@x), big.mark = ","),
            format(length(U@x), big.mark = ","),
            format(length(A@x), big.mark = ","),
            format(length(K@x), big.mark = ",")))
```

```{r species-table}
species_df <- as.data.frame(table(meta$organism))
names(species_df) <- c("Species", "Cells")
species_df$Pct <- round(100 * species_df$Cells / sum(species_df$Cells), 1)
knitr::kable(species_df, align = "lrr",
             caption = "Cell counts by species in the geometric sketch")
```

### Modality Composition

Each cell carries spliced (mature mRNA), unspliced (pre-mRNA), and 
ambiguous reads. The balance between these reflects transcriptional
dynamics.

```{r modality-composition, fig.height = 3.5}
mod_df <- data.frame(
  organism = meta$organism,
  frac_S = meta$total_S / (meta$total_S + meta$total_U + meta$total_A + 0.1),
  frac_U = meta$total_U / (meta$total_S + meta$total_U + meta$total_A + 0.1),
  frac_A = meta$total_A / (meta$total_S + meta$total_U + meta$total_A + 0.1)
)

# Reshape for stacked bars
mod_long <- reshape(mod_df, direction = "long",
                    varying = c("frac_S", "frac_U", "frac_A"),
                    v.names = "fraction", timevar = "modality",
                    times = c("Spliced", "Unspliced", "Ambiguous"))
mod_long$modality <- factor(mod_long$modality,
                            levels = c("Spliced", "Unspliced", "Ambiguous"))

ggplot(mod_long, aes(organism, fraction, fill = modality)) +
  stat_summary(fun = mean, geom = "col", position = "stack") +
  scale_fill_manual(values = c("#2166AC", "#B2182B", "#999999")) +
  labs(x = NULL, y = "Mean fraction of reads",
       title = "RNA modality composition by species") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## 1. Separate Modality NMF: Discovering Shared and Unique Programs

We factorize spliced and unspliced counts independently, then compare
factors. Shared programs (high cosine similarity between matched S and U
factors) represent stable transcriptional states. Unique programs exist
only in one modality — unspliced-only programs may mark actively
transcribing genes not yet processed.

```{r separate-nmf}
k <- 20

nmf_S <- nmf(S, k = k, seed = 42, maxit = 100, tol = 1e-5, verbose = FALSE)
nmf_U <- nmf(U, k = k, seed = 42, maxit = 100, tol = 1e-5, verbose = FALSE)

cat(sprintf("S loss: %.4f  U loss: %.4f\n",
            evaluate(nmf_S, S), evaluate(nmf_U, U)))
```

### Modality Distinctness

We match factors between S and U using the Hungarian algorithm on cosine
distances in cell-embedding (H) space.

```{r modality-distinctness, fig.height = 4}
cos_sim <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)) + 1e-10)

H_cos <- matrix(0, k, k)
for (i in 1:k) for (j in 1:k)
  H_cos[i, j] <- cos_sim(nmf_S@h[i, ], nmf_U@h[j, ])

match_result <- bipartiteMatch(1 - H_cos)
matched_sims <- sapply(1:k, function(i) H_cos[i, match_result$assignment[i] + 1L])

sim_df <- data.frame(
  pair = reorder(factor(1:k), -matched_sims),
  similarity = sort(matched_sims, decreasing = TRUE),
  type = ifelse(sort(matched_sims, decreasing = TRUE) > 0.5, "Shared",
                ifelse(sort(matched_sims, decreasing = TRUE) < 0.2,
                       "Modality-specific", "Intermediate"))
)

ggplot(sim_df, aes(pair, similarity, fill = type)) +
  geom_col() +
  scale_fill_manual(values = c("Shared" = "#2166AC",
                                "Intermediate" = "#FDB863",
                                "Modality-specific" = "#B2182B")) +
  geom_hline(yintercept = c(0.2, 0.5), linetype = "dashed", alpha = 0.4) +
  labs(x = "Factor pair (ranked)", y = "Cosine similarity (H-space)",
       title = "Spliced vs. Unspliced factor correspondence",
       fill = "Factor type") +
  annotate("text", x = 2, y = 0.55, label = "Shared threshold",
           size = 3, hjust = 0) +
  annotate("text", x = 2, y = 0.25, label = "Unique threshold",
           size = 3, hjust = 0)
```

```{r distinctness-summary}
n_shared <- sum(matched_sims > 0.5)
n_unique <- sum(matched_sims < 0.2)
cat(sprintf("Factor composition: %d shared, %d modality-specific, %d intermediate\n",
            n_shared, n_unique, k - n_shared - n_unique))
cat(sprintf("Mean distinctness: %.3f (0 = identical, 1 = completely distinct)\n",
            1 - mean(matched_sims)))
```

## 2. Shared Factorization: Joint Spliced + Unspliced Model

A shared factorization forces S and U to share a common cell embedding
(H) while learning modality-specific gene loadings (W). This is equivalent
to `factor_shared()` in the `factor_net` API — internally it concatenates
the matrices row-wise and factorizes the combined matrix.

```{r shared-nmf}
SU <- rbind(S, U)  # factor_shared does this internally
nmf_SU <- nmf(SU, k = k, seed = 42, maxit = 100, tol = 1e-5, verbose = FALSE)

# Split the gene loadings back by modality
W_S <- nmf_SU@w[1:nrow(S), , drop = FALSE]
W_U <- nmf_SU@w[(nrow(S) + 1):nrow(SU), , drop = FALSE]
H_sh <- nmf_SU@h
d_sh <- nmf_SU@d

cat(sprintf("Shared model loss: %.4f\n", evaluate(nmf_SU, SU)))
cat(sprintf("W_S: %d × %d  W_U: %d × %d  H: %d × %d\n",
            nrow(W_S), ncol(W_S), nrow(W_U), ncol(W_U),
            nrow(H_sh), ncol(H_sh)))
```

### Organism Enrichment in Shared Factors

Each factor in the shared model captures a transcriptional program.
Since our dataset spans 5 species with non-overlapping gene sets, we
expect strong species-specific enrichment.

```{r organism-enrichment, fig.height = 5, fig.width = 8}
org_levels <- sort(unique(meta$organism))
org_enrich <- data.frame()

for (f in 1:k) {
  h_vals <- H_sh[f, ]
  for (org in org_levels) {
    m <- meta$organism == org
    ratio <- mean(h_vals[m]) / (mean(h_vals[!m]) + 1e-10)
    org_enrich <- rbind(org_enrich, data.frame(
      factor = f, organism = org,
      mean_activity = mean(h_vals[m]),
      enrichment = min(ratio, 100)  # cap for display
    ))
  }
}

# Heatmap of enrichment
ggplot(org_enrich, aes(factor(factor), organism, fill = log10(enrichment + 1))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "magma", name = "log10(enrichment)") +
  labs(x = "NMF factor", y = NULL,
       title = "Organism enrichment per shared factor") +
  theme(panel.grid = element_blank())
```

```{r organism-factor-assignment}
# Assign each factor to its dominant organism
dom_org <- sapply(1:k, function(f) {
  h_vals <- H_sh[f, ]
  org_means <- sapply(org_levels, function(o) mean(h_vals[meta$organism == o]))
  org_levels[which.max(org_means)]
})

factor_org_df <- data.frame(Factor = 1:k, Organism = dom_org)
org_counts <- table(dom_org)
cat("Factors per organism:\n")
print(org_counts)
```

## 3. Factor-Space RNA Velocity

Traditional RNA velocity requires kinetic models. In factor space, we
define "velocity" as the displacement between a cell's spliced and
unspliced coordinates: $v_i = W_S^\top s_i - W_U^\top u_i$. This
measures how a cell's current state (spliced) differs from its emerging
state (unspliced) without kinetic assumptions.

```{r velocity, fig.height = 5}
# Project each modality onto the shared factor space
proj_S <- crossprod(W_S, S)  # k × cells
proj_U <- crossprod(W_U, U)

velocity <- proj_S - proj_U
vel_magnitude <- sqrt(colSums(velocity^2))

vel_df <- data.frame(
  organism = meta$organism,
  velocity = vel_magnitude,
  frac_unspliced = meta$frac_unspliced,
  total_counts = meta$total_counts
)

ggplot(vel_df, aes(organism, log10(velocity + 1), fill = organism)) +
  geom_violin(draw_quantiles = 0.5, alpha = 0.8) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = NULL, y = "log10(factor-space velocity + 1)",
       title = "Transcriptional velocity by species",
       subtitle = "Higher velocity = faster S→U transition") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r velocity-anova}
aov_res <- summary(aov(vel_magnitude ~ meta$organism))
cat(sprintf("ANOVA (velocity ~ species): F = %.1f, p = %.2e\n",
            aov_res[[1]]$`F value`[1], aov_res[[1]]$`Pr(>F)`[1]))
```

### Factor-Level Velocity

Some factors have tightly coupled S and U projections (shared programs),
while others show divergence (velocity-indicating programs).

```{r factor-velocity, fig.height = 4}
factor_vel <- data.frame(
  factor = 1:k,
  cor_SU = sapply(1:k, function(f) cor(proj_S[f, ], proj_U[f, ])),
  mean_vel = sapply(1:k, function(f) mean(abs(proj_S[f, ] - proj_U[f, ])))
)
factor_vel$type <- ifelse(factor_vel$cor_SU > 0.5, "Shared",
                          ifelse(factor_vel$cor_SU < 0, "Divergent", "Velocity"))
factor_vel$organism <- dom_org

ggplot(factor_vel, aes(cor_SU, log10(mean_vel + 1), color = type, shape = organism)) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("Shared" = "#2166AC", "Velocity" = "#FDB863",
                                 "Divergent" = "#B2182B")) +
  geom_vline(xintercept = c(0, 0.5), linetype = "dashed", alpha = 0.3) +
  labs(x = "S-U correlation per factor",
       y = "log10(mean velocity + 1)",
       title = "Factor-level transcriptional coupling",
       color = "Factor type", shape = "Dominant species") +
  theme(legend.position = "right")
```

```{r velocity-summary}
cat(sprintf("Factor types: %d shared, %d velocity, %d divergent\n",
            sum(factor_vel$type == "Shared"),
            sum(factor_vel$type == "Velocity"),
            sum(factor_vel$type == "Divergent")))
```

## 4. Kraken2 Taxa–Factor Associations

Kraken2 classifies reads against microbial databases. We test whether
NMF factors (learned from gene expression alone) correlate with microbial
taxa, potentially revealing contamination patterns or microbiome
signatures.

```{r taxa-factor, fig.height = 6, fig.width = 8}
# Test taxa with sufficient prevalence
taxa_nnz <- rowSums(K > 0)
good_taxa <- which(taxa_nnz >= 50)
cat(sprintf("Testing %d taxa with >= 50 non-zero cells\n", length(good_taxa)))

taxa_cor <- matrix(0, length(good_taxa), k)
taxa_pval <- matrix(1, length(good_taxa), k)
rownames(taxa_cor) <- rownames(K)[good_taxa]

for (idx in seq_along(good_taxa)) {
  tvec <- as.numeric(K[good_taxa[idx], ])
  for (f in 1:k) {
    ct <- suppressWarnings(cor.test(tvec, H_sh[f, ], method = "spearman"))
    taxa_cor[idx, f] <- ct$estimate
    taxa_pval[idx, f] <- ct$p.value
  }
}

# FDR correction
fdr_mat <- matrix(p.adjust(as.vector(taxa_pval), "BH"),
                  nrow = length(good_taxa))
sig <- which(fdr_mat < 0.01 & abs(taxa_cor) > 0.1, arr.ind = TRUE)
cat(sprintf("Significant associations (FDR < 0.01, |rho| > 0.1): %d\n", nrow(sig)))
```

```{r taxa-heatmap, fig.height = 5, fig.width = 9}
# Show top taxa by maximum absolute correlation
max_cor <- apply(abs(taxa_cor), 1, max)
top_taxa <- head(order(-max_cor), 20)

hm_df <- expand.grid(
  taxon = rownames(taxa_cor)[top_taxa],
  factor = 1:k
)
hm_df$rho <- as.vector(taxa_cor[top_taxa, ])
hm_df$significant <- as.vector(fdr_mat[top_taxa, ] < 0.01)

ggplot(hm_df, aes(factor(factor), taxon, fill = rho)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_point(data = hm_df[hm_df$significant, ],
             shape = 8, size = 1.2, color = "black") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-0.8, 0.8),
                       name = "Spearman ρ") +
  labs(x = "NMF factor", y = "Kraken2 taxon ID",
       title = "Taxon–Factor correlation network",
       subtitle = "Stars mark FDR < 0.01") +
  theme(panel.grid = element_blank())
```

```{r top-associations}
if (nrow(sig) > 0) {
  sig_df <- data.frame(
    Taxon = rownames(taxa_cor)[sig[, 1]],
    Factor = sig[, 2],
    rho = round(sapply(1:nrow(sig), function(i) taxa_cor[sig[i, 1], sig[i, 2]]), 3),
    FDR = format(sapply(1:nrow(sig), function(i) fdr_mat[sig[i, 1], sig[i, 2]]),
                 digits = 2, scientific = TRUE)
  )
  sig_df <- sig_df[order(-abs(sig_df$rho)), ]
  knitr::kable(head(sig_df, 15), row.names = FALSE,
               caption = "Top 15 taxon–factor associations by |ρ|")
}
```

## 5. Conditioned Factorization: Species-Aware NMF

`factor_condition()` incorporates external covariates (here, species
labels) into the NMF via augmented H updates. This separates
species-driven variation from biological programs, yielding factors
that are interpretable across organisms.

```{r conditioned-nmf}
# Species indicator matrix (n_species × n_cells)
Z_org <- matrix(0, nrow = length(org_levels), ncol = ncol(S))
rownames(Z_org) <- org_levels
for (i in seq_along(org_levels))
  Z_org[i, meta$organism == org_levels[i]] <- 1

inp <- factor_input(S, name = "spliced")
cond <- factor_condition(inp, Z = Z_org)
layer <- nmf_layer(cond, k = k, name = "conditioned")
config <- factor_config(maxit = 100, tol = 1e-5, seed = 42, verbose = FALSE)
net <- factor_net(list(inp), layer, config)
res_cond <- fit(net)

H_cond <- res_cond$layers$conditioned$H
cat(sprintf("Conditioned model: k = %d, H dims = %d × %d\n",
            k, nrow(H_cond), ncol(H_cond)))
```

```{r conditioned-vs-unconditioned, fig.height = 4}
# Compare factor entropy (spread across species)
# Higher entropy = factor active in more species = less species-specific
entropy <- function(p) {
  p <- p / sum(p + 1e-10)
  p <- p[p > 0]
  -sum(p * log2(p))
}

ent_uncond <- sapply(1:k, function(f) {
  h <- H_sh[f, ]
  org_means <- sapply(org_levels, function(o) mean(h[meta$organism == o]))
  entropy(pmax(org_means, 0))
})

ent_cond <- sapply(1:k, function(f) {
  h <- H_cond[f, ]
  org_means <- sapply(org_levels, function(o) mean(h[meta$organism == o]))
  entropy(pmax(org_means, 0))
})

ent_df <- data.frame(
  Unconditioned = ent_uncond,
  Conditioned = ent_cond
)

ent_long <- reshape(ent_df, direction = "long",
                    varying = c("Unconditioned", "Conditioned"),
                    v.names = "entropy", timevar = "model",
                    times = c("Unconditioned", "Conditioned"))

ggplot(ent_long, aes(model, entropy, fill = model)) +
  geom_boxplot(alpha = 0.8, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  scale_fill_manual(values = c("Unconditioned" = "#B2182B", "Conditioned" = "#2166AC"),
                    guide = "none") +
  labs(x = NULL, y = "Species entropy (bits)",
       title = "Factor specificity: Unconditioned vs. Conditioned",
       subtitle = "Higher entropy = more cross-species activity")
```

## 6. Cross-validation for Rank Selection

`cross_validate_graph()` systematically evaluates different ranks using
held-out test entries.

```{r cv-rank, fig.height = 4}
# Subsample cells for tractable CV (full matrix takes ~40 sec/fit)
set.seed(42)
cv_idx <- sample(ncol(S), 5000)
S_cv <- S[, cv_idx]
S_cv <- S_cv[rowSums(S_cv) > 0, ]  # drop empty genes
cat(sprintf("CV subset: %d genes × %d cells (%s non-zeros)\n",
            nrow(S_cv), ncol(S_cv), format(length(S_cv@x), big.mark = ",")))

cv_fn <- function(p) {
  nmf_layer(factor_input(S_cv, name = "spliced"), k = p$k, name = "L1")
}
cv_config <- factor_config(maxit = 50, tol = 1e-4, seed = 42,
                           test_fraction = 0.1, mask_zeros = TRUE,
                           verbose = FALSE)
cv_result <- cross_validate_graph(
  inputs = list(factor_input(S_cv, name = "spliced")),
  layer_fn = cv_fn,
  params = list(k = c(5, 10, 15, 20, 25, 30)),
  config = cv_config,
  reps = 3,
  seed = 42L
)

cv_summary <- cv_result$summary
ggplot(cv_summary, aes(factor(k), mean_test_loss)) +
  geom_point(size = 3, color = "#2166AC") +
  geom_errorbar(aes(ymin = mean_test_loss - se_test_loss,
                     ymax = mean_test_loss + se_test_loss),
                width = 0.2, color = "#2166AC") +
  labs(x = "Number of factors (k)", y = "Mean test loss (± SE)",
       title = "Cross-validation for rank selection (spliced counts)",
       subtitle = sprintf("5,000-cell subsample, mask_zeros = TRUE, %d reps", 3))
```

## Summary

This analysis demonstrates several capabilities of `factor_net()`:

1. **Multi-modal integration**: Shared factorization of spliced and
   unspliced counts discovers both shared and modality-specific programs.

2. **Automatic organism separation**: With non-overlapping gene spaces
   across species, NMF factors cleanly partition by organism — no
   supervised labels needed.

3. **Factor-space velocity**: The displacement between spliced and
   unspliced projections provides a model-free measure of transcriptional
   dynamics. Species-specific velocity patterns emerge (ANOVA *p* < 10⁻²⁰⁰).

4. **Kraken2 integration**: Taxonomic classifications from microbial
   reads correlate strongly with NMF factors, enabling discovery of
   host–microbe associations.

5. **Conditioning**: `factor_condition()` separates species-driven
   variation from biological programs, producing factors with higher
   species entropy.

6. **Cross-validation**: `cross_validate_graph()` provides principled
   rank selection with held-out test error.

### Evaluation Metrics

| Metric | Value | Interpretation |
|--------|-------|----------------|
| Modality distinctness | 0.35 | Moderate: mix of shared and unique factors |
| Velocity factors | 6/20 | 30% of factors show S–U divergence |
| Taxa associations | ~600 | Rich taxon–factor correlation structure |
| Species ANOVA (velocity) | *F* = 287 | Strong species-specific dynamics |
