---
title: "Getting Started with RcppML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RcppML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## What Is RcppML?

**RcppML** decomposes a nonnegative matrix $A$ into low-rank factors
$A \approx W \cdot \text{diag}(d) \cdot H$, where $W$ captures features,
$H$ captures sample loadings, and $d$ gives per-factor scale.

Built on Rcpp and Eigen with OpenMP parallelism and optional CUDA GPU support,
RcppML is designed for large sparse matrices common in genomics, recommender
systems, and image analysis. Key capabilities include:
 
- **NMF** with coordinate-descent and Cholesky NNLS solvers
- **SVD / PCA** with optional non-negativity and L1 constraints
- **Cross-validation** via speckled holdout for automatic rank selection
- **Distribution-aware losses**: Gaussian, Poisson, Generalized Poisson, Negative Binomial, Gamma, Inverse Gaussian, Tweedie — plus zero-inflation
- **Regularization**: L1, L2, L21, angular, graph Laplacian, upper bounds
- **Consensus clustering** and divisive clustering (`dclust`)
- **Composable factorization graphs** via `factor_net()`
- **StreamPress** `.spz` format for out-of-core computation
- **GPU acceleration** via CUDA

## Installation

Install from CRAN:

```r
install.packages("RcppML")
```

Or install the development version:

```r
remotes::install_github("zdebruine/RcppML")
```

For GPU support, see the [GPU Acceleration](gpu-acceleration.html) vignette.

## Quick NMF Demo

Generate a synthetic nonnegative matrix and recover its factors:

```{r nmf-demo}
sim <- simulateNMF(200, 100, k = 5, noise = 0.3, seed = 42)
model <- nmf(sim$A, k = 5, seed = 1)
```

```{r nmf-summary, echo = FALSE}
knitr::kable(data.frame(
  Rows = nrow(sim$A),
  Columns = ncol(sim$A),
  Rank = 5,
  `Reconstruction MSE` = round(evaluate(model, sim$A), 4),
  Iterations = model@misc$iter,
  `Runtime (sec)` = round(as.numeric(model@misc$runtime, units = "secs"), 2),
  check.names = FALSE
), caption = "NMF model summary")
```

With a low-noise signal and the correct rank, NMF recovers the generating factors and achieves near-zero reconstruction error.

```{r nmf-convergence, echo = FALSE, fig.height = 3.5}
# Show convergence of a quick NMF run
track_loss <- function() {
  losses <- numeric(0)
  list(
    callback = function(iter, train_loss, test_loss) { losses[iter] <<- train_loss },
    get = function() losses
  )
}
tracker <- track_loss()
invisible(nmf(sim$A, k = 5, seed = 1, tol = 1e-6, maxit = 50, on_iteration = tracker$callback))
loss_df <- data.frame(Iteration = seq_along(tracker$get()), Loss = tracker$get())
ggplot(loss_df, aes(x = Iteration, y = Loss)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  scale_y_log10() +
  labs(title = "NMF Convergence", x = "Iteration", y = "Reconstruction Loss (log)") +
  theme_minimal()
```

## Quick SVD Demo

Compute a rank-5 truncated SVD of the AML chromatin-accessibility dataset:

```{r svd-demo}
data(aml)
sv <- svd(aml, k = 5)
```

```{r svd-table, echo = FALSE}
ve <- variance_explained(sv)
knitr::kable(data.frame(
  Component = 1:5,
  `Variance Explained` = paste0(round(ve * 100, 1), "%"),
  `Cumulative` = paste0(round(cumsum(ve) * 100, 1), "%"),
  check.names = FALSE
), caption = "Variance explained by the first 5 components")
```

```{r svd-scree, echo = FALSE, fig.height = 3}
ve_df <- data.frame(Component = 1:5, VE = ve * 100)
ggplot(ve_df, aes(x = Component, y = VE)) +
  geom_col(fill = "steelblue", width = 0.6) +
  labs(title = "Scree Plot: AML Chromatin Accessibility", x = "Component", y = "% Variance Explained") +
  theme_minimal()
```

## Quick Cross-Validation Demo

Use speckled holdout cross-validation to find the best rank for the AML data:

```{r cv-demo}
cv <- nmf(aml, k = c(2, 4, 6, 8, 10), test_fraction = 0.2, seed = 42, maxit = 50)
```

```{r cv-plot, echo = FALSE, fig.height = 3.5}
plot(cv)
```

The test loss curve reveals the rank at which the model begins to overfit. The
minimum identifies the optimal number of factors for this dataset.

## Built-in Datasets

RcppML ships seven datasets spanning diverse domains:

```{r dataset-table, echo = FALSE}
ds <- data.frame(
  Dataset = c("`aml`", "`golub`", "`hawaiibirds`", "`movielens`",
              "`olivetti`", "`digits`", "`pbmc3k`"),
  Dimensions = c("824 × 135", "38 × 5,000", "183 × 1,183",
                  "3,867 × 610", "400 × 4,096", "1,797 × 64",
                  "8,000 × 500"),
  Type = c("Dense matrix", "Sparse (dgCMatrix)", "Sparse (dgCMatrix)",
           "Sparse (dgCMatrix)", "Sparse (dgCMatrix)", "Sparse (dgCMatrix)",
           "SPZ raw bytes"),
  Domain = c("Chromatin accessibility (AML)", "Gene expression (leukemia)",
             "Bird species counts", "Movie ratings",
             "Face images (Olivetti)", "Handwritten digit images",
             "Single-cell RNA-seq (PBMCs)"),
  check.names = FALSE
)
knitr::kable(ds, caption = "Shipped datasets")
```

The `pbmc3k` dataset is stored as StreamPress-compressed raw bytes (a variance-selected
subset of the full 13,714 × 2,700 10x Genomics PBMC 3k dataset). Decompress
it with `st_read()` — see the [StreamPress](streampress.html) vignette.

## Where to Go Next

### Core Techniques

- [NMF Fundamentals](nmf-fundamentals.html) — Solvers, diagonal scaling, convergence, and basic workflow
- [SVD and PCA](svd-pca.html) — Truncated SVD, PCA, non-negative and sparse variants
- [Cross-Validation](cross-validation.html) — Speckled holdout for automatic rank selection
- [Statistical Distributions](distributions.html) — Distribution-aware losses and zero-inflation handling
- [Regularization and Constraints](regularization.html) — L1, L2, L21, angular penalties, and upper bounds

### Advanced Methods

- [Clustering, Consensus, and Classification](clustering.html) — Divisive clustering, consensus NMF, and sample classification
- [Factorization Graphs](factor-graphs.html) — Multi-modal and guided factorization via `factor_net()`

### Applications

- [Image Decomposition](image-decomposition.html) — Learning visual parts from face and digit images
- [Recommendation Systems](recommendation-systems.html) — Building a movie recommender with sparse NMF

### Infrastructure

- [StreamPress](streampress.html) — High-performance sparse matrix compression and out-of-core NMF
- [GPU Acceleration](gpu-acceleration.html) — CUDA-based GPU support for large-scale factorization
