---
title: "Getting Started with RcppML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RcppML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

RcppML provides high-performance Non-negative Matrix Factorization (NMF) for R. It decomposes a non-negative matrix $A$ (features × samples) into three factors:

$$A \approx W \cdot \text{diag}(d) \cdot H$$

where:

- **W** (features × k): feature loadings (non-negative)
- **d** (length k): scaling diagonal
- **H** (k × samples): sample embeddings (non-negative)
- **k**: the factorization rank (number of factors)

The diagonal $d$ ensures that columns of $W$ and rows of $H$ each sum to 1, providing consistent factor scalings regardless of initialization.

## Installation

```{r eval = FALSE}
# From GitHub (development version)
devtools::install_github("zdebruine/RcppML")
```

## Quick Start

```{r}
library(RcppML)
library(Matrix)
```

### Simulating NMF Data

`simulateNMF()` generates synthetic data with known ground truth factors:

```{r}
sim <- simulateNMF(nrow = 100, ncol = 80, k = 5, noise = 0.1, dropout = 0.3, seed = 42)
A <- sim$A  # input matrix
dim(A)
```

### Fitting a Basic NMF Model

```{r}
model <- nmf(A, k = 5, seed = 123, verbose = FALSE)
model
```

The result is an S4 object of class `nmf` with slots:

- `model@w`: the W factor matrix
- `model@d`: the diagonal scaling vector
- `model@h`: the H factor matrix
- `model@misc`: metadata (iterations, loss, runtime, etc.)

You can also use `$` access:

```{r}
dim(model$w)
length(model$d)
dim(model$h)
```

### Evaluating the Fit

```{r}
# Mean squared error
evaluate(model, A)

# Reconstruction
A_approx <- prod(model)  # W * diag(d) * H
dim(A_approx)
```

## Working with Real Data

RcppML includes several example datasets:

```{r}
data(hawaiibirds)
dim(hawaiibirds)
```

```{r}
model_birds <- nmf(hawaiibirds, k = 4, seed = 42, verbose = FALSE)
model_birds
```

### Interpreting Factors

Each column of W represents a "basis pattern" across features, and each row of H represents how much each sample loads onto that pattern.

```{r}
# Top features for factor 1
w1 <- model_birds$w[, 1]
head(sort(w1, decreasing = TRUE), 5)
```

### Sparsity of Factors

```{r}
sparsity(model_birds)
```

### Plotting

If `track_loss_history = TRUE` was set during fitting, you can visualize convergence:

```{r, eval = FALSE}
model_tracked <- nmf(hawaiibirds, k = 10, track_loss_history = TRUE)
plot(model_tracked)
```

## Dense vs Sparse Input

RcppML automatically selects optimized code paths for dense (`matrix`) and sparse (`dgCMatrix`) inputs:
  
```{r}
# Dense
A_dense <- as.matrix(sim$A)
m_dense <- nmf(A_dense, k = 5, seed = 42, maxit = 20, verbose = FALSE)

# Sparse
A_sparse <- as(sim$A, "dgCMatrix")
m_sparse <- nmf(A_sparse, k = 5, seed = 42, maxit = 20, verbose = FALSE)

# Both converge to similar solutions
cat("Dense loss:", m_dense$loss, "\n")
cat("Sparse loss:", m_sparse$loss, "\n")
```

## Regularization

### L1 (LASSO) — Encourage Sparsity

L1 penalties drive small factor values to exactly zero:

```{r}
m_L1 <- nmf(A, k = 5, L1 = c(0.1, 0.1), seed = 42, verbose = FALSE)

# Compare sparsity
cat("No L1 — H zeros:", sum(model$h == 0), "\n")
cat("L1=0.1 — H zeros:", sum(m_L1$h == 0), "\n")
```

### L2 (Ridge) — Shrinkage

L2 penalties shrink factor magnitudes:

```{r}
m_L2 <- nmf(A, k = 5, L2 = c(0.1, 0.1), seed = 42, verbose = FALSE)
```

### Upper Bound Constraints

Constrain factor values to a maximum:

```{r}
m_bounded <- nmf(A, k = 5, upper_bound = c(1.0, 1.0), seed = 42, verbose = FALSE)
cat("Max W value:", max(m_bounded$w), "\n")
cat("Max H value:", max(m_bounded$h), "\n")
```

## Loss Functions

RcppML supports four loss functions. MSE is the default; the others use IRLS (Iteratively Reweighted Least Squares):

```{r}
# Mean Squared Error (default)
m_mse <- nmf(A, k = 5, loss = "mse", seed = 42, verbose = FALSE)

# Mean Absolute Error (robust to outliers)
m_mae <- nmf(A, k = 5, loss = "mae", seed = 42, verbose = FALSE)

# Huber loss (quadratic near zero, linear for large errors)
m_huber <- nmf(A, k = 5, loss = "huber", huber_delta = 1.0, seed = 42, verbose = FALSE)
```

## Semi-NMF

By default, both W and H are constrained to be non-negative. Semi-NMF relaxes the constraint on W:

```{r}
# Allow negative values in W
m_semi <- nmf(A, k = 5, nonneg = c(FALSE, TRUE), seed = 42, verbose = FALSE)
cat("Min W value:", min(m_semi$w), " (can be negative)\n")
cat("Min H value:", min(m_semi$h), " (still non-negative)\n")
```

## Reproducibility

Setting `seed` ensures deterministic results:

```{r}
m1 <- nmf(A, k = 3, seed = 999, maxit = 10, verbose = FALSE)
m2 <- nmf(A, k = 3, seed = 999, maxit = 10, verbose = FALSE)
identical(m1$w, m2$w)  # TRUE
```

## S4 Methods

The `nmf` class supports many standard R operations:

```{r}
# Dimensions: m, n, k
dim(model)

# Transpose (swap W and H)
t_model <- t(model)
dim(t_model)

# Sort by factor importance (decreasing d)
sorted <- sort(model)
sorted$d  # decreasing order

# Subset factors
sub <- model[1:3]
dim(sub)
```

## Projection and Prediction

Project a trained model onto new data:

```{r}
# Split data
A_train <- A[, 1:60]
A_test <- A[, 61:80]

# Train on subset
m_train <- nmf(A_train, k = 5, seed = 42, verbose = FALSE)

# Project onto test data
H_test <- predict(m_train, A_test)
dim(H_test)
```

## NNLS Solver

The standalone NNLS solver is available for custom use:

```{r}
# Create a small system: minimize ||Ax - b||^2  subject to x >= 0
set.seed(42)
w <- matrix(runif(50), 10, 5)
A_small <- matrix(runif(30), 10, 3)
h <- nnls(w = w, A = A_small)
dim(h)
```

## Parallelism

Set the number of threads for OpenMP parallelism:

```{r eval = FALSE}
options(RcppML.threads = 4)  # Use 4 threads
options(RcppML.threads = 0)  # Use all available threads (default)
```

## Next Steps

- **Cross-Validation**: See `vignette("cross-validation")` for rank selection and speckled holdout
- **Clustering**: Use `bipartition()` and `dclust()` for divisive clustering
- **SparsePress**: Use `sp_write()` / `sp_read()` for compressed sparse matrix I/O
- **Consensus NMF**: Use `consensus_nmf()` for robust factor estimation
