---
title: "GPU-Accelerated Computing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPU-Accelerated Computing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## Motivation

At scale, dimensionality reduction is bottlenecked by dense linear algebra. Each
NMF iteration computes Gram matrices ($H H^T$, $W^T W$) and right-hand-side
products ($A H^T$, $W^T A$) — operations that map directly onto GPU hardware
where thousands of cores execute matrix multiplies in parallel. RcppML's GPU
backend uses NVIDIA CUDA with cuBLAS and cuSPARSE to accelerate these operations,
achieving **10–40× speedups** on real single-cell datasets even against a
fully-utilized 56-thread CPU.

The GPU advantage is not uniform across operations. Standard NMF already benefits
significantly, but **cross-validation and SVD randomized show the largest
gains** because these modes perform dense BLAS work (Gram matrix corrections,
block matrix multiplies) that maps directly onto GPU hardware. The GPU is
optional — all code falls back to CPU transparently when CUDA is unavailable.

## Quick Start

```{r eval = FALSE}
library(RcppML)

# Check GPU availability
gpu_available()
#> [1] TRUE
gpu_info()
#>   device              name total_mb free_mb
#> 1      0  NVIDIA H100 NVL    95830   93500

# NMF with GPU — same API, add resource = "gpu"
model <- nmf(A, k = 20, resource = "gpu", seed = 42)
```

The `resource` parameter accepts `"gpu"`, `"cpu"`, or `"auto"` (default). With
`"auto"`, RcppML uses GPU when available and falls back to CPU otherwise.

## Benchmarks on Real Data

How much faster is GPU in practice? We benchmarked 10 operations spanning standard
NMF, cross-validation, IRLS (KL divergence), sparse-mask NMF, and SVD
on two real single-cell RNA-seq datasets.

### Data Preparation

```{r eval = FALSE}
library(RcppML)
library(Matrix)
library(SeuratData)
library(Seurat)

# ── hcabm40k: 40,000 bone marrow cells ──
data("hcabm40k")
hca <- UpdateSeuratObject(hcabm40k)
counts <- hca[["RNA"]]$counts                    # 17,369 genes × 40,000 cells

# Select top 5,000 variable genes, log-normalize
gene_vars <- Matrix::rowMeans(counts^2) - Matrix::rowMeans(counts)^2
top5k <- head(order(gene_vars, decreasing = TRUE), 5000)
hca_5k <- counts[top5k, ]
lib_sizes <- Matrix::colSums(hca_5k)
hca_norm <- log1p(hca_5k %*% Diagonal(x = 1e4 / lib_sizes))
hca_norm <- as(hca_norm, "dgCMatrix")
# 5,000 × 40,000, nnz ≈ 33M, density ≈ 16.5%

# 10K-cell subset for expensive k = 64 operations
set.seed(42)
hca_10k <- hca_norm[, sample(ncol(hca_norm), 10000)]
# 5,000 × 10,000, nnz ≈ 8.3M

# ── pbmc3k: 500 PBMCs (bundled with RcppML) ──
data(pbmc3k, package = "RcppML")
tmp <- tempfile(fileext = ".spz")
writeBin(pbmc3k, tmp)
pbmc <- st_read(tmp)
pbmc_norm <- log1p(pbmc %*% Diagonal(x = 1e4 / Matrix::colSums(pbmc)))
pbmc_norm <- as(pbmc_norm, "dgCMatrix")
# 8,000 × 500, nnz ≈ 412K
```

### Results

All benchmarks ran on an **NVIDIA H100 NVL** (96 GB HBM3) against a **56-thread
Intel Xeon Gold 6238R** (2 × 28 cores, all threads active). NMF iterations
were forced with `tol = 0` and `maxit = 20`.

```{r benchmark-table, echo = FALSE}
bench_df <- data.frame(
  Operation = c(
    "NMF (k=20, 40K cells)",
    "NMF (k=64, 10K cells)",
    "Cross-validation NMF (k=64, 10K cells)",
    "Cross-validation NMF (k=16, pbmc3k)",
    "KL-divergence NMF (k=16, pbmc3k)",
    "Sparse-mask NMF (k=20, 10K cells)",
    "SVD Lanczos (k=10, 40K cells)",
    "SVD randomized (k=10, 40K cells)",
    "SVD IRLBA (k=10, 40K cells)",
    "NMF (k=20, pbmc3k)"
  ),
  Dataset = c(
    "hcabm40k", "hcabm40k (10K)", "hcabm40k (10K)", "pbmc3k",
    "pbmc3k", "hcabm40k (10K)",
    "hcabm40k", "hcabm40k", "hcabm40k",
    "pbmc3k"
  ),
  CPU_s = c(38.45, 29.23, 75.31, 4.04,
            23.37, 10.50,
            4.78, 17.77, 5.30,
            2.18),
  GPU_s = c(2.78, 0.88, 2.39, 0.20,
            1.98, 0.75,
            0.44, 0.41, 0.38,
            0.21),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
bench_df$Speedup <- bench_df$CPU_s / bench_df$GPU_s
bench_df <- bench_df[order(bench_df$Speedup, decreasing = TRUE), ]

knitr::kable(
  data.frame(
    Operation = bench_df$Operation,
    `CPU (s)` = sprintf("%.1f", bench_df$CPU_s),
    `GPU (s)` = sprintf("%.2f", bench_df$GPU_s),
    Speedup = sprintf("%.0f×", bench_df$Speedup),
    check.names = FALSE
  ),
  align = "lrrr",
  caption = "CPU (56-thread Xeon) vs. GPU (H100 NVL) benchmarks"
)
```

```{r benchmark-plot, echo = FALSE, fig.width = 7, fig.height = 4.2}
library(ggplot2)

bench_df$Type <- ifelse(grepl("Cross-validation", bench_df$Operation), "Cross-validation",
                 ifelse(grepl("KL", bench_df$Operation), "IRLS (KL loss)",
                 ifelse(grepl("SVD", bench_df$Operation), "SVD",
                 ifelse(grepl("Sparse-mask", bench_df$Operation), "Sparse-mask NMF",
                 "Standard NMF"))))

# Shorter labels for plot readability
bench_df$Short <- gsub(" \\(.*", "", bench_df$Operation)
bench_df$Label <- sprintf("%s  (%.0f\u00d7)", bench_df$Short, bench_df$Speedup)
bench_df$Label <- factor(bench_df$Label, levels = rev(bench_df$Label))

ggplot(bench_df, aes(x = Speedup, y = Label, fill = Type)) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = c(
    "Cross-validation" = "#e63946",
    "IRLS (KL loss)" = "#f4a261",
    "Standard NMF" = "#457b9d",
    "Sparse-mask NMF" = "#2a9d8f",
    "SVD" = "#a8dadc"
  )) +
  labs(x = "GPU speedup (\u00d7 faster than CPU)", y = NULL, fill = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(0.4, "cm"),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 8.5),
    plot.title = element_text(face = "bold", size = 11)
  ) +
  ggtitle("GPU speedup by operation (H100 NVL vs. 56-thread Xeon)")
```

### Key Observations

Three patterns stand out:

1. **SVD randomized shows the largest speedup** (43×). Randomized SVD performs
   full matrix–dense block multiplies ($A Q$, $A^T B$) that translate into a
   handful of cuSPARSE/cuBLAS calls on GPU. Cross-validation NMF at high rank is
   close behind (32×) because CV corrects the Gram matrix for every held-out
   entry each iteration — a $O(k^2)$ update per test-set column that batches
   naturally on GPU.

2. **All operations achieve ≥ 10× speedup**, even against a fully-utilized
   56-thread Xeon. NMF at high rank ($k = 64$, 33×), IRLS KL-divergence (12×),
   and sparse-mask NMF (14×) all benefit from GPU BLAS acceleration.

3. **Speedups scale with problem size and rank**. The 40K-cell NMF at $k = 20$
   (14×) benefits less than $k = 64$ on 10K cells (33×) because higher rank
   increases the Gram matrix and NNLS work that the GPU absorbs efficiently.
   With fewer CPU threads (e.g., 8), speedups increase proportionally.

## Why GPU Gains Are Larger for CV and IRLS

Standard NMF spends most of each iteration on two operations:

$$H H^T \quad (k \times k), \qquad A H^T \quad (m \times k)$$

Cross-validation adds a correction step: for each held-out column $j$ in the test
set, the Gram matrix is adjusted by subtracting the contribution of the missing
entries:

$$G_j = H H^T - \sum_{i \in \text{mask}(j)} h_i h_i^T$$

On CPU, this per-column correction loop is sequential. On GPU, the corrections
for all test columns are batched into a single sparse outer-product kernel. The
result: CV overhead that dominates CPU time becomes negligible on GPU, pushing
CV speedups well above standard NMF speedups.

IRLS losses (KL divergence, Gamma–Poisson, etc.) similarly increase per-iteration
work by solving weighted NNLS subproblems that require additional Gram matrix
updates with observation-specific weights — again, heavy BLAS work that the GPU
absorbs with minimal overhead.

## API Reference

### Resource Override Priority

The compute backend is determined by (highest priority first):

1. `resource` parameter in `nmf()` or `svd()`
2. `RCPPML_RESOURCE` environment variable
3. `options(RcppML.gpu = TRUE/FALSE)`
4. Auto-detection (GPU if available, else CPU)

```{r eval = FALSE}
# Force CPU even when GPU is available
model <- nmf(A, k = 20, resource = "cpu", seed = 42)

# Session-wide override via environment variable
Sys.setenv(RCPPML_RESOURCE = "cpu")

# Or via R option
options(RcppML.gpu = FALSE)
```

### GPU-Supported Features

| Feature | GPU Support |
|:--------|:------------|
| Sparse NMF | Yes |
| Dense NMF | Yes |
| Cross-validation NMF | Yes |
| MSE loss | Yes |
| MAE / Huber (via `robust`) | Yes |
| KL / distribution losses | Yes |
| L1, L2 regularization | Yes |
| L21, angular regularization | Yes |
| Graph regularization | CPU only |
| Upper bound constraints | Yes |
| Bipartition / dclust | Yes |
| StreamPress streaming NMF | Yes (CPU decompression → GPU compute) |

When a GPU-unsupported feature is requested, RcppML falls back to CPU
automatically with no error.

### GPU-Direct StreamPress Reading

For very large datasets stored as `.spz` files, `st_read_gpu()` transfers the
matrix directly into GPU memory, avoiding a full CPU-side `dgCMatrix` copy:

```{r eval = FALSE}
gpu_handle <- st_read_gpu("large_dataset.spz")
model <- nmf(gpu_handle, k = 30, seed = 42)
st_free_gpu(gpu_handle)   # optional — GC handles cleanup
```

See the [StreamPress](streampress.html) vignette for `.spz` format details.

## Building the GPU Library

The GPU shared library is built separately from the R package using `nvcc`.

### Requirements

- **NVIDIA GPU** with compute capability $\geq$ 7.0 (Volta or newer)
- **CUDA Toolkit** $\geq$ 12.x (cuBLAS and cuSPARSE included)

### Build Steps

```{bash eval = FALSE}
# 1. Load CUDA on your system
module load cuda/12.8.1   # HPC example

# 2. Build the GPU shared library (use the project Makefile)
cd RcppML/src/
make -f Makefile.gpu install

# Or manually:
nvcc -O3 \
  -gencode arch=compute_80,code=sm_80 \
  -gencode arch=compute_90,code=sm_90 \
  -std=c++17 -Xcompiler -fPIC --shared \
  -I../inst/include \
  -o ../inst/lib/RcppML_gpu.so \
  gpu_bridge_nmf.cu gpu_bridge_svd.cu gpu_bridge_cluster.cu \
  gpu_bridge_utils.cu sp_gpu_bridge.cu \
  -lcublas -lcusparse -lcuda

# 3. Verify
Rscript -e 'library(RcppML); cat("GPU:", gpu_available(), "\n")'
#> GPU: TRUE
```

Adapt the `-gencode` flags to your GPU architecture:

```{r arch-table, echo = FALSE}
arch_df <- data.frame(
  `GPU Family` = c("V100", "A100", "H100", "RTX 3090", "RTX 4090"),
  `Compute Capability` = c("7.0", "8.0", "9.0", "8.6", "8.9"),
  Flag = c("arch=compute_70,code=sm_70",
           "arch=compute_80,code=sm_80",
           "arch=compute_90,code=sm_90",
           "arch=compute_86,code=sm_86",
           "arch=compute_89,code=sm_89"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(arch_df, align = "llr",
             caption = "CUDA architecture flags by GPU family")
```

### Library Search Path

RcppML searches for `RcppML_gpu.so` in order:
`inst/lib/` → `src/` → package `lib/` → working directory root.

### GPU Memory

NMF GPU memory usage is approximately:

$$\text{Memory} \approx \text{nnz} \times 12 + (m + n) \times k \times 8 \text{ bytes}$$

where nnz is the number of non-zeros, $m$ and $n$ are matrix dimensions, and $k$
is rank. For a 30,000 × 20,000 matrix with 5% density at $k = 30$, this is roughly
3.6 GB — well within a 32 GB GPU.

### HPC / SLURM Usage

```{bash eval = FALSE}
#!/bin/bash
#SBATCH --partition=gpu
#SBATCH --gres=gpu:1
#SBATCH --cpus-per-task=8
#SBATCH --mem=32G
#SBATCH --time=04:00:00

module load r/4.5.2 cuda/12.8.1
Rscript my_analysis.R
```

## Troubleshooting

**GPU not detected?**

```{r eval = FALSE}
gpu_available(force_recheck = TRUE)
#> [1] FALSE
```

Common causes: CUDA drivers missing or outdated (`nvidia-smi` to check), GPU
compute capability < 7.0, `RcppML_gpu.so` not on the search path, or architecture
mismatch (library compiled for `sm_90` won't load on a V100).

**Precision**: GPU uses FP32 for core operations and FP64 selectively for
numerically sensitive accumulations. Results are equivalent to CPU (FP64) within
floating-point tolerance.

## Key Takeaways

1. **Same API, different backend**: `resource = "gpu"` is the only change needed.
2. **10–40× speedups** on real single-cell data against a fully-utilized
   56-thread CPU, with SVD randomized and cross-validation showing the largest
   gains. With fewer CPU threads, speedups scale proportionally.
3. **Automatic fallback**: If GPU is unavailable or a feature is CPU-only,
   RcppML falls back silently.
4. **Direct disk-to-GPU** via `st_read_gpu()` for very large `.spz` datasets.

*See the [NMF Fundamentals](nmf-fundamentals.html) vignette for comprehensive
parameter documentation and the [StreamPress](streampress.html) vignette for
`.spz` file format details.*
