---
title: "GPU Acceleration Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPU Acceleration Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

RcppML supports GPU-accelerated NMF via NVIDIA CUDA. The GPU backend uses cuBLAS and cuSPARSE for dense linear algebra and custom CUDA kernels for sparse operations and the coordinate descent NNLS solver.

GPU acceleration is **optional**: the package installs and functions fully on CPU. The GPU shared library (`RcppML_gpu.so`) is loaded at runtime only when CUDA hardware is detected.

## Requirements

- **NVIDIA GPU** with compute capability ≥ 7.0 (Volta or newer: V100, A100, H100, RTX 20xx+)
- **CUDA Toolkit** ≥ 12.x
- **cuBLAS** and **cuSPARSE** (included with CUDA Toolkit)

## Building the GPU Library

The GPU component is compiled separately from the main R package using `nvcc`:

```bash
# 1. Load CUDA on your system (HPC example)
module load cuda/12.8.1

# 2. Navigate to the CUDA build directory
cd benchmarks/cuda/

# 3. Build the production benchmark (to test GPU alone)
make production_benchmark

# 4. Build the GPU shared library for R
cd ../../src/
nvcc -O3 \
  -gencode arch=compute_80,code=sm_80 \
  -gencode arch=compute_90,code=sm_90 \
  -std=c++17 -Xcompiler -fPIC \
  --shared \
  -I../inst/include \
  -o ../inst/lib/RcppML_gpu.so \
  gpu_bridge.cu \
  -lcublas -lcusparse -lcuda
```

Adapt the `-gencode` flags to your GPU architecture:

| GPU Family | Compute Capability | Flag |
|:-----------|:-------------------|:-----|
| V100 | 7.0 | `arch=compute_70,code=sm_70` |
| A100 | 8.0 | `arch=compute_80,code=sm_80` |
| H100 | 9.0 | `arch=compute_90,code=sm_90` |
| RTX 3090 | 8.6 | `arch=compute_86,code=sm_86` |
| RTX 4090 | 8.9 | `arch=compute_89,code=sm_89` |

## Library Search Path

RcppML searches for `RcppML_gpu.so` (or `.dll` on Windows) in these locations (in order):

1. `<package_install_path>/lib/`
2. `<package_install_path>/src/`
3. `<working_directory>/src/`
4. `<working_directory>/inst/lib/`
5. `<working_directory>/`

For development, placing the library in `inst/lib/` is recommended.

## Checking GPU Availability

```{r eval = FALSE}
library(RcppML)

# Check if GPU is available
gpu_available()
#> [1] TRUE

# Get device details
gpu_info()
#>   device   name total_mb  free_mb
#> 1      0 GPU 0    32510    31200
```

## Using GPU NMF

GPU dispatch happens automatically when a GPU is available:

```{r eval = FALSE}
library(Matrix)
A <- rsparsematrix(10000, 5000, 0.05)

# Auto-dispatch (uses GPU if available, CPU otherwise)
model <- nmf(A, k = 20, verbose = TRUE)
#> [RcppML] Using GPU backend for NMF

# Force CPU even when GPU is available
model_cpu <- nmf(A, k = 20, resource = "cpu")

# Force GPU (errors if unavailable)
model_gpu <- nmf(A, k = 20, resource = "gpu")
```

### Resource Override Priority

The compute backend is determined by (highest priority first):

1. `resource` parameter in `nmf()`
2. `RCPPML_RESOURCE` environment variable
3. `options(RcppML.gpu = TRUE/FALSE)`  
4. Auto-detection

```{r eval = FALSE}
# Environment variable
Sys.setenv(RCPPML_RESOURCE = "cpu")

# R option
options(RcppML.gpu = FALSE)
```

## GPU-Supported Features

| Feature | GPU Support |
|:--------|:------------|
| Sparse NMF | Yes |
| Dense NMF | Yes |
| Cross-validation NMF | Yes |
| MSE loss | Yes |
| MAE / Huber / KL loss | Yes |
| L1, L2 regularization | Yes |
| L21, angular regularization | Yes |
| Graph regularization | CPU only |
| Upper bound constraints | Yes |
| Bipartition | Yes |
| Divisive clustering (dclust) | Yes |
| SparsePress streaming NMF | CPU only |

When a GPU-unsupported feature is requested, RcppML automatically falls back to the CPU backend.

## Performance Expectations

GPU acceleration is most beneficial for:

- **Large matrices**: >1000 rows, >1000 columns
- **Moderate to high rank**: k ≥ 10
- **Many iterations**: the GPU amortizes setup costs over iterations

Expected speedups over a multi-threaded CPU (48 cores):

| Operation | Typical Speedup |
|:----------|:----------------|
| Gram matrix (G = H'H) | 5–10× |
| Sparse RHS (B = A·H') | 10–30× |
| NNLS batch solve | 20–50× |
| **Full NMF iteration** | **10–20×** |

For small matrices or low rank (k < 5), CPU may be faster due to GPU kernel launch overhead.

## Precision

GPU computation currently uses double precision. Single precision support is planned
for a future release to enable faster performance on consumer GPUs with fewer FP64 units.

## Troubleshooting

### GPU Not Detected

```{r eval = FALSE}
gpu_available(force_recheck = TRUE)  # Force re-probe
```

Common causes:
- CUDA drivers not installed or outdated
- GPU compute capability < 7.0
- `RcppML_gpu.so` not found in search path
- Library compiled for wrong architecture (e.g., sm_90 on a V100)

### CUDA Out of Memory

NMF GPU memory usage is approximately:

$$\text{Memory} \approx \text{nnz} \times 12 + (m + n) \times k \times 8 \text{ bytes}$$

where nnz is the number of non-zeros, m and n are matrix dimensions, and k is rank.

For very large matrices, reduce memory pressure:
- Reduce rank k
- Use CPU for the largest problems

### Runtime Errors

Set verbose mode to diagnose issues:

```{r eval = FALSE}
model <- nmf(A, k = 20, resource = "gpu", verbose = TRUE)
```

## HPC / SLURM Usage

On SLURM-managed clusters, request GPU resources:

```bash
# Interactive GPU session
salloc --partition=gpu --gres=gpu:1 --cpus-per-task=8 --mem=32G --time=02:00:00

# Batch job
sbatch --partition=gpu --gres=gpu:1 --wrap="Rscript my_nmf_script.R"
```

Example SLURM batch script:

```bash
#!/bin/bash
#SBATCH --partition=gpu
#SBATCH --gres=gpu:1
#SBATCH --cpus-per-task=8
#SBATCH --mem=32G
#SBATCH --time=04:00:00
#SBATCH --output=nmf_gpu_%j.log

module load r/4.5.2 cuda/12.8.1

Rscript -e '
library(RcppML)
cat("GPU available:", gpu_available(), "\n")
print(gpu_info())

data(pbmc3k)
model <- nmf(pbmc3k, k = 20, resource = "gpu", verbose = TRUE)
cat("Loss:", model@misc$loss, "\n")
'
```
