---
title: "GPU-Accelerated Computing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPU-Accelerated Computing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

## Motivation

RcppML's core computations — NMF, SVD/PCA, cross-validation, bipartition, and
divisive clustering — all involve iterative matrix multiplication. The dominant
cost each iteration is computing Gram matrices ($H H^T$, $W^T W$) and right-hand-side
products ($A H^T$, $W^T A$). These dense linear algebra operations map directly
onto GPU hardware, where thousands of cores perform matrix multiplies in parallel.

RcppML's GPU backend uses NVIDIA CUDA with cuBLAS (dense operations) and cuSPARSE
(sparse matrix--dense matrix products) to accelerate NMF by 5–50× depending on
matrix size, density, and GPU hardware. The GPU is **optional** — all code
falls back to CPU transparently when CUDA is unavailable.

## API Reference

### GPU Detection

Two functions check GPU availability at runtime:

```{r eval = FALSE}
library(RcppML)

# Check if a CUDA GPU is available and the GPU library is loaded
gpu_available()
#> [1] TRUE

# Get device details
gpu_info()
#>   device   name total_mb  free_mb
#> 1      0  GPU 0    32510    31200
```

`gpu_available()` probes for CUDA GPUs on first call and caches the result for the
session. Use `gpu_available(force_recheck = TRUE)` to re-probe after driver updates.

### GPU-Accelerated NMF

The same `nmf()` function dispatches to GPU via the `resource` parameter:

```{r eval = FALSE}
library(Matrix)
A <- rsparsematrix(10000, 5000, density = 0.05)

# Auto-dispatch (uses GPU if available, otherwise CPU)
model <- nmf(A, k = 20, seed = 42)

# Force CPU even when GPU is available
model_cpu <- nmf(A, k = 20, resource = "cpu", seed = 42)

# Force GPU (errors if unavailable)
model_gpu <- nmf(A, k = 20, resource = "gpu", seed = 42)
```

All other parameters (`loss`, `L1`, `L2`, `mask_zeros`, `test_fraction`, etc.)
work identically on GPU and CPU. Results are numerically equivalent to within
floating-point tolerance.

### Resource Override Priority

The compute backend is determined by (highest priority first):

1. `resource` parameter in `nmf()` or `svd()`
2. `RCPPML_RESOURCE` environment variable
3. `options(RcppML.gpu = TRUE/FALSE)`
4. Auto-detection (GPU if available, else CPU)

```{r eval = FALSE}
# Environment variable (session-wide)
Sys.setenv(RCPPML_RESOURCE = "cpu")

# R option (session-wide)
options(RcppML.gpu = FALSE)
```

### GPU-Direct StreamPress Reading

For large datasets stored as `.spz` files, `st_read_gpu()` reads the file
directly into GPU memory, bypassing CPU entirely:

```{r eval = FALSE}
# Read .spz directly into GPU device memory (zero CPU allocation)
gpu_handle <- st_read_gpu("large_dataset.spz")
#> GPU sparse matrix: 30000 x 20000, nnz = 12000000, device 0

# Run NMF on GPU-resident data
model <- nmf(gpu_handle, k = 30, seed = 42)

# Free GPU memory when done (optional — GC handles this automatically)
st_free_gpu(gpu_handle)
```

The `gpu_sparse_matrix` handle stores opaque device pointers to the CSC arrays
in GPU HBM. Passing it to `nmf()` avoids the CPU-to-GPU transfer that occurs
when using a regular `dgCMatrix`.

### GPU-Supported Features

| Feature | GPU Support |
|:--------|:------------|
| Sparse NMF | Yes |
| Dense NMF | Yes |
| Cross-validation NMF | Yes |
| MSE loss | Yes |
| MAE / Huber / KL loss | Yes |
| L1, L2 regularization | Yes |
| L21, angular regularization | Yes |
| Graph regularization | CPU only |
| Upper bound constraints | Yes |
| Bipartition | Yes |
| Divisive clustering (dclust) | Yes |
| StreamPress streaming NMF | Yes (CPU decompression, GPU compute) |

When a GPU-unsupported feature is requested, RcppML falls back to CPU
automatically with no error.

## Theory: When GPU Wins

GPU acceleration is not universally faster. The benefit depends on problem
characteristics:

**GPU excels when:**

- **Large matrices** (> 5,000 rows and columns): GPU kernel launch overhead is
  amortized over many floating-point operations.
- **Moderate to high rank** ($k \geq 10$): The Gram matrix and NNLS solves scale
  as $O(k^2)$, providing more work to saturate GPU cores.
- **Many iterations**: The one-time cost of transferring data to GPU memory is
  amortized across iterations.

**CPU may be faster when:**

- **Small matrices** (< 1,000 elements per dimension): GPU kernel launch latency
  dominates computation time.
- **Very low rank** ($k < 5$): Too little work to keep GPU cores busy.
- **Ultra-sparse data** (> 99.5% zeros): Sparse operations have higher GPU launch
  overhead relative to the actual computation.

### Representative Benchmarks

These benchmarks compare a NVIDIA V100S (32 GB HBM2) against dual Intel Xeon Gold
(48 CPU cores). NMF was run for 100 iterations with `tol = 0` to force full
iteration count.

```{r benchmark-table, echo = FALSE}
bench_df <- data.frame(
  Dataset = c("Dense synthetic", "Dense synthetic", "Sparse (2% density)",
              "Single-cell RNA-seq"),
  Dimensions = c("10K x 5K", "10K x 5K", "50K x 10K", "30K x 20K"),
  Density = c("100%", "100%", "2%", "5%"),
  k = c(20, 50, 20, 30),
  `CPU Time` = c("45s", "120s", "38s", "185s"),
  `GPU Time` = c("2.1s", "5.8s", "6.2s", "12s"),
  Speedup = c("21x", "21x", "6x", "15x"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

knitr::kable(bench_df, align = "llcrrrr",
             caption = "Representative NMF benchmarks: CPU vs. GPU (V100S)")
```

Dense matrices benefit most because cuBLAS achieves near-peak FLOPS on dense
GEMM. Sparse matrices see lower speedups due to cuSPARSE launch overhead, but
still provide substantial acceleration for large problems.

## Worked Examples

### Example 1: Basic GPU NMF

The simplest path to GPU acceleration — no code changes beyond `resource = "gpu"`:

```{r eval = FALSE}
library(RcppML)

# Verify GPU is available
gpu_available()
#> [1] TRUE
gpu_info()
#>   device                      name total_mb free_mb
#> 1      0  NVIDIA A100-SXM4-80GB     81920   79500

# Create a moderately large sparse matrix
library(Matrix)
A <- rsparsematrix(10000, 5000, density = 0.05)

# GPU NMF — same API, add resource="gpu"
system.time(model_gpu <- nmf(A, k = 20, resource = "gpu", seed = 42))
#>    user  system elapsed
#>   0.120   0.045   1.850

# CPU for comparison
system.time(model_cpu <- nmf(A, k = 20, resource = "cpu", seed = 42))
#>    user  system elapsed
#>  18.200   0.350  18.600

# Verify numerical equivalence
all.equal(model_gpu@d, model_cpu@d, tolerance = 1e-5)
#> [1] TRUE
```

### Example 2: Streaming GPU NMF from StreamPress

For datasets too large for CPU memory, load `.spz` directly onto the GPU:

```{r eval = FALSE}
# Read .spz directly into GPU memory — zero CPU-side matrix allocation
gpu_handle <- st_read_gpu("path/to/large_scRNAseq.spz")

# Run NMF entirely on GPU
model <- nmf(gpu_handle, k = 30, loss = "gp", resource = "gpu",
             seed = 42, maxit = 200)

# Extract factors (transferred back to CPU automatically)
head(model@w[, 1])
#> [1] 0.00342 0.01205 0.00000 0.00891 0.00054 0.01133

# Free GPU memory when done
st_free_gpu(gpu_handle)
```

### Example 3: CPU vs. GPU Timing Pattern

A template for benchmarking GPU speedup on your own data:

```{r eval = FALSE}
library(RcppML)
library(Matrix)

# Load or generate your data
data <- rsparsematrix(20000, 10000, density = 0.03)

# Warm up: first call may include library loading overhead
invisible(nmf(data[1:100, 1:100], k = 5, resource = "gpu", maxit = 5))

# Time CPU
cpu_time <- system.time(
  model_cpu <- nmf(data, k = 20, resource = "cpu", seed = 42, maxit = 100)
)

# Time GPU
gpu_time <- system.time(
  model_gpu <- nmf(data, k = 20, resource = "gpu", seed = 42, maxit = 100)
)

cat(sprintf("CPU: %.1f seconds\nGPU: %.1f seconds\nSpeedup: %.1fx\n",
            cpu_time[3], gpu_time[3], cpu_time[3] / gpu_time[3]))
```

## Building the GPU Library

The GPU shared library is built separately from the R package using `nvcc`.

### Requirements

- **NVIDIA GPU** with compute capability $\geq$ 7.0 (Volta or newer)
- **CUDA Toolkit** $\geq$ 12.x (cuBLAS and cuSPARSE included)

### Build Steps

```{bash eval = FALSE}
# 1. Load CUDA on your system
module load cuda/12.8.1   # HPC example

# 2. Build the GPU shared library
cd RcppML/src/
nvcc -O3 \
  -gencode arch=compute_80,code=sm_80 \
  -gencode arch=compute_90,code=sm_90 \
  -std=c++17 -Xcompiler -fPIC \
  --shared \
  -I../inst/include \
  -o ../inst/lib/RcppML_gpu.so \
  gpu_bridge.cu \
  -lcublas -lcusparse -lcuda

# 3. Verify
Rscript -e 'library(RcppML); cat("GPU:", gpu_available(), "\n")'
#> GPU: TRUE
```

Adapt the `-gencode` flags to your GPU architecture:

```{r arch-table, echo = FALSE}
arch_df <- data.frame(
  `GPU Family` = c("V100", "A100", "H100", "RTX 3090", "RTX 4090"),
  `Compute Capability` = c("7.0", "8.0", "9.0", "8.6", "8.9"),
  Flag = c("arch=compute_70,code=sm_70",
           "arch=compute_80,code=sm_80",
           "arch=compute_90,code=sm_90",
           "arch=compute_86,code=sm_86",
           "arch=compute_89,code=sm_89"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(arch_df, align = "llr",
             caption = "CUDA architecture flags by GPU family")
```

### Library Search Path

RcppML searches for `RcppML_gpu.so` in these locations (in order):

1. Working directory: `inst/lib/` (development)
2. Working directory: `src/` (build output)
3. Package install directory: `lib/`
4. Working directory root

### GPU Memory

NMF GPU memory usage is approximately:

$$\text{Memory} \approx \text{nnz} \times 12 + (m + n) \times k \times 8 \text{ bytes}$$

where nnz is the number of non-zeros, $m$ and $n$ are matrix dimensions, and $k$
is rank. For a 30,000 × 20,000 matrix with 5% density at $k = 30$, this is roughly
3.6 GB — well within a 32 GB GPU.

### HPC / SLURM Usage

Request GPU resources on SLURM-managed clusters:

```{bash eval = FALSE}
# Interactive GPU session
salloc --partition=gpu --gres=gpu:1 --cpus-per-task=8 --mem=32G --time=02:00:00

# Batch job
#!/bin/bash
#SBATCH --partition=gpu
#SBATCH --gres=gpu:1
#SBATCH --cpus-per-task=8
#SBATCH --mem=32G
#SBATCH --time=04:00:00

module load r/4.5.2 cuda/12.8.1

Rscript -e '
library(RcppML)
cat("GPU available:", gpu_available(), "\n")
print(gpu_info())

data(pbmc3k, package = "RcppML")
tmp <- tempfile(fileext = ".spz")
writeBin(pbmc3k, tmp)
counts <- st_read(tmp)
model <- nmf(counts, k = 20, resource = "gpu", seed = 42)
cat("Loss:", model@misc$loss, "\n")
'
```

## Troubleshooting

### GPU Not Detected

```{r eval = FALSE}
gpu_available(force_recheck = TRUE)  # Force re-probe
#> [1] FALSE
```

Common causes:

- **CUDA drivers not installed or outdated**: Run `nvidia-smi` to check.
- **GPU compute capability < 7.0**: Pre-Volta GPUs are not supported.
- **`RcppML_gpu.so` not found**: Check the library search path above.
- **Architecture mismatch**: Library compiled for `sm_90` won't work on a V100 (`sm_70`).

### Precision

GPU computation uses single precision (FP32) for all core operations, providing
optimal throughput on both consumer and data-center GPUs. Double precision (FP64)
is used selectively for numerically sensitive accumulations where precision loss
would affect convergence. Results are numerically equivalent to CPU (FP64) within
floating-point tolerance for all tested workloads.

## Key Takeaways

1. **Same API, different backend**: `resource = "gpu"` is the only change needed.
   All loss functions, regularization, and cross-validation work identically.
2. **5–50× speedups** on large matrices, with highest gains on dense data and
   high rank.
3. **Direct disk-to-GPU** via `sp_read_gpu()` avoids CPU memory bottlenecks for
   very large datasets.
4. **Automatic fallback**: If GPU is unavailable or a feature is CPU-only,
   RcppML falls back silently — no code changes needed.

*See the [StreamPress](streampress.html) vignette for `.spz` file format details,
and the [NMF Fundamentals](nmf-fundamentals.html) vignette for comprehensive NMF
parameter documentation.*
