---
title: "Learning Parts from Images: Visual Feature Extraction with NMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning Parts from Images: Visual Feature Extraction with NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  eval = NOT_CRAN
)
```

## Parts-Based Representation: Why NMF Sees Differently

A foundational insight of Lee & Seung (1999) is that NMF naturally learns
**parts-based representations** of data. While PCA (or SVD) decomposes data into
holistic eigenmodes that combine addition and subtraction, NMF's non-negativity
constraint forces it to build a whole from non-negative parts. For images, this
means:

- **PCA** learns "eigenfaces" — ghostly, holistic patterns where adding and
  subtracting features creates each face.
- **NMF** learns facial "parts" — eyes, noses, mouths, hairlines — that are
  combined additively to reconstruct each face.

This matters beyond aesthetics: parts-based representations are more interpretable,
more robust to occlusion, and better aligned with how biological visual systems
process information.

## Controlled Validation: The Swimmer Dataset

Before applying NMF to real images, we validate with a synthetic dataset where
the ground truth is known. The **swimmer** dataset generates stick-figure images
with independently moving limbs:

```{r swimmer-generate, fig.width=7, fig.height=3.5}
library(RcppML)
library(ggplot2)
library(Matrix)

# Generate swimmer dataset: 256 images with known factors
swimmer <- simulateSwimmer(n_images = 256, style = "stick",
                           noise = 0.01, seed = 42,
                           return_factors = TRUE)

# simulateSwimmer returns images as ROWS (256 x 1024).
# NMF expects features (pixels) as rows, observations (images) as columns.
A_swim <- as.matrix(t(swimmer$A))  # 1024 pixels × 256 images

cat("Data dimensions:", nrow(A_swim), "pixels ×", ncol(A_swim), "images\n")
cat("True rank:", ncol(swimmer$w), "factors\n")
cat("True factors: 4 limbs × 4 positions = 16\n")
cat("(The torso is constant across all images, not a learned factor)\n")

# Visualize sample images
par(mfrow = c(2, 8), mar = c(0.5, 0.5, 0.5, 0.5))
for (i in sample(ncol(A_swim), 16)) {
  img <- matrix(A_swim[, i], 32, 32)
  image(t(img[32:1, ]), axes = FALSE, col = gray.colors(64, start = 1, end = 0))
}
```

### Recovering True Parts

With 16 ground-truth factors (4 limbs × 4 positions each), can NMF recover them?

```{r swimmer-nmf, fig.width=7, fig.height=6}
model_swim <- nmf(A_swim, k = 16, L1 = c(0.2, 0),
                  tol = 1e-6, maxit = 300, seed = 42, verbose = FALSE)

# Visualize learned factors — each should show one limb position
par(mfrow = c(4, 4), mar = c(0.5, 0.5, 1.5, 0.5))
for (j in 1:16) {
  img <- matrix(model_swim@w[, j], 32, 32)
  image(t(img[32:1, ]), axes = FALSE,
        col = gray.colors(64, start = 1, end = 0),
        main = paste0("F", j), cex.main = 0.9)
}
```

Each factor should correspond to a single limb position or the torso — demonstrating
that NMF correctly identifies the compositional structure of the images.

### Reconstruction Quality

```{r swimmer-recon, fig.width=7, fig.height=3}
recon <- prod(model_swim)

par(mfrow = c(2, 8), mar = c(0.5, 0.5, 1.2, 0.5))
for (i in 1:8) {
  img <- matrix(A_swim[, i], 32, 32)
  image(t(img[32:1, ]), axes = FALSE,
        col = gray.colors(64, start = 1, end = 0), main = "Original", cex.main = 0.7)
}
for (i in 1:8) {
  img <- matrix(recon[, i], 32, 32)
  image(t(img[32:1, ]), axes = FALSE,
        col = gray.colors(64, start = 1, end = 0), main = "NMF", cex.main = 0.7)
}
```

```{r swimmer-mse}
cat("Reconstruction MSE:", evaluate(model_swim, A_swim), "\n")
```

## Ground Truth Validation with Synthetic NMF

The `simulateNMF()` function generates data from a known $W$, $H$ decomposition,
allowing us to measure how accurately NMF recovers the true factors:

```{r simulate-nmf}
# Create ground truth with sparse factors for identifiable recovery
set.seed(42)
k_true <- 5; m <- 200; n <- 150
W_true <- matrix(0, m, k_true)
H_true <- matrix(0, k_true, n)
for (j in 1:k_true) {
  # Each factor uses ~30 features and ~25 samples (sparse, non-overlapping)
  w_idx <- ((j - 1) * 40 + 1):min(j * 40, m)
  h_idx <- ((j - 1) * 30 + 1):min(j * 30, n)
  W_true[w_idx, j] <- runif(length(w_idx), 0.5, 1.5)
  H_true[j, h_idx] <- runif(length(h_idx), 0.5, 1.5)
}
A_sim <- W_true %*% H_true + abs(matrix(rnorm(m * n, 0, 0.01), m, n))
cat("Simulated data:", m, "×", n, "with true rank", k_true, "\n")
cat("True W sparsity:", round(mean(W_true == 0), 2), "\n")

# Fit NMF and compare
model_sim <- nmf(A_sim, k = k_true, tol = 1e-6, maxit = 300, seed = 42, verbose = FALSE)

# Align recovered factors to true factors using bipartite matching
cos_raw <- cosine(W_true, model_sim@w)
cost <- 1 - abs(cos_raw)
match_result <- bipartiteMatch(cost)
model_w_aligned <- model_sim@w[, match_result$assignment + 1]  # 0-indexed

cos_aligned <- cosine(W_true, model_w_aligned)
cat("\nCosine similarity (true vs recovered, after alignment):\n")
cat("  Diagonal:", round(diag(cos_aligned), 3), "\n")
cat("  Mean diagonal:", round(mean(diag(cos_aligned)), 3), "\n")
```

With sparse, non-overlapping ground truth factors, NMF achieves high cosine
similarity after alignment — confirming that the algorithm correctly identifies
the true compositional structure. The `bipartiteMatch()` function solves the
assignment problem to find the best one-to-one correspondence between true and
recovered factors.

## Real Images: Olivetti Faces

The **Olivetti** dataset contains 400 grayscale face images (64×64 pixels) from
40 subjects, 10 images per subject, with varying lighting, expressions, and
minor pose changes:

```{r olivetti-load}
data(olivetti)
img_shape <- attr(olivetti, "image_shape")
subjects <- attr(olivetti, "subject")

cat("Dimensions:", nrow(olivetti), "images ×", ncol(olivetti), "pixels\n")
cat("Image shape:", img_shape, "\n")
cat("Subjects:", nlevels(subjects), "\n")
```

Note: in this dataset, **rows are images** and **columns are pixels**. We'll
transpose for NMF so that columns are observations (images).

```{r olivetti-show, fig.width=7, fig.height=3.5}
# Show sample faces (one per subject)
par(mfrow = c(2, 8), mar = c(0.5, 0.5, 0.5, 0.5))
show_subs <- seq(1, 40, length.out = 16)
for (s in show_subs) {
  idx <- which(subjects == levels(subjects)[s])[1]
  img <- matrix(olivetti[idx, ], img_shape[1], img_shape[2])
  image(t(img[img_shape[1]:1, ]), axes = FALSE,
        col = gray.colors(64, start = 0, end = 1))
}
```

### Learning Facial Parts

We decompose the transposed matrix so each column is a flattened face image:

```{r olivetti-nmf}
# Transpose: pixels × images for NMF
A_faces <- t(olivetti)

model_faces <- nmf(A_faces, k = 25, L1 = c(0.2, 0),
                   tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)
```

### Convergence Analysis

The multi-panel convergence plot reveals how the factorization stabilized:

```{r convergence-plot, fig.width=7, fig.height=4}
plot(model_faces, type = "convergence")
```

### Visualizing Learned Parts

```{r face-parts, fig.width=7, fig.height=6}
par(mfrow = c(5, 5), mar = c(0.5, 0.5, 1.5, 0.5))
for (j in 1:25) {
  img <- matrix(model_faces@w[, j], img_shape[1], img_shape[2])
  image(t(img[img_shape[1]:1, ]), axes = FALSE,
        col = gray.colors(64, start = 0, end = 1),
        main = paste0("Factor ", j), cex.main = 0.8)
}
```

Unlike PCA eigenfaces, these factors should show localized facial features: some
capture eye regions, others nose/mouth areas, hairlines, or facial contours. The
non-negativity constraint prevents cancellation, forcing each factor to contribute
additively.

## The Role of Regularization

Regularization profoundly affects what NMF learns. Let's compare:

### L1 Sparsity: Sharper Parts

L1 regularization pushes small values to zero, creating sparser factors with
crisper localization:

```{r l1-effect, fig.width=7, fig.height=5}
m_sparse <- nmf(A_faces, k = 25, L1 = c(0.5, 0), tol = 1e-5,
                maxit = 200, seed = 42, verbose = FALSE)

# Show regularization diagnostics
plot(m_sparse, type = "regularization")
```

### Sparsity Comparison

```{r sparsity-compare, fig.width=7, fig.height=4}
sp_none <- sparsity(model_faces)
sp_l1 <- sparsity(m_sparse)

sp_none$Regularization <- "No L1"
sp_l1$Regularization <- "L1 = 0.5"
sp_df <- rbind(sp_none, sp_l1)

ggplot(sp_df, aes(x = factor, y = sparsity, fill = Regularization)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  facet_wrap(~model, scales = "free_y",
             labeller = labeller(model = c(w = "W (features)", h = "H (samples)"))) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(title = "Effect of L1 Regularization on Factor Sparsity",
       x = "Factor", y = "Fraction of Zeros") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
```

### L21 Group Sparsity: Automatic Factor Selection

L21 regularization penalizes the $\ell_{2,1}$ norm, encouraging the model to
concentrate its capacity in fewer factors. While RcppML normalizes $W$ columns
to unit norm during fitting, the factor importance is captured in the $d$
vector — L21 drives less important factors' $d$ values toward zero:

```{r l21-effect, fig.width=7, fig.height=4}
# Use strong L21 to demonstrate factor pruning
m_l21 <- nmf(A_faces, k = 25, L21 = c(2.0, 0), tol = 1e-5,
             maxit = 200, seed = 42, verbose = FALSE)

# With L21, factor importance shifts to the d vector
# Factors with small d values are effectively deactivated
d_sorted <- sort(m_l21@d, decreasing = TRUE)
cat("Factor d values (sorted):\n")
cat(round(d_sorted, 1), "\n")

# Count effectively active factors (d > 1% of max)
threshold <- max(m_l21@d) * 0.01
active <- sum(m_l21@d > threshold)
cat("Active factors (d > 1% of max):", active, "of 25\n")

barplot(d_sorted, names.arg = 1:25,
        main = "Factor Importance Under L21 Regularization",
        xlab = "Factor (sorted)", ylab = "d value",
        col = ifelse(d_sorted > threshold, "steelblue", "gray80"),
        border = NA)
abline(h = threshold, lty = 2, col = "red")
```

### Angular Regularization: Diverse Parts

Angular regularization encourages orthogonality between factors, preventing
redundant parts that capture similar features:

```{r angular-effect, fig.width=7, fig.height=3}
m_angular <- nmf(A_faces, k = 25, angular = c(0.3, 0), tol = 1e-5,
                 maxit = 200, seed = 42, verbose = FALSE)

# Measure factor similarity
cos_no_reg <- cosine(model_faces@w)
cos_angular <- cosine(m_angular@w)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
image(abs(cos_no_reg), main = "No Angular Reg",
      col = heat.colors(32), axes = FALSE,
      xlab = "Factor", ylab = "Factor")
image(abs(cos_angular), main = "Angular Reg (0.3)",
      col = heat.colors(32), axes = FALSE,
      xlab = "Factor", ylab = "Factor")
```

With angular regularization, the off-diagonal elements of the cosine similarity
matrix shrink — factors become more distinct from each other.

## Semi-NMF: When Signs Matter

Standard NMF constrains both $W$ and $H$ to be non-negative. But some data naturally
contain both positive and negative components (e.g., centered images, residuals).
**Semi-NMF** relaxes the constraint on one factor:

```{r semi-nmf}
# Center the face data (makes values positive and negative)
A_centered <- as.matrix(A_faces) - rowMeans(as.matrix(A_faces))

# Semi-NMF handles the signed data directly — no need for abs()
# W is unconstrained (can be negative), H remains non-negative
m_semi <- nmf(A_centered, k = 10, nonneg = c(FALSE, TRUE),
              tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)

cat("Input range:", round(range(A_centered), 3), "\n")
cat("W range:", round(range(m_semi@w), 3), "\n")
cat("H range:", round(range(m_semi@h), 3), "(constrained non-negative)\n")
```

In semi-NMF, $W$ factors can have negative entries (representing "anti-features"),
while $H$ loadings remain non-negative (each observation is a non-negative
combination of possibly signed basis elements). Negative values in $W$ capture
regions that are specifically *not* associated with a factor — a richer model
than standard NMF.

## Sparsity Analysis

The sparsity bar chart gives a per-factor view of zero fractions:

```{r sparsity-plot, fig.width=7, fig.height=4}
plot(m_sparse, type = "sparsity")
```

## Key Takeaways

1. **NMF learns parts, not modes**: Unlike PCA, NMF's non-negativity constraint
   produces localized, additive features — facial parts, limb positions, or image
   patches that combine to form wholes.

2. **Ground truth validation** with `simulateSwimmer()` and `simulateNMF()` confirms
   that NMF can exactly recover known compositional structure.

3. **L1 regularization** sharpens parts by driving small contributions to zero,
   producing crisper, more localized features.

4. **L21 group sparsity** shifts factor importance to the $d$ vector,
   concentrating the model's capacity in fewer dominant factors while
   down-weighting less important ones.

5. **Angular regularization** enforces diversity between factors, preventing
   redundant representations.

6. **Semi-NMF** extends NMF to signed data while maintaining non-negative
   combination coefficients.

7. **Convergence diagnostics** (`plot(model, type = "convergence")`) confirm
   that the factorization has stabilized, and regularization diagnostics
   show the contribution of each penalty term.

## Further Reading

- [Getting started with RcppML](getting-started.html) — Basic NMF concepts and regularization overview
- [Cross-validation for rank selection](cross-validation.html) — Detailed CV guide
- [Ecological biodiversity patterns](ecological-biodiversity.html) — Graph regularization for spatial data
- [Cancer subtype discovery](cancer-subtype-discovery.html) — Consensus NMF for robust clustering
