---
title: "NMF Deep Dive: Regularization, Masking, and Model Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF Deep Dive: Regularization, Masking, and Model Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 100,
  eval = NOT_CRAN
)
```

## Introduction

Non-negative matrix factorization (NMF) decomposes a non-negative matrix $A \in \mathbb{R}_{\geq 0}^{m \times n}$ into lower-rank factors:

$$A \approx W \cdot \text{diag}(d) \cdot H$$

where $W \in \mathbb{R}_{\geq 0}^{m \times k}$ contains feature loadings, $H \in \mathbb{R}_{\geq 0}^{k \times n}$ contains sample embeddings, and $d \in \mathbb{R}_{>0}^k$ is a diagonal scaling vector.

The non-negativity constraint ensures that factors can only _add_, never subtract. This produces **parts-based representations** where each factor captures a distinct component of the data. But the basic algorithm is only the starting point. RcppML exposes a comprehensive set of controls — loss functions, regularization penalties, masking strategies, and cross-validation — that shape the factorization to match the structure of your problem.

This vignette walks through these capabilities with synthetic experiments (where ground truth is known) and real demonstrations on built-in datasets. For domain-specific applications, see the links in [Further Reading](#further-reading).

```{r libraries}
library(RcppML)
library(ggplot2)
library(Matrix)
```

## 1. The Optimization Problem

RcppML minimizes a loss function over the non-negative factors:

$$\min_{W \geq 0,\, H \geq 0} \; \mathcal{L}(A, W \cdot \text{diag}(d) \cdot H) + \lambda \cdot \mathcal{R}(W, H)$$

where $\mathcal{L}$ is the reconstruction loss and $\mathcal{R}$ is a regularization penalty.

The algorithm uses **alternating least squares** (ALS): it fixes $H$ and solves for $W$ (as a batch of non-negative least squares problems), then fixes $W$ and solves for $H$. Each NNLS subproblem is solved by fast sequential coordinate descent initialized from an unconstrained Cholesky solution. After each full update cycle, columns of $W$ and rows of $H$ are rescaled to unit norm, with the scale factors absorbed into $d$.

**The diagonal scaling $d$**: After each iteration, RcppML normalizes columns of $W$ and rows of $H$ to unit $\ell_1$ norm (by default), storing the absorbed scale in $d$. This means:

- $W$ and $H$ contain normalized, comparable factor profiles
- $d$ captures the overall "importance" or magnitude of each factor
- Regularization (L1, L2, upper bound) acts on the normalized factors, not the raw product

This design is critical for understanding how penalties interact with the model. For instance, an upper bound constraint of 0.1 on $W$ means that *no single feature can account for more than 10% of a factor's total loading*.

### Convergence

Convergence is measured as the relative change in the loss function between iterations. When the change drops below `tol`, the algorithm terminates:

$$\frac{|\mathcal{L}^{(t)} - \mathcal{L}^{(t-1)}|}{|\mathcal{L}^{(t-1)}| + \epsilon} < \texttt{tol}$$

```{r convergence-demo, fig.width=7, fig.height=5}
# Generate data with known rank
sim <- simulateNMF(nrow = 200, ncol = 150, k = 5, noise = 0.1, seed = 42)
A <- sim$A

# Fit with different tolerances and track convergence
m_tight  <- nmf(A, k = 5, tol = 1e-6, maxit = 500, seed = 1, verbose = FALSE)
m_normal <- nmf(A, k = 5, tol = 1e-4, maxit = 500, seed = 1, verbose = FALSE)
m_loose  <- nmf(A, k = 5, tol = 1e-2, maxit = 500, seed = 1, verbose = FALSE)

cat(sprintf("Tight  (tol=1e-6): %3d iterations, loss = %.6f\n",
            m_tight$iter, m_tight$loss))
cat(sprintf("Normal (tol=1e-4): %3d iterations, loss = %.6f\n",
            m_normal$iter, m_normal$loss))
cat(sprintf("Loose  (tol=1e-2): %3d iterations, loss = %.6f\n",
            m_loose$iter, m_loose$loss))

# Plot per-iteration loss convergence
loss_hist <- m_tight@misc$loss_history
if (!is.null(loss_hist) && length(loss_hist) > 1) {
  df_conv <- data.frame(
    iteration = seq_along(loss_hist),
    loss = loss_hist
  )
  ggplot(df_conv, aes(x = iteration, y = loss)) +
    geom_line(color = "steelblue", linewidth = 0.8) +
    geom_hline(yintercept = m_normal@misc$loss, linetype = "dashed",
               color = "coral", linewidth = 0.5) +
    annotate("text", x = max(df_conv$iteration) * 0.6,
             y = m_normal@misc$loss * 1.02,
             label = sprintf("tol=1e-4 stops at iter %d", m_normal$iter),
             color = "coral", size = 3.2) +
    labs(title = "Convergence: Per-Iteration Loss",
         x = "Iteration", y = "Loss") +
    theme_minimal()
}
```

In practice, `tol = 1e-4` (the default) provides a good balance between accuracy and speed. Tightening to `1e-6` yields marginal loss improvement at substantially higher iteration cost.

## 2. Loss Functions

The choice of loss function determines how the algorithm weighs different types of errors.

### Mean Squared Error (MSE)

The default. Minimizes the Frobenius norm squared:

$$\mathcal{L}_{\text{MSE}} = \frac{1}{mn} \sum_{ij} (A_{ij} - \hat{A}_{ij})^2$$

MSE is solved directly via least squares — no reweighting needed — making it the fastest option.

### Mean Absolute Error (MAE)

Penalizes errors linearly rather than quadratically, providing **robustness to outliers**:

$$\mathcal{L}_{\text{MAE}} = \frac{1}{mn} \sum_{ij} |A_{ij} - \hat{A}_{ij}|$$

MAE is implemented via Iteratively Reweighted Least Squares (IRLS): each element receives a weight inversely proportional to its residual magnitude, so large errors are effectively down-weighted.

### Huber Loss

A smooth interpolation between MSE (for small residuals) and MAE (for large residuals):

$$\mathcal{L}_{\delta}(r) = \begin{cases} \frac{1}{2}r^2 & \text{if } |r| \leq \delta \\ \delta(|r| - \frac{1}{2}\delta) & \text{otherwise} \end{cases}$$

The `huber_delta` parameter controls the transition point. Small $\delta$ behaves like MAE; large $\delta$ behaves like MSE.

### Kullback-Leibler Divergence (KL)

Natural for count data (e.g., scRNA-seq UMI counts):

$$\mathcal{L}_{\text{KL}} = \sum_{ij} \left( A_{ij} \log \frac{A_{ij}}{\hat{A}_{ij}} - A_{ij} + \hat{A}_{ij} \right)$$

### Comparing Loss Functions

```{r loss-comparison, fig.width=7, fig.height=5}
# Add outliers to test robustness
set.seed(99)
A_clean <- sim$A
A_noisy <- as.matrix(A_clean)
outlier_idx <- sample(length(A_noisy), 50)
A_noisy[outlier_idx] <- A_noisy[outlier_idx] + runif(50, 10, 30)
A_noisy <- as(A_noisy, "dgCMatrix")

# Fit with each loss
m_mse   <- nmf(A_noisy, k = 5, loss = "mse",   tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
m_mae   <- nmf(A_noisy, k = 5, loss = "mae",   tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
m_huber <- nmf(A_noisy, k = 5, loss = "huber", huber_delta = 1.0,
               tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)

# Evaluate all on the CLEAN data (how well did each recover the true signal?)
cat("Reconstruction error on clean data (lower = better recovery):\n")
cat(sprintf("  MSE loss:   %.5f\n", evaluate(m_mse,   A_clean)))
cat(sprintf("  MAE loss:   %.5f\n", evaluate(m_mae,   A_clean)))
cat(sprintf("  Huber loss: %.5f\n", evaluate(m_huber, A_clean)))

# Measure factor recovery via cosine similarity to ground truth
recover_quality <- function(model, W_true) {
  cos_mat <- cosine(W_true, model@w)
  cost <- 1 - abs(cos_mat)
  match <- bipartiteMatch(cost)
  mean(1 - cost[cbind(1:nrow(cost), match$assignment + 1)])
}

cat("\nFactor recovery (cosine similarity to ground truth):\n")
cat(sprintf("  MSE loss:   %.4f\n", recover_quality(m_mse,   sim$w)))
cat(sprintf("  MAE loss:   %.4f\n", recover_quality(m_mae,   sim$w)))
cat(sprintf("  Huber loss: %.4f\n", recover_quality(m_huber, sim$w)))
```

When outliers are present, MAE and Huber produce factors that better recover the true signal. The tradeoff: IRLS-based losses are 10-20× slower per iteration than MSE.

## 3. Regularization

Regularization adds penalties to the loss function, trading reconstruction accuracy for desirable structural properties. The total objective is:

$$\text{Total loss} = \underbrace{\mathcal{L}(A, WdH)}_{\text{reconstruction}} + \underbrace{\lambda \cdot \mathcal{R}(W, H)}_{\text{regularization}}$$

A higher reconstruction error in a regularized model is not a failure — it means the penalties are doing their job, imposing the desired structural constraints.

### L1 (LASSO): Encouraging Sparsity

L1 regularization adds an $\ell_1$-norm penalty that drives small factor values to exactly zero:

$$\mathcal{R}_{L_1} = \alpha_w \|W\|_1 + \alpha_h \|H\|_1$$

This produces **sparse factors** — each factor is defined by a small number of features and each sample loads on a small number of factors — improving interpretability and reducing noise.

```{r l1-demo, fig.width=7, fig.height=5}
sim_sparse <- simulateNMF(nrow = 300, ncol = 200, k = 8, noise = 0.05,
                          dropout = 0.5, seed = 42)
A_sp <- sim_sparse$A

# Sweep L1 values
l1_vals <- c(0, 0.01, 0.05, 0.1, 0.3, 0.5)
results <- data.frame()
l1_models <- list()

for (l1 in l1_vals) {
  m <- nmf(A_sp, k = 8, L1 = c(l1, l1), tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
  l1_models[[as.character(l1)]] <- m
  sp <- sparsity(m)
  w_sp <- mean(sp$sparsity[sp$model == "w"])
  h_sp <- mean(sp$sparsity[sp$model == "h"])
  mse_val <- evaluate(m, A_sp)
  results <- rbind(results, data.frame(L1 = l1, W_sparsity = w_sp,
                                       H_sparsity = h_sp, MSE = mse_val))
}

# Sparsity-MSE tradeoff plot
df_trade <- data.frame(
  L1 = rep(results$L1, 2),
  value = c(results$W_sparsity, results$MSE / max(results$MSE)),
  metric = rep(c("W Sparsity", "Relative MSE"), each = nrow(results))
)
ggplot(df_trade, aes(x = L1, y = value, color = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("W Sparsity" = "steelblue", "Relative MSE" = "coral")) +
  labs(title = "L1 Regularization: Sparsity-Accuracy Tradeoff",
       x = "L1 Penalty", y = "Value (sparsity or relative MSE)", color = "") +
  theme_minimal()
```

The plot shows the fundamental tradeoff: as L1 increases, sparsity rises (more zeros in the factors) while reconstruction error increases. Values in the range 0.01–0.1 typically give a good balance. The penalty is applied within the coordinate descent NNLS solver, so it is cheap — there is essentially no runtime cost for using L1.

```{r l1-histograms, fig.width=7, fig.height=6}
# Histograms of W values under three L1 settings
w_list <- list()
for (lbl in c("0", "0.05", "0.3")) {
  w_vals <- as.vector(l1_models[[lbl]]@w)
  w_pos  <- w_vals[w_vals > 0]
  w_list <- c(w_list, list(data.frame(
    value = w_pos,
    L1 = paste0("L1 = ", lbl),
    stringsAsFactors = FALSE
  )))
}
df_w <- do.call(rbind, w_list)

ggplot(df_w, aes(x = value, fill = L1)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_wrap(~ L1, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = c("L1 = 0" = "steelblue",
                                "L1 = 0.05" = "coral",
                                "L1 = 0.3" = "forestgreen")) +
  labs(title = "Distribution of Non-Zero W Values Under L1 Regularization",
       x = "W value", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

# Same for H
h_list <- list()
for (lbl in c("0", "0.05", "0.3")) {
  h_vals <- as.vector(l1_models[[lbl]]@h)
  h_pos  <- h_vals[h_vals > 0]
  h_list <- c(h_list, list(data.frame(
    value = h_pos,
    L1 = paste0("L1 = ", lbl),
    stringsAsFactors = FALSE
  )))
}
df_h <- do.call(rbind, h_list)

ggplot(df_h, aes(x = value, fill = L1)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_wrap(~ L1, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = c("L1 = 0" = "steelblue",
                                "L1 = 0.05" = "coral",
                                "L1 = 0.3" = "forestgreen")) +
  labs(title = "Distribution of Non-Zero H Values Under L1 Regularization",
       x = "H value", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

The histograms reveal L1's mechanism: small values are clipped to exactly zero (creating sparsity), and the surviving non-zero values tend to be larger and more concentrated. Under strong L1, the distribution shifts rightward — the model retains only the most important loadings.

### L2 (Ridge): Shrinkage

L2 regularization penalizes the squared $\ell_2$ norm of factor values:

$$\mathcal{R}_{L_2} = \beta_w \|W\|_F^2 + \beta_h \|H\|_F^2$$

Unlike L1, L2 does not produce exact zeros — it **shrinks** all values toward zero proportionally. This is useful for preventing individual features from dominating a factor and for numerical stability when the problem is ill-conditioned.

```{r l2-demo, fig.width=7, fig.height=5}
l2_vals <- c(0, 0.01, 0.1, 1.0)
l2_models <- list()
results_l2 <- data.frame()

for (l2 in l2_vals) {
  m <- nmf(A_sp, k = 8, L2 = c(l2, l2), tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
  l2_models[[as.character(l2)]] <- m
  max_w <- max(m@w)
  mean_w <- mean(m@w[m@w > 0])
  mse_val <- evaluate(m, A_sp)
  results_l2 <- rbind(results_l2, data.frame(L2 = l2, max_W = max_w,
                                             mean_W = mean_w, MSE = mse_val))
}

cat("Effect of L2 regularization on W factor values:\n")
print(results_l2, row.names = FALSE)

# Histogram comparison under L2
w_l2_list <- list()
for (lbl in c("0", "0.01", "0.1", "1")) {
  w_vals <- as.vector(l2_models[[lbl]]@w)
  w_pos  <- w_vals[w_vals > 0]
  w_l2_list <- c(w_l2_list, list(data.frame(
    value = w_pos,
    L2 = paste0("L2 = ", lbl),
    stringsAsFactors = FALSE
  )))
}
df_l2w <- do.call(rbind, w_l2_list)

ggplot(df_l2w, aes(x = value, fill = L2)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_wrap(~ L2, ncol = 2, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "W Value Distribution Under L2 Regularization",
       subtitle = "L2 compresses values toward zero without producing exact zeros",
       x = "W value", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

L2 is applied by augmenting the diagonal of the Gram matrix: $G(i,i) \mathrel{+}= \beta$. This also improves conditioning, which can help convergence for near-degenerate problems. The histograms show how L2 compresses the distribution of factor values toward smaller magnitudes without introducing exact zeros — in contrast to L1's hard thresholding.

### L21 (Group Sparsity): Adaptive Row-Wise Penalization

L21-norm regularization (group LASSO) applies an **adaptive L2 penalty per row** of the factor matrix. Unlike a uniform L2 penalty, L21 penalizes each row inversely proportional to its norm:

$$G(i,i) \mathrel{+}= \frac{\lambda}{\|W_{i,:}\|_2}$$

Rows with small norm receive a larger penalty (pushing them further toward zero), while rows with large norm receive a smaller penalty (preserving them). This creates a "rich get richer" dynamic: strong factors are preserved and weak factors are suppressed.

Because the $d$ diagonal captures the overall scale of each factor after normalization, L21 affects the $d$ values — concentrating model energy into fewer dominant factors while suppressing minor ones. This makes L21 a tool for **soft rank selection**: fit with a rank higher than expected and let L21 determine how many factors the data supports.

We demonstrate L21 on the Hawaiian birds dataset — a real ecological matrix with meaningful factor structure. By fitting with $k = 12$ (more than the expected number of bird communities), we test whether L21 suppresses unnecessary factors:

```{r l21-demo, fig.width=7, fig.height=5}
data(hawaiibirds)
meta_w <- attr(hawaiibirds, "metadata_w")
meta_h <- attr(hawaiibirds, "metadata_h")

# Fit with k=12 (deliberately over-specified) under different L21 strengths
l21_vals <- c(0, 0.5, 1.0, 2.0, 5.0)
l21_results <- data.frame()
l21_d_list <- list()

for (l21 in l21_vals) {
  m <- nmf(hawaiibirds, k = 12, L21 = c(l21, 0), tol = 1e-5, maxit = 200,
           seed = 42, verbose = FALSE)
  d_sorted <- sort(m@d, decreasing = TRUE)
  l21_results <- rbind(l21_results, data.frame(
    L21 = l21,
    sum_d = sum(m@d),
    max_d = max(m@d),
    min_d = min(m@d),
    ratio = max(m@d) / max(min(m@d), 1e-10),
    MSE = evaluate(m, hawaiibirds)
  ))
  l21_d_list <- c(l21_d_list, list(data.frame(
    factor = 1:12,
    d = d_sorted,
    L21 = paste0("L21 = ", l21)
  )))
}

cat("L21 effect on diagonal scaling (d) and MSE:\n")
print(l21_results[, c("L21", "sum_d", "max_d", "min_d", "ratio", "MSE")], row.names = FALSE)

df_d <- do.call(rbind, l21_d_list)
df_d$L21 <- factor(df_d$L21, levels = paste0("L21 = ", l21_vals))

ggplot(df_d, aes(x = factor, y = d)) +
  geom_col(fill = "steelblue", width = 0.7) +
  facet_wrap(~ L21, ncol = 3, scales = "free_y") +
  labs(title = "L21 Effect on Factor Importance (Hawaiian Birds, k = 12)",
       subtitle = "Sorted diagonal d values — L21 suppresses minor factors",
       x = "Factor (sorted by importance)", y = "d value") +
  theme_minimal()
```

The bar charts show how L21 redistributes model energy. Without L21, the 12 factors have relatively uniform $d$ values. As L21 increases, the gap between dominant and minor factors widens: the `ratio` column (max $d$ / min $d$) quantifies this effect. Strong factors that capture real community structure are preserved, while excess factors are driven toward negligible $d$ values. In practice, you can inspect the $d$ distribution to decide how many factors are truly needed.

### Angular Regularization: Factor Diversity

Angular regularization encourages orthogonality between factor columns by penalizing cross-factor correlations through the Gram matrix:

$$G \mathrel{+}= \eta \cdot (W \cdot W^T - \text{diag})$$

where only the off-diagonal terms penalize overlap between factors. This pushes the model toward solutions where each factor captures a distinct, non-redundant pattern.

```{r angular-demo, fig.width=7, fig.height=6}
K <- 6
ang_vals <- c(0, 0.1, 0.5, 1.0)
results_ang <- data.frame()
ang_models <- list()

for (ang in ang_vals) {
  m <- nmf(hawaiibirds, k = K, angular = c(ang, 0), tol = 1e-5, maxit = 200,
           seed = 42, verbose = FALSE)
  ang_models[[as.character(ang)]] <- m
  cos_mat <- abs(cosine(m@w))
  diag(cos_mat) <- NA
  mean_cos <- mean(cos_mat, na.rm = TRUE)
  max_cos <- max(cos_mat, na.rm = TRUE)
  results_ang <- rbind(results_ang, data.frame(angular = ang, mean_cosine = mean_cos,
                                               max_cosine = max_cos,
                                               MSE = evaluate(m, hawaiibirds)))
}

cat("Effect of angular regularization on factor similarity:\n")
print(results_ang, row.names = FALSE, digits = 4)

# Cosine similarity heatmaps with ggplot2, viridis "B", 1:1 aspect ratio
heat_list <- list()
for (ang_str in names(ang_models)) {
  m <- ang_models[[ang_str]]
  cos_mat <- abs(cosine(m@w))
  for (i in seq_len(nrow(cos_mat))) {
    for (j in seq_len(ncol(cos_mat))) {
      heat_list <- c(heat_list, list(data.frame(
        row = i, col = j,
        cosine = cos_mat[i, j],
        angular = paste0("Angular = ", ang_str)
      )))
    }
  }
}
df_heat <- do.call(rbind, heat_list)
df_heat$angular <- factor(df_heat$angular,
                          levels = paste0("Angular = ", c("0", "0.1", "0.5", "1")))

ggplot(df_heat, aes(x = col, y = row, fill = cosine)) +
  geom_tile() +
  facet_wrap(~ angular, ncol = 2) +
  scale_fill_viridis_c(option = "B", limits = c(0, 1), name = "|cosine|") +
  scale_y_reverse() +
  coord_fixed() +
  labs(title = "Inter-Factor Cosine Similarity Under Angular Regularization",
       subtitle = "Hawaiian Birds (k = 6) — off-diagonal cooling indicates more distinct factors",
       x = "Factor", y = "Factor") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Higher angular penalties produce factors with lower inter-factor cosine similarity. The heatmaps show how the off-diagonal blocks cool (approach zero) as the penalty increases, indicating more distinct, non-redundant factors. This is particularly useful in ecological data like Hawaiian birds, where you want each factor to represent a distinct bird community.

### Upper Bound Constraints

Upper bound constraints clip factor values to a maximum during coordinate descent. Because RcppML normalizes factor columns to unit $\ell_1$ norm after each iteration, a factor value of $x$ in $W$ means that feature accounts for fraction $x$ of the factor's total loading. An upper bound of 0.1 ensures no single feature can dominate — each factor's loading is spread across at least 10 features.

```{r upper-bound-demo, fig.width=7, fig.height=5}
A5 <- sim$A  # 5 true factors
m_unbound <- nmf(A5, k = 5, tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)

cat("Unconstrained W summary:\n")
cat("  Range:", round(range(m_unbound@w), 4), "\n")
cat("  Max:  ", round(max(m_unbound@w), 4), "\n")
cat("  Mean (nonzero):", round(mean(m_unbound@w[m_unbound@w > 0]), 4), "\n\n")

# Use bounds that actually constrain values
bounds <- c(0.3, 0.1, 0.05)
bound_models <- list()
for (b in bounds) {
  bound_models[[as.character(b)]] <- nmf(A5, k = 5, upper_bound = c(b, b),
                                          tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
}

# Superimposed histogram comparison using ggplot2
ub_list <- list()
ub_list <- c(ub_list, list(data.frame(
  value = as.vector(m_unbound@w[m_unbound@w > 0]),
  bound = "Unconstrained"
)))
for (b in bounds) {
  w_b <- bound_models[[as.character(b)]]@w
  ub_list <- c(ub_list, list(data.frame(
    value = as.vector(w_b[w_b > 0]),
    bound = paste0("UB = ", b)
  )))
}
df_ub <- do.call(rbind, ub_list)
df_ub$bound <- factor(df_ub$bound,
                       levels = c("Unconstrained", paste0("UB = ", bounds)))

ggplot(df_ub, aes(x = value, fill = bound)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_wrap(~ bound, ncol = 2, scales = "free_y") +
  geom_vline(data = data.frame(
    bound = factor(paste0("UB = ", bounds),
                   levels = c("Unconstrained", paste0("UB = ", bounds))),
    xint = bounds
  ), aes(xintercept = xint), linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "W Value Distribution Under Upper Bound Constraints",
       x = "W value", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r upper-bound-table}
# d values comparison
cat("Diagonal scaling (d) under upper bounds:\n")
d_df <- data.frame(factor = 1:5, unconstrained = sort(m_unbound@d, decreasing = TRUE))
for (b in bounds) {
  d_df[[paste0("UB_", b)]] <- sort(bound_models[[as.character(b)]]@d, decreasing = TRUE)
}
print(d_df, row.names = FALSE, digits = 2)

# MSE comparison
cat("\nReconstruction MSE:\n")
cat(sprintf("  Unconstrained: %.5f\n", evaluate(m_unbound, A5)))
for (b in bounds) {
  cat(sprintf("  Upper bound = %s: %.5f\n", b, evaluate(bound_models[[as.character(b)]], A5)))
}
```

When the upper bound clips values, the distribution shifts: peak loadings are capped and the model compensates by distributing weight more evenly across features. The dashed red lines show the bound. The $d$ values may increase to compensate for the restricted per-element range. This is useful when you want factors with broad, distributed support rather than concentrated on a few features.

### Combining Regularizations

Regularization types can be combined. A common pattern is L1 for sparsity plus L2 for numerical stability:

```{r combined-reg, fig.width=7, fig.height=5}
m_combined <- nmf(A_sp, k = 8, L1 = c(0.1, 0.1), L2 = c(0.01, 0.01),
                  tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
m_l1_only <- nmf(A_sp, k = 8, L1 = c(0.1, 0.1),
                 tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
m_none    <- nmf(A_sp, k = 8, tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)

# MSE comparison
mse_vals <- data.frame(
  config = factor(c("None", "L1 only", "L1 + L2"),
                  levels = c("None", "L1 only", "L1 + L2")),
  MSE = c(evaluate(m_none, A_sp), evaluate(m_l1_only, A_sp), evaluate(m_combined, A_sp))
)

# Sparsity comparison
sp_none <- sparsity(m_none)
sp_l1 <- sparsity(m_l1_only)
sp_comb <- sparsity(m_combined)

sp_vals <- data.frame(
  config = factor(c("None", "L1 only", "L1 + L2"),
                  levels = c("None", "L1 only", "L1 + L2")),
  sparsity = c(
    mean(sp_none$sparsity[sp_none$model == "w"]),
    mean(sp_l1$sparsity[sp_l1$model == "w"]),
    mean(sp_comb$sparsity[sp_comb$model == "w"])
  )
)

df_comb <- merge(mse_vals, sp_vals, by = "config")
df_comb_long <- data.frame(
  config = rep(df_comb$config, 2),
  metric = rep(c("Reconstruction MSE", "W Sparsity"), each = nrow(df_comb)),
  value = c(df_comb$MSE / max(df_comb$MSE), df_comb$sparsity)
)

ggplot(df_comb_long, aes(x = config, y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Reconstruction MSE" = "coral", "W Sparsity" = "steelblue")) +
  labs(title = "Combining Regularizations: MSE vs. Sparsity",
       subtitle = "Regularized models trade reconstruction accuracy for structure",
       x = "", y = "Normalized Value", fill = "") +
  theme_minimal()
```

The regularized models have higher reconstruction error but sparser, more interpretable factors. The algorithm converges to a solution that minimizes the _total_ objective (reconstruction loss + penalty terms), not the reconstruction error alone.

## 4. Masking: Handling Missing and Special Data

NMF can handle incomplete matrices via masking — ignoring certain entries during optimization.

### NA Masking

If your data contains `NA` entries (genuinely missing observations), set `mask = "NA"`:

```{r na-mask-demo, fig.width=7, fig.height=4}
# Introduce NAs into the data
A5 <- sim$A
A_na <- as.matrix(A5)
na_frac <- 0.15
na_idx <- sample(length(A_na), floor(length(A_na) * na_frac))
A_na[na_idx] <- NA

cat(sprintf("Matrix has %.1f%% NAs (%d entries)\n", na_frac * 100, length(na_idx)))

# Fit with NA masking
m_na <- nmf(A_na, k = 5, mask = "NA", tol = 1e-5, maxit = 200, seed = 1, verbose = FALSE)
cat(sprintf("Converged in %d iterations\n", m_na$iter))

# Check that the model recovers the structure
cat(sprintf("Factor recovery: %.4f\n", recover_quality(m_na, sim$w)))
```

Masked entries do not contribute to the loss or gradients. The model learns from the observed entries and implicitly imputes the missing values in the reconstruction $W \cdot \text{diag}(d) \cdot H$.

### Zero Masking (`mask_zeros = TRUE`)

For data where zeros represent **missing observations** rather than true zero values — as in recommendation systems where a zero rating means "not rated" rather than "rated zero stars" — set `mask_zeros = TRUE`:

```{r sparse-mask-demo}
# MovieLens: zeros are "not rated", not "zero stars"
data(movielens)
cat("MovieLens: ", nrow(movielens), "movies x", ncol(movielens), "users\n")
cat("Sparsity:", round(1 - nnzero(movielens) / prod(dim(movielens)), 4), "\n")
cat("(Zeros represent unobserved ratings)\n\n")

# Compare with and without zero masking
m_dense  <- nmf(movielens, k = 8, mask_zeros = FALSE, tol = 1e-5, maxit = 100,
                seed = 42, verbose = FALSE)
m_sparse <- nmf(movielens, k = 8, mask_zeros = TRUE, tol = 1e-5, maxit = 100,
                seed = 42, verbose = FALSE)

# Compute RMSE on observed entries
A_dense_mat <- as.matrix(movielens)
obs_mask <- A_dense_mat != 0

recon_d <- prod(m_dense)
recon_s <- prod(m_sparse)

rmse_d <- sqrt(mean((A_dense_mat[obs_mask] - recon_d[obs_mask])^2))
rmse_s <- sqrt(mean((A_dense_mat[obs_mask] - recon_s[obs_mask])^2))

cat(sprintf("RMSE on observed ratings:\n"))
cat(sprintf("  mask_zeros=FALSE: %.3f\n", rmse_d))
cat(sprintf("  mask_zeros=TRUE:  %.3f\n", rmse_s))
```

With `mask_zeros = TRUE`, the model focuses exclusively on reconstructing the observed entries, typically yielding better predictions for recommendation tasks.

### Custom Masking

For more complex missing-data patterns, pass an explicit mask matrix. The mask should be the same dimensions as the data, with non-zero entries marking values to exclude:

```{r custom-mask-demo}
# Mask the top 5% of values (e.g., suspected measurement artifacts)
A_dense <- as.matrix(A5)
threshold <- quantile(A_dense[A_dense > 0], 0.95)
mask_mat <- Matrix(1 * (A_dense > threshold), sparse = TRUE)
cat(sprintf("Custom mask: %d entries (%.1f%% of non-zeros)\n",
            nnzero(mask_mat), 100 * nnzero(mask_mat) / sum(A_dense > 0)))

m_masked <- nmf(A5, k = 5, mask = mask_mat, tol = 1e-5, maxit = 200,
                seed = 1, verbose = FALSE)
cat(sprintf("Factor recovery with masking: %.4f\n", recover_quality(m_masked, sim$w)))
```

## 5. Semi-NMF: Relaxing the Non-Negativity Constraint

Standard NMF requires $W \geq 0$ and $H \geq 0$. **Semi-NMF** relaxes one or both constraints, enabling factorization of data with both positive and negative values:

```{r semi-nmf-demo, fig.width=7, fig.height=4}
# Center the data (introduces negative values)
A_centered <- as.matrix(A5) - rowMeans(as.matrix(A5))
cat("Centered data range:", round(range(A_centered), 3), "\n")

# Standard NMF would clip negatives. Semi-NMF handles them directly.
m_semi <- nmf(A_centered, k = 5, nonneg = c(FALSE, TRUE), tol = 1e-5,
              maxit = 200, seed = 1, verbose = FALSE)

cat("W range:", round(range(m_semi@w), 3), " (unconstrained)\n")
cat("H range:", round(range(m_semi@h), 3), " (non-negative)\n")

# With nonneg = c(FALSE, TRUE):
#   W can have negative values -> captures "anti-features"
#   H remains non-negative -> each sample is a non-negative mixture of basis elements
```

Semi-NMF is useful for centered data, log-fold-change matrices, and any setting where features can be both positively and negatively associated with a factor.

## 6. Cross-Validation for Rank Selection

The most important hyperparameter in NMF is the rank $k$. Cross-validation provides a principled, data-driven way to select it.

### The Speckled Holdout Approach

RcppML uses **speckled masking**: a random subset of individual matrix entries is held out (set to zero during training). After fitting, the model's ability to predict these unseen entries is measured. Unlike row/column holdout, speckled holdout:

- Preserves the global structure of the matrix
- Tests prediction across all features and samples simultaneously
- Avoids the "cold start" problem of holding out entire samples

### Convergence and Stopping Criteria

Two parameters control when training stops for each rank:

- **`maxit`**: The maximum number of iterations. Should be set high enough to allow full convergence.
- **`patience`**: Early stopping for cross-validation. If test loss does not improve for `patience` consecutive iterations, training stops — this prevents wasting computation on iterations that overfit the training data but degrade test performance.

Setting `patience` too aggressively (small values) can terminate training prematurely before the model has found a good factorization. More generous patience values allow the model to explore longer, making overfitting patterns more visible in the test loss curves.

### Rank Sweep

We demonstrate with the Hawaiian birds dataset — a sparse ecological matrix where the target rank (number of bird communities) has an ecological interpretation:

```{r cv-rank-sweep, fig.width=7, fig.height=5}
# Cross-validate across ranks 2-25 with 3 replicates
# Higher-rank models test whether overfitting emerges
cv <- nmf(hawaiibirds, k = 2:25, test_fraction = 0.2, cv_seed = 1:3,
          tol = 1e-6, maxit = 200, patience = 20, verbose = FALSE)
plot(cv) + ggtitle("Hawaiian Birds: Cross-Validation for Optimal Rank")
```

The plot shows test MSE vs. rank. Multiple replicates (different `cv_seed` values) provide confidence intervals. With sufficient `patience` and a wide rank range (up to 25), higher-rank models may show test loss increasing after an initial decrease — a clear signature of overfitting that identifies the ideal rank.

```{r cv-optimal}
mean_test <- tapply(cv$test_mse, cv$k, mean)
optimal_k <- as.integer(names(which.min(mean_test)))
cat("Optimal rank:", optimal_k, "\n")
cat("Test MSE at optimal k:", round(min(mean_test), 5), "\n")
```

### Test Fraction Selection

The `test_fraction` parameter controls how much data is held out. Larger fractions give more stable estimates but remove more training signal:

```{r test-fraction, fig.width=7, fig.height=5}
# Compare different holdout fractions — including a large one to test stability
fracs <- c(0.05, 0.20, 0.50)
cv_results <- list()
for (frac in fracs) {
  cv_f <- nmf(hawaiibirds, k = 2:20, test_fraction = frac, cv_seed = 42,
              tol = 1e-6, maxit = 200, patience = 20, verbose = FALSE)
  cv_f$fraction <- frac
  cv_results[[as.character(frac)]] <- cv_f
}

cv_all <- do.call(rbind, cv_results)

ggplot(cv_all, aes(x = k, y = test_mse, color = factor(fraction))) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(values = c("0.05" = "steelblue", "0.2" = "coral", "0.5" = "forestgreen")) +
  labs(title = "Effect of Test Fraction on Cross-Validation Curves",
       subtitle = "Hawaiian Birds — larger holdout fractions are more stable but noisier",
       color = "Test Fraction", x = "Rank (k)", y = "Test MSE") +
  theme_minimal()
```

At `test_fraction = 0.5`, half the data is held out — this severely limits training signal and can make the CV curve noisy, but if the optimal rank is robust to this stress test, you can be confident in the result. For most datasets, 10–20% is the practical sweet spot.

### Sparse vs. Dense Test Sets

The `mask_zeros` parameter in cross-validation controls whether zero entries are included in the test set:

- `mask_zeros = FALSE` (default): The test set includes both zero and non-zero entries. The model is evaluated on full matrix reconstruction. Appropriate for dense data where zeros are informative.

- `mask_zeros = TRUE`: The test set contains only non-zero entries. The model is evaluated on predicting observed values. Appropriate for recommendation data where zeros mean "unobserved."

```{r sparse-cv, fig.width=7, fig.height=5}
# MovieLens: compare sparse vs dense test set semantics
cv_dense  <- nmf(movielens, k = 2:15, test_fraction = 0.2, mask_zeros = FALSE,
                 cv_seed = 42, tol = 1e-5, maxit = 80, verbose = FALSE)
cv_sparse <- nmf(movielens, k = 2:15, test_fraction = 0.2, mask_zeros = TRUE,
                 cv_seed = 42, tol = 1e-5, maxit = 80, verbose = FALSE)

# Faceted comparison
cv_dense$mode <- "Dense (all entries)"
cv_sparse$mode <- "Sparse (non-zeros only)"
cv_both <- rbind(cv_dense, cv_sparse)

ggplot(cv_both, aes(x = k, y = test_mse)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 2) +
  facet_wrap(~ mode, scales = "free_y") +
  labs(title = "Cross-Validation: Dense vs. Sparse Test Sets (MovieLens)",
       subtitle = "Different test set definitions produce different curve shapes and scales",
       x = "Rank (k)", y = "Test MSE") +
  theme_minimal()
```

The absolute MSE values differ substantially between the two modes because they have different test sets and denominators. The _shape_ of the curve may also differ, reflecting different definitions of "good fit." For recommendation data, `mask_zeros = TRUE` is the correct choice since it evaluates the model on the task it actually performs — predicting observed ratings.

### Automatic Rank Selection

For convenience, RcppML can search for the optimal rank automatically:

```{r auto-rank, eval=FALSE}
# Binary search over [2, 30]
auto_model <- nmf(hawaiibirds, k = "auto", cv_k_range = c(2, 30),
                  test_fraction = 0.2, cv_seed = 42, verbose = TRUE)
cat("Selected rank:", auto_model$rank_search$k_optimal, "\n")
```

The auto search uses a bisection strategy to efficiently narrow the rank range without evaluating every integer.

## 7. Graph Regularization: Incorporating Prior Knowledge

When features or samples have known relationships (spatial proximity, gene networks, taxonomic trees), graph Laplacian regularization encourages connected nodes to have similar factor loadings:

$$\mathcal{R}_{\text{graph}} = \lambda \cdot \text{tr}(H \cdot L_H \cdot H^T)$$

where $L_H$ is the graph Laplacian of the sample adjacency matrix.

### Building a Spatial Graph

The Hawaiian birds dataset includes geographic coordinates for each survey site. We construct a within-island adjacency graph connecting nearby sites:

```{r graph-setup}
# Build spatial adjacency graph for Hawaiian bird survey sites
coords <- cbind(meta_h$lng, meta_h$lat)
n_sites <- nrow(coords)
threshold <- 0.1  # ~11 km

i_vec <- integer(0)
j_vec <- integer(0)
for (isl in levels(meta_h$island)) {
  idx <- which(meta_h$island == isl)
  if (length(idx) < 2) next
  d <- as.matrix(dist(coords[idx, ]))
  pairs <- which(d > 0 & d < threshold, arr.ind = TRUE)
  if (nrow(pairs) > 0) {
    i_vec <- c(i_vec, idx[pairs[, 1]])
    j_vec <- c(j_vec, idx[pairs[, 2]])
  }
}

graph_H <- sparseMatrix(i = i_vec, j = j_vec, x = 1,
                        dims = c(n_sites, n_sites))
graph_H <- as((graph_H + t(graph_H)) / 2, "dgCMatrix")
cat("Spatial graph:", n_sites, "nodes,", nnzero(graph_H) / 2, "edges\n")
```

### Comparing Standard vs. Graph-Regularized NMF

We test multiple graph regularization strengths, including strong values ($\lambda = 2.0$) to make the effect clearly visible:

```{r graph-compare, fig.width=7, fig.height=6}
graph_lambdas <- c(0, 0.1, 0.5, 2.0)
graph_models <- list()
graph_results <- data.frame()

for (lam in graph_lambdas) {
  m <- nmf(hawaiibirds, k = K, L1 = c(0.1, 0),
           graph_H = if (lam > 0) graph_H else NULL,
           graph_lambda = c(0, lam),
           tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)
  graph_models[[as.character(lam)]] <- m
  
  # Measure spatial smoothness: mean squared difference between connected nodes
  h_mat <- m@h
  edges <- which(graph_H != 0, arr.ind = TRUE)
  edges <- edges[edges[,1] < edges[,2], , drop = FALSE]  # unique edges
  
  if (nrow(edges) > 0) {
    sq_diff <- rowMeans(sapply(seq_len(K), function(f) {
      (h_mat[f, edges[,1]] - h_mat[f, edges[,2]])^2
    }))
    smoothness <- mean(sq_diff)
  } else {
    smoothness <- NA
  }
  
  graph_results <- rbind(graph_results, data.frame(
    lambda = lam,
    MSE = evaluate(m, hawaiibirds),
    smoothness = smoothness
  ))
}

cat("Graph regularization: MSE vs. spatial smoothness\n")
cat("(Lower smoothness = more spatially coherent factors)\n\n")
print(graph_results, row.names = FALSE, digits = 5)

# Plot tradeoff
df_graph <- data.frame(
  lambda = rep(graph_results$lambda, 2),
  value = c(graph_results$MSE / max(graph_results$MSE),
            graph_results$smoothness / max(graph_results$smoothness, na.rm = TRUE)),
  metric = rep(c("Reconstruction MSE (normalized)", "Spatial Roughness (normalized)"),
               each = nrow(graph_results))
)

ggplot(df_graph, aes(x = factor(lambda), y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Reconstruction MSE (normalized)" = "coral",
                                "Spatial Roughness (normalized)" = "steelblue")) +
  labs(title = "Graph Regularization: Reconstruction vs. Smoothness Tradeoff",
       subtitle = "Higher lambda produces smoother factors at the cost of reconstruction accuracy",
       x = expression("Graph"~lambda), y = "Normalized Value", fill = "") +
  theme_minimal()
```

### Spatial Maps

```{r graph-maps, fig.width=7, fig.height=8}
# Spatial visualization: compare standard vs. strong graph regularization
m_std <- graph_models[["0"]]
m_graph <- graph_models[["2"]]

# Find the two factors with the most spatial variation
h_var_std <- apply(m_std@h, 1, var)
show_factors <- order(h_var_std, decreasing = TRUE)[1:2]

map_list <- list()
for (j in show_factors) {
  for (info in list(list(m_std, "Standard NMF"), list(m_graph, "Graph NMF (lambda = 2)"))) {
    model <- info[[1]]
    label_text <- info[[2]]
    map_list <- c(map_list, list(data.frame(
      lng = meta_h$lng,
      lat = meta_h$lat,
      loading = model@h[j, ],
      model = label_text,
      factor_label = paste0("Factor ", j)
    )))
  }
}
df_map <- do.call(rbind, map_list)

ggplot(df_map, aes(x = lng, y = lat, color = loading)) +
  geom_point(size = 0.6) +
  facet_grid(factor_label ~ model) +
  scale_color_viridis_c(option = "B", name = "Loading") +
  coord_fixed() +
  labs(title = "Spatial Factor Loadings: Standard vs. Graph-Regularized NMF",
       subtitle = "Graph regularization produces smoother, more spatially contiguous patterns",
       x = "Longitude", y = "Latitude") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The spatial maps show how graph regularization produces smoother, more spatially contiguous factor loadings. Without it, nearby sites can have very different factor values; with it, communities transition gradually across geographic space. The **smoothness** metric quantifies this: it measures the mean squared difference in factor loadings between connected sites — lower values indicate more spatially coherent patterns. The tradeoff bar chart shows that stronger graph regularization increases reconstruction MSE while decreasing spatial roughness.

## 8. Practical Guidelines

### Choosing Regularization

| Goal | Regularization | Typical Values |
|:-----|:---------------|:---------------|
| Sparse, interpretable factors | L1 | 0.01 – 0.3 |
| Numerical stability | L2 | 0.001 – 0.1 |
| Soft factor importance shrinkage | L21 | 0.5 – 5.0 |
| Diverse, non-redundant factors | Angular | 0.1 – 1.0 |
| Bounded factor values | Upper bound | Problem-specific |
| Spatially smooth factors | Graph | 0.01 – 2.0 |

### Choosing a Loss Function

| Data Type | Recommended Loss | Reason |
|:----------|:----------------|:-------|
| Clean data, no outliers | MSE | Fastest, optimal for Gaussian noise |
| Data with outliers | MAE or Huber | Robust to extreme values |
| Count data (scRNA-seq) | KL | Matches Poisson generative model |
| Mixed / uncertain | Huber | Good default for robustness |

### Choosing Test Set Semantics

| Data Type | `sparse` Setting | Reason |
|:----------|:----------------|:-------|
| Dense matrices (images, expression) | `FALSE` | Zeros are real values |
| Ratings / interactions | `TRUE` | Zeros mean "unobserved" |
| Sparse counts (scRNA-seq) | Either | `FALSE` for reconstruction, `TRUE` for imputation |

### General Workflow

1. **Start simple**: Fit with no regularization to establish a baseline.
2. **Select rank** via cross-validation (`k = 2:K_max, test_fraction = 0.1-0.2`).
3. **Add L1** if you need interpretable, sparse factors.
4. **Add L2** if you encounter numerical instability or ill-conditioned problems.
5. **Try L21** as a soft penalty on factor importance.
6. **Use angular** if factors appear redundant (high cosine similarity).
7. **Validate** with `evaluate()`, `sparsity()`, and `summary(group_by = ...)`.

## Further Reading {#further-reading}

- [Getting started with RcppML](getting-started.html) — Quick-start guide and basic API
- [Cross-validation for rank selection](cross-validation.html) — Detailed CV methodology
- [Image decomposition](image-decomposition.html) — Parts-based visual feature extraction
- [Ecological biodiversity](ecological-biodiversity.html) — Graph regularization for spatial data
- [Building a recommendation engine](recommendation-systems.html) — Sparse masking for ratings
- [Cancer subtype discovery](cancer-subtype-discovery.html) — Consensus NMF for robust clustering
- [Discovering cell types from scRNA-seq](single-cell-genomics.html) — NMF for genomics
- [GPU acceleration](gpu-acceleration.html) — Using CUDA for large-scale NMF

## References

- Lee, D.D. & Seung, H.S. (1999). Learning the parts of objects by non-negative matrix factorization. *Nature*, 401(6755), 788-791.
- Lee, D.D. & Seung, H.S. (2001). Algorithms for non-negative matrix factorization. *Advances in Neural Information Processing Systems*, 13.
- Kim, H. & Park, H. (2007). Sparse non-negative matrix factorizations via alternating non-negativity-constrained least squares for microarray data analysis. *Bioinformatics*, 23(12), 1495-1502.
- Cai, D., He, X., Han, J., & Huang, T.S. (2011). Graph regularized nonnegative matrix factorization for data representation. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 33(8), 1548-1560.
- DeBruine, Z.J., Melcher, K., & Triche, T.J. (2021). High-performance non-negative matrix factorization for large single-cell data. *BioRXiv*.
