---
title: "Building a Movie Recommendation Engine with NMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Movie Recommendation Engine with NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  eval = NOT_CRAN
)
```

## Collaborative Filtering with Matrix Factorization

Recommendation systems power some of the most impactful applications of machine
learning — from Netflix suggestions to Spotify playlists. The core idea behind
**collaborative filtering** is simple: users who agreed in the past will agree in the
future.

Mathematically, we observe a sparse ratings matrix $A$ where entry $A_{ij}$ is user
$j$'s rating of movie $i$ (or zero if unrated). NMF decomposes this into:

$$A \approx W \cdot \text{diag}(d) \cdot H$$

where:

- $W$ (movies × k): **movie factors** — latent attributes of each movie (e.g., "action
  content", "romantic tone", "critical acclaim")
- $H$ (k × users): **user factors** — each user's affinity for these latent attributes
- $d$: relative importance of each latent dimension

To predict whether user $j$ would enjoy movie $i$, we simply compute the dot product
of their latent representations: $\hat{A}_{ij} = (W \cdot \text{diag}(d) \cdot H)_{ij}$.

## The MovieLens Dataset

The `movielens` dataset contains 100,000 ratings from 610 users across 3,867 movies:

```{r load-data}
library(RcppML)
library(ggplot2)
library(Matrix)

data(movielens)
cat("Dimensions:", nrow(movielens), "movies ×", ncol(movielens), "users\n")
cat("Non-zero ratings:", nnzero(movielens), "\n")
cat("Sparsity:", round(1 - nnzero(movielens) / prod(dim(movielens)), 4), "\n")
cat("Rating range:", range(movielens@x), "\n")

# Genre metadata
genres <- attr(movielens, "genres")
cat("\nGenre matrix:", nrow(genres), "genres ×", ncol(genres), "movies\n")
cat("Genres:", rownames(genres), "\n")
```

This matrix is over 95% sparse — each user has rated only a small fraction of
available movies. This extreme sparsity is what makes recommendation challenging
and interesting.

### Rating Distribution

```{r rating-dist, fig.width=6, fig.height=3.5}
hist(movielens@x, breaks = seq(0.25, 5.25, 0.5), col = "coral",
     border = "white", main = "Distribution of Ratings",
     xlab = "Rating (stars)", ylab = "Count")
```

Ratings are discrete (0.5 to 5 stars) with a left skew — users tend to rate movies
they like, creating selection bias that the model must account for.

## The Prediction Challenge: Sparse Cross-Validation

For recommendation, the key question is: **how well can we predict ratings a user
*would* give to movies they haven't seen?**

Standard cross-validation holds out random entries, but the semantics matter:

- `mask_zeros = TRUE`: Only existing ratings are split into train/test. The model never
  tries to predict zeros (unobserved entries). This matches the recommendation task.
- `mask_zeros = FALSE`: All entries (including zeros) can be in the test set. This
  measures full matrix reconstruction.

For recommendation, `mask_zeros = TRUE` is the correct choice:

```{r cross-validation, fig.width=7, fig.height=4}
cv <- nmf(movielens, k = 2:20, test_fraction = 0.2, mask_zeros = TRUE,
          tol = 1e-5, maxit = 100, verbose = FALSE)
plot(cv) + ggtitle("Cross-Validation: Predicting Held-Out Ratings")
```

The test error curve reveals the optimal complexity for this rating matrix — too
few factors underfit (can't capture genre nuances), too many overfit (memorize
noise in ratings).

```{r optimal-rank}
best_k <- cv$k[which.min(cv$test_mse)]
cat("Optimal rank:", best_k, "\n")
```

## Fitting the Recommendation Model

We fit the final model with regularization appropriate for the recommendation setting:

- **L1 on H** (user factors): Encourages each user to be described by a few dominant
  taste dimensions, reflecting that most people have a handful of strong preferences.
- **L2 on W** (movie factors): Prevents any single latent dimension from dominating,
  producing more balanced factor representations.

```{r fit-model}
model <- nmf(movielens, k = best_k, L1 = c(0, 0.3), L2 = c(0.1, 0),
             tol = 1e-6, maxit = 300, seed = 42, verbose = FALSE)
model
```

## Comparing Loss Functions

Ratings data can be noisy — users occasionally give ratings that don't reflect
their typical preferences. Different loss functions handle this differently:

- **MSE** (mean squared error): The standard choice, penalizing errors quadratically.
  Sensitive to outliers but produces the minimum-variance estimate.
- **MAE** (mean absolute error): Penalizes errors linearly, potentially more robust
  to extreme ratings but may converge to a different optimum.

```{r compare-losses, fig.width=7, fig.height=5}
model_mse <- nmf(movielens, k = best_k, L1 = c(0, 0.3), loss = "mse",
                 tol = 1e-5, maxit = 150, seed = 42, verbose = FALSE)
model_mae <- nmf(movielens, k = best_k, L1 = c(0, 0.3), loss = "mae",
                 tol = 1e-5, maxit = 150, seed = 42, verbose = FALSE)

compare_nmf(MSE = model_mse, MAE = model_mae) +
  ggtitle("Training Convergence: MSE vs MAE Loss")
```

Note that the convergence curves are not directly comparable since each model
optimizes a different objective. Let's evaluate both models on common metrics:

```{r evaluate-losses}
cat("Evaluation on all observed ratings:\n")
cat(sprintf("  MSE model — MSE: %.3f, MAE: %.3f\n",
            evaluate(model_mse, movielens, loss = "mse"),
            evaluate(model_mse, movielens, loss = "mae")))
cat(sprintf("  MAE model — MSE: %.3f, MAE: %.3f\n",
            evaluate(model_mae, movielens, loss = "mse"),
            evaluate(model_mae, movielens, loss = "mae")))
```

Each model optimizes its own loss best, as expected. In practice, the choice between
MSE and MAE depends on whether you care more about closely predicting typical ratings
(MSE) or about reducing the average absolute prediction error (MAE).

## Discovering Latent Taste Dimensions

The movie factors $W$ reveal what each latent dimension captures. By linking factors
to genre metadata, we can interpret what each dimension represents:

```{r genre-analysis, fig.width=7, fig.height=5}
# Compute average genre loading per factor
# genres is 19 x 3867 (genre x movies), W is 3867 x k (movies x factors)
genre_factor <- as.matrix(genres %*% model@w)  # 19 x k

# Normalize by number of movies in each genre
genre_counts <- rowSums(genres)
genre_factor_norm <- genre_factor / genre_counts

# Heatmap
genre_df <- expand.grid(
  Genre = rownames(genres),
  Factor = paste0("F", 1:ncol(model@w))
)
genre_df$Loading <- as.vector(genre_factor_norm)

ggplot(genre_df, aes(x = Factor, y = Genre, fill = Loading)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Genre Signatures of Latent Taste Dimensions",
       x = "Latent Factor", y = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Each column shows the "genre fingerprint" of a latent factor. A factor that loads
heavily on Action and Sci-Fi but not Romance likely captures affinity for
action-oriented films.

## Projecting New Users with NNLS

When a new user arrives and rates a few movies, we can compute their taste profile
without refitting the entire model. The `nnls()` function solves the non-negative
least squares problem: given the movie factors $W$, find the user loadings $h$ that
best reconstruct the new user's ratings.

```{r nnls-projection}
# Simulate a new user who rated 20 movies
set.seed(123)
rated_movies <- sample(nrow(movielens), 20)
actual_ratings <- sample(1:5, 20, replace = TRUE)

# Key: only pass the RATED movies to NNLS (not the full vector with zeros)
# Since the model was fit with sparse=TRUE, zeros mean "unobserved",
# so we must not include them in the projection.
W_scaled <- model@w * model@d[col(model@w)]  # m x k, scaled by d
W_rated <- W_scaled[rated_movies, , drop = FALSE]
A_rated <- matrix(actual_ratings, ncol = 1)

# Project onto existing model using only observed ratings
h_new <- nnls(w = W_rated, A = A_rated, nonneg = TRUE)

cat("User taste profile:\n")
barplot(h_new[, 1], names.arg = paste0("F", 1:nrow(h_new)),
        col = "steelblue", border = NA,
        main = "New User's Latent Taste Profile",
        ylab = "Factor Loading")
```

```{r predict-ratings}
# Predict this user's rating for all movies using the full W matrix
predicted <- as.vector(W_scaled %*% h_new)

# Show top recommended movies (highest predicted rating among unrated)
unrated <- setdiff(1:nrow(movielens), rated_movies)
top_recs <- unrated[order(predicted[unrated], decreasing = TRUE)[1:10]]
cat("\nTop 10 recommended movies:\n")
for (i in seq_along(top_recs)) {
  cat(sprintf("  %d. %s (predicted: %.1f stars)\n",
              i, rownames(movielens)[top_recs[i]], predicted[top_recs[i]]))
}
```

## Regularization Effects on Recommendations

Different regularization strategies produce qualitatively different recommendations.
Let's compare:

```{r regularization-compare, fig.width=7, fig.height=4}
# No regularization
m_noreg <- nmf(movielens, k = best_k, tol = 1e-5, maxit = 150,
               seed = 42, verbose = FALSE)

# L1 sparsity
m_l1 <- nmf(movielens, k = best_k, L1 = c(0.3, 0.3), tol = 1e-5,
            maxit = 150, seed = 42, verbose = FALSE)

# L2 ridge
m_l2 <- nmf(movielens, k = best_k, L2 = c(0.1, 0.1), tol = 1e-5,
            maxit = 150, seed = 42, verbose = FALSE)

compare_nmf(
  `No Regularization` = m_noreg,
  `L1 (Sparse)` = m_l1,
  `L2 (Ridge)` = m_l2
) + ggtitle("Effect of Regularization on Training Loss")
```

```{r reg-eval}
cat("Training reconstruction error (MSE):\n")
for (name in c("No reg", "L1", "L2")) {
  m <- list(m_noreg, m_l1, m_l2)[[match(name, c("No reg", "L1", "L2"))]]
  cat(sprintf("  %s: %.4f\n", name, evaluate(m, movielens, loss = "mse")))
}
```

Regularization increases training error (the model fits the observed data less
tightly) but can improve generalization to unseen ratings. The tradeoff depends
on data sparsity and user behavior:

- **No regularization** fits training data tightly but may overfit for sparse users.
- **L1** produces sharper taste profiles (fewer dominant factors per user) — better
  for users with narrow tastes.
- **L2** produces smoother predictions — better for users with diverse viewing habits.

## Key Takeaways

1. **Sparse cross-validation** (`mask_zeros = TRUE`) is essential for recommendation:
   it evaluates the model on the task it actually performs — predicting observed
   ratings, not reconstructing zeros.

2. **Rank selection via CV** prevents overfitting: too few factors underfit (can't
   capture genre nuances), too many overfit (memorize noise in ratings).

3. **MSE and MAE loss** optimize different objectives: MSE minimizes squared
   prediction error (sensitive to outliers), MAE minimizes absolute error
   (more uniform weighting). Choose based on your evaluation criteria.

4. **NNLS projection** enables real-time recommendation: new users can be profiled
   in milliseconds without refitting the entire model.

5. **Genre analysis** of latent factors confirms that NMF discovers interpretable
   taste dimensions that align with meaningful movie attributes.

6. **Regularization** controls the complexity–accuracy tradeoff: L1 for focused
   recommendations (specialist users), L2 for balanced recommendations (generalists).

## Further Reading

- [Getting started with RcppML](getting-started.html) — Core NMF concepts
- [Cross-validation for rank selection](cross-validation.html) — Detailed CV guide
- [Ecological community analysis](ecological-biodiversity.html) — NMF for ecological data
- [Image decomposition](image-decomposition.html) — Parts-based visual representations
