---
title: "Building a Movie Recommendation Engine with NMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Movie Recommendation Engine with NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## Motivation

Recommendation systems power some of the most impactful applications of machine
learning — from Netflix suggestions to Spotify playlists. The core idea behind
**collaborative filtering** is simple: users who agreed in the past will agree in
the future. Given a sparse matrix of user--item ratings, NMF discovers latent
preference dimensions that explain the observed ratings and predict unobserved
ones.

A critical subtlety: zeros in a ratings matrix mean "not rated," not "rated zero."
Standard NMF treats zeros as targets to reconstruct, which wastes model capacity
on predicting absence. RcppML's `mask_zeros` parameter solves this by restricting
the loss to observed entries only.

## API Reference

### Core Recommendation Pattern

The primary interface is `nmf()` with `mask_zeros = TRUE`:

```r
model <- nmf(data, k, mask_zeros = TRUE, seed = 42)
```

**Key parameters for recommendation:**

| Parameter | Type | Default | Description |
|:----------|:-----|:--------|:------------|
| `k` | integer or vector | — | Rank (number of latent factors). Pass a vector for cross-validation. |
| `mask_zeros` | logical | `FALSE` | If `TRUE`, zero entries do not contribute to the loss. |
| `test_fraction` | numeric | `0` | Fraction of *non-zero* entries held out for cross-validation. |
| `L1` | numeric(2) | `c(0, 0)` | L1 penalty on (W, H). Promotes sparse factor profiles. |
| `L2` | numeric(2) | `c(0, 0)` | L2 penalty on (W, H). Prevents large factor values. |
| `loss` | string | `"mse"` | Loss function: `"mse"`, `"mae"`, `"huber"`, or distribution-based. |
| `seed` | integer | `NULL` | Random seed for reproducible initialization. |

### Cross-Validation

Pass a vector of ranks to `k` for automatic cross-validation:

```r
cv <- nmf(data, k = 2:20, test_fraction = 0.2, mask_zeros = TRUE, seed = 42)
plot(cv)  # visualize test error vs. rank
```

### Reconstruction and Evaluation

```r
# Reconstruct the full predicted rating matrix
Ahat <- model@w %*% diag(model@d) %*% model@h

# Evaluate model loss on data
evaluate(model, data, loss = "mse")
```

### The MovieLens Dataset

`movielens` is a 3,867 × 610 sparse ratings matrix (movies × users) with ratings
from 1 to 5 stars. Genre metadata is available via `attr(movielens, "genres")`,
a 19 × 3,867 binary matrix mapping each movie to its genres.

## Theory

### Matrix Completion via NMF

NMF decomposes the ratings matrix as $A \approx W \cdot \text{diag}(d) \cdot H$,
where $W$ (movies × k) captures latent movie attributes, $H$ (k × users) captures
user preferences, and $d$ scales each dimension's importance.

The predicted rating for user $j$ on movie $i$ is:

$$\hat{A}_{ij} = \sum_{\ell=1}^{k} W_{i\ell} \, d_\ell \, H_{\ell j}$$

### Why `mask_zeros` Matters

With `mask_zeros = FALSE`, the loss function sums over *all* entries:

$$\mathcal{L} = \sum_{i,j} (A_{ij} - \hat{A}_{ij})^2$$

Since over 95% of entries are zero (unrated), the model spends most effort predicting
zeros rather than learning actual preferences. With `mask_zeros = TRUE`:

$$\mathcal{L} = \sum_{(i,j) \in \Omega} (A_{ij} - \hat{A}_{ij})^2$$

where $\Omega$ is the set of observed (non-zero) ratings. This focuses the model on
the signal that matters for recommendation.

## Worked Examples

### Example 1: Discovering Latent Taste Dimensions

We fit NMF to the MovieLens data and interpret the latent factors through their
genre composition.

```{r fit-model}
data(movielens)
genres <- attr(movielens, "genres")

model <- nmf(movielens, k = 10, mask_zeros = TRUE,
             L1 = c(0, 0.2), tol = 1e-5, maxit = 100, seed = 42,
             verbose = FALSE)
```

Each column of $W$ represents a latent "taste dimension." By examining which genres
load most heavily on each factor, we can interpret what preference axis it captures.

```{r genre-profile, fig.height = 5}
# Coerce pattern matrix to numeric for arithmetic
genres_num <- as(genres, "dgCMatrix") * 1.0

# Compute genre loadings per factor: genres (19 x 3867) times W (3867 x 10)
genre_factor <- as.matrix(genres_num %*% model@w)
genre_counts <- Matrix::rowSums(genres_num)
genre_factor_norm <- genre_factor / genre_counts

genre_df <- expand.grid(
  Genre = rownames(genres),
  Factor = paste0("F", seq_len(ncol(model@w)))
)
genre_df$Loading <- as.vector(genre_factor_norm)

ggplot(genre_df, aes(x = Factor, y = Genre, fill = Loading)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  labs(title = "Genre Signatures of Latent Taste Dimensions",
       x = "Latent Factor", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Each column in this heatmap reveals a coherent preference axis. Factors that load
heavily on Action and Sci-Fi but not Romance capture affinity for action-oriented
films. Factors dominated by Drama and Romance represent art-house or character-driven
taste profiles. The NMF discovers these genre clusters without any genre labels in
the input — only the rating patterns themselves.

```{r top-movies}
# Top 5 movies for first 4 factors
top_movies <- do.call(rbind, lapply(1:4, function(f) {
  top_idx <- order(model@w[, f], decreasing = TRUE)[1:5]
  movie_genres <- vapply(top_idx, function(idx) {
    col <- as.logical(as.vector(genres_num[, idx]))
    paste(rownames(genres)[col], collapse = ", ")
  }, character(1))
  data.frame(
    Factor = paste0("F", f),
    Movie = rownames(movielens)[top_idx],
    Loading = round(model@w[top_idx, f], 3),
    Genres = movie_genres,
    stringsAsFactors = FALSE
  )
}))

knitr::kable(top_movies, row.names = FALSE,
             caption = "Top 5 movies per latent factor (first 4 factors)")
```

### Example 2: Rank Selection via Cross-Validation

Choosing the right number of latent factors is critical. Too few factors underfit
(can't capture genre nuances); too many overfit (memorize noise in individual
ratings). RcppML's built-in cross-validation holds out a fraction of observed
ratings and measures prediction error on the held-out set.

```{r cv-rank, fig.width = 7, fig.height = 4}
cv <- nmf(movielens, k = c(3, 5, 8, 10, 15, 20), test_fraction = 0.2,
          mask_zeros = TRUE, tol = 1e-5, maxit = 100, seed = 42, verbose = FALSE)

knitr::kable(cv[, c("k", "train_mse", "test_mse")], digits = 3,
             col.names = c("Rank (k)", "Train MSE", "Test MSE"),
             caption = "Cross-validation: prediction error on held-out ratings")
```

Test MSE decreases as rank increases, showing that additional factors capture
real preference structure. The gap between train and test error indicates
generalization capacity — a widening gap signals overfitting.

### Example 3: Why `mask_zeros` Matters

The `mask_zeros` parameter fundamentally changes what the model optimizes.
We compare both modes at the same ranks to illustrate the difference.

```{r mask-compare}
cv_no_mask <- nmf(movielens, k = c(3, 5, 8, 10, 15, 20), test_fraction = 0.2,
                  mask_zeros = FALSE, tol = 1e-5, maxit = 100, seed = 42,
                  verbose = FALSE)

compare_df <- data.frame(
  k = cv$k,
  `Test MSE (mask_zeros=TRUE)` = round(cv$test_mse, 3),
  `Test MSE (mask_zeros=FALSE)` = round(cv_no_mask$test_mse, 4),
  check.names = FALSE
)

knitr::kable(compare_df, align = "r",
             caption = "Cross-validation test error: masking zeros vs. not")
```

The results reveal two fundamentally different optimization targets:

- **`mask_zeros = TRUE`**: Test MSE is on the scale of squared rating errors
  (~7–10), because the test set contains only actual ratings (values 1–5).
  Higher rank consistently improves predictions.
- **`mask_zeros = FALSE`**: Test MSE is much smaller (~0.3) because the test set
  is dominated by zero entries that are trivially predicted. But test error
  *increases* with rank beyond k = 5, revealing classic overfitting — the model
  spends its capacity fitting the rare non-zero entries at the expense of
  predicting the sea of zeros.

For recommendation, `mask_zeros = TRUE` is the correct choice: it evaluates the
model on the task it actually performs — predicting what users will rate, not
predicting which entries are missing.

## Key Takeaways

1. **`mask_zeros = TRUE`** is essential for recommendation: zeros mean "unobserved,"
   not "zero-rated." Masking them focuses the model on actual preferences.
2. **Rank selection** via cross-validation prevents overfitting. The test error
   curve reveals the optimal complexity for the data.
3. **Genre analysis** of latent factors confirms that NMF discovers interpretable
   preference dimensions from rating patterns alone — no genre labels needed.
4. **NNLS projection** (via `nnls()`) enables real-time recommendation for new users
   without refitting the model.

*See the [Cross-Validation](cross-validation.html) vignette for automated rank
selection with `test_fraction`, and the [Distributions](distributions.html)
vignette for non-Gaussian loss functions on ratings data.*
