---
title: "Regularization and Constraints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regularization and Constraints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## Why Regularize?

Unregularized NMF often produces dense factors where every feature loads on every component. The result is hard to interpret — factors overlap extensively and capture redundant structure. Regularization gives you control: **sparse** factors (few active features per component), **decorrelated** factors (minimal redundancy between components), or **smooth** factors (respecting spatial or network structure in the data).

RcppML provides six distinct penalty mechanisms, all applied independently to W and H.

## API Reference

### Sparsity Penalties

- `L1 = c(w_penalty, h_penalty)` — LASSO penalty, range [0, 1]. Drives small coefficients to exactly zero, producing element-wise sparsity.
- `L2 = c(w_penalty, h_penalty)` — Ridge penalty, range ≥ 0. Shrinks large coefficients toward zero without producing exact zeros.
- `L21 = c(w_penalty, h_penalty)` — Group sparsity, range ≥ 0. Zeros entire rows or columns of W or H, eliminating full features or samples from a factor.

### Structural Penalties

- `angular = c(w_penalty, h_penalty)` — Orthogonality penalty, range ≥ 0. Penalizes correlation between factor pairs, encouraging decorrelated components.
- `graph_W`, `graph_H` — Sparse graph Laplacian matrices (`dgCMatrix`). Penalizes differences between connected nodes, enforcing spatial or network smoothness.
- `graph_lambda = c(w_strength, h_strength)` — Controls the strength of graph regularization.

### Box Constraints

- `upper_bound = c(w_bound, h_bound)` — Maximum coefficient value. Useful for bounded data such as probabilities or normalized scores.
- `nonneg = c(TRUE, TRUE)` — Nonnegativity toggle. Set `FALSE` to allow negative coefficients (semi-NMF).

### Parameter Conventions

All penalty parameters accept either a single value (applied to both W and H) or a length-2 vector `c(w, h)` for independent control. Zero means no penalty (the default for all).

Each penalty modifies the NNLS subproblem: L1 subtracts from the RHS (soft thresholding), L2 adds to the Gram diagonal (Tikhonov), angular adds off-diagonal terms to the Gram (decorrelation), and graph Laplacian adds the Laplacian matrix to the Gram (smoothness).

## Theory

NMF solves $A \approx W \cdot \text{diag}(d) \cdot H$ by alternating NNLS updates. Each penalty extends the per-column subproblem:
$$\min_{h \geq 0} \|Gh - b\|^2 + \lambda_1 \|h\|_1 + \lambda_2 \|h\|^2_2$$

The full penalized NMF objective is:
$$\mathcal{L} = \|A - WDH\|_F^2 + \lambda_{L1} \sum_{ij} |W_{ij}| + \lambda_{L2} \sum_{ij} W_{ij}^2 + \lambda_{\text{ang}} \sum_{i < j} \cos(w_i, w_j) + \lambda_G \, \text{tr}(H L H^\top)$$

Each penalty modifies the NNLS subproblem:

- **L1 (LASSO)**: $\lambda_1 \|h\|_1$ — soft-thresholds the RHS, driving small coefficients to exactly zero. Ideal for sparse factors.
- **L2 (Ridge)**: $\lambda_2 \|h\|_2^2$ — adds $\lambda_2 I$ to the Gram diagonal (Tikhonov regularization). Shrinks coefficients smoothly.
- **L21 (Group LASSO)**: $\lambda_{21} \sum_g \|h_g\|_2$ — zeros entire groups (rows/columns). Useful for feature selection.
- **Angular**: $\lambda_a \sum_{i<j} \frac{w_i^\top w_j}{\|w_i\|\|w_j\|}$ — adds off-diagonal terms to the Gram, penalizing cosine similarity between factor pairs.
- **Graph Laplacian**: $\lambda_G \, \text{tr}(H L H^\top)$ where $L = D - A$ — adds the Laplacian to the Gram, encouraging connected nodes to share similar values.

| Goal | Penalty | Typical Range |
|------|---------|---------------|
| Sparse features per factor | L1 on W | 0.01–0.5 |
| Sparse sample loadings | L1 on H | 0.01–0.3 |
| Remove irrelevant features | L21 on W | 0.01–0.1 |
| Decorrelated factors | angular on W | 0.1–1.0 |
| Spatially smooth factors | graph_H + graph_lambda | 0.1–1.0 |
| Bounded coefficients | upper_bound | domain-specific |

## Example 1: L1 Sparsity Spectrum on Swimmers

The swimmer benchmark generates 32×32 images of stick figures with 4 limbs in 4 positions each (true rank = 16). We fit NMF at four L1 levels to see how sparsity affects factor interpretability.

```{r swimmer-l1-spectrum}
sim <- simulateSwimmer(256, style = "gaussian", seed = 42, return_factors = TRUE)
A <- t(sim$A)  # pixels (1024) x images (256)

l1_values <- c(0, 0.1, 0.3, 0.5)
results <- lapply(l1_values, function(l1) {
  nmf(A, k = 16, L1 = c(l1, 0), seed = 42, maxit = 50, verbose = FALSE)
})

sparsity_table <- data.frame(
  L1 = l1_values,
  `W Sparsity (% zeros)` = sapply(results, function(m) round(100 * mean(m@w == 0), 1)),
  `Reconstruction MSE` = sapply(results, function(m) round(evaluate(m, A), 5)),
  check.names = FALSE
)
knitr::kable(sparsity_table, digits = c(2, 1, 5),
             caption = "L1 sparsity vs. reconstruction tradeoff on the swimmer benchmark")
```

At L1 = 0 (no penalty), factors are dense and hard to interpret. As L1 increases, more coefficients are driven to zero — each factor isolates a specific limb position. Beyond L1 = 0.5, factors become over-sparse and lose structure.

```{r swimmer-l1-figure, fig.width = 7, fig.height = 7}
# Show one representative factor at each L1 level
plot_data <- do.call(rbind, lapply(seq_along(l1_values), function(i) {
  model <- results[[i]]
  # Show 4 diverse factors
  factor_ids <- c(1, 5, 9, 13)
  do.call(rbind, lapply(factor_ids, function(fi) {
    if (fi > ncol(model@w)) return(NULL)
    vals <- model@w[, fi]
    grid <- expand.grid(x = 1:32, y = 1:32)
    grid$value <- vals
    grid$factor <- paste("Factor", fi)
    grid$L1 <- paste0("L1 = ", l1_values[i])
    grid
  }))
}))

ggplot(plot_data, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_gradient(low = "white", high = "black") +
  scale_y_reverse() +
  facet_grid(L1 ~ factor) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        axis.title = element_blank(), legend.position = "none",
        strip.text = element_text(size = 8)) +
  ggtitle("Swimmer factors at increasing L1 sparsity")
```

The progression is clear: L1 = 0 produces diffuse factors; L1 = 0.3 yields clean, localized limb positions; L1 = 0.5 begins to over-penalize.

## Example 2: Angular Decorrelation on Face Data

When factors are correlated, they capture overlapping structure. Angular regularization penalizes pairwise correlation between factors, producing more distinct components.

```{r angular-faces}
data(olivetti)
A_faces <- t(olivetti[1:100, ])  # 4096 pixels x 100 images

baseline <- nmf(A_faces, k = 10, seed = 42, maxit = 50)
decorr <- nmf(A_faces, k = 10, angular = c(0.2, 0), seed = 42, maxit = 50)

# Compute factor correlation matrices
cor_base <- cor(baseline@w)
cor_decorr <- cor(decorr@w)

# Mean absolute off-diagonal correlation
mean_cor_base <- mean(abs(cor_base[upper.tri(cor_base)]))
mean_cor_decorr <- mean(abs(cor_decorr[upper.tri(cor_decorr)]))

cor_table <- data.frame(
  Method = c("Baseline (no penalty)", "Angular (0.2)"),
  `Mean |correlation|` = c(mean_cor_base, mean_cor_decorr),
  `Max |correlation|` = c(max(abs(cor_base[upper.tri(cor_base)])),
                          max(abs(cor_decorr[upper.tri(cor_decorr)]))),
  `Reconstruction MSE` = c(evaluate(baseline, A_faces),
                           evaluate(decorr, A_faces)),
  check.names = FALSE
)
knitr::kable(cor_table, digits = 4,
             caption = "Angular regularization reduces inter-factor correlation")
```

```{r angular-heatmap, fig.width = 7, fig.height = 3.5}
# Build correlation heatmap data for both models
make_cor_df <- function(cor_mat, label) {
  n <- nrow(cor_mat)
  grid <- expand.grid(Factor1 = 1:n, Factor2 = 1:n)
  grid$correlation <- as.vector(cor_mat)
  grid$method <- label
  grid
}

heatmap_data <- rbind(
  make_cor_df(cor_base, "Baseline"),
  make_cor_df(cor_decorr, "Angular (0.2)")
)

ggplot(heatmap_data, aes(x = Factor1, y = Factor2, fill = correlation)) +
  geom_raster() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(aspect.ratio = 1) +
  labs(fill = "Correlation", title = "Factor correlation: Baseline vs. Angular regularization")
```

Angular regularization produces factors that capture more distinct facial features, with substantially reduced pairwise correlation. The reconstruction cost is modest.

## Example 3: Graph Laplacian for Spatial Smoothness

When data has spatial or network structure, graph Laplacian regularization encourages connected nodes to share similar factor values. We build a k-nearest-neighbor graph from the geographic coordinates of Hawaiian bird survey grid cells and use it to produce spatially coherent community patterns.

```{r graph-laplacian}
data(hawaiibirds)
meta <- attr(hawaiibirds, "metadata_h")

# Build k-NN adjacency matrix from lat/lng coordinates
coords <- as.matrix(meta[, c("lat", "lng")])
n <- nrow(coords)
k_nn <- 10

# Compute pairwise distances and find k nearest neighbors
dist_mat <- as.matrix(dist(coords))
adj <- Matrix::sparseMatrix(i = integer(0), j = integer(0), x = numeric(0),
                            dims = c(n, n))
for (i in seq_len(n)) {
  neighbors <- order(dist_mat[i, ])[2:(k_nn + 1)]  # exclude self
  adj[i, neighbors] <- 1
  adj[neighbors, i] <- 1  # symmetric
}

# Graph Laplacian: L = D - A
degree <- Matrix::rowSums(adj)
laplacian <- Matrix::Diagonal(x = degree) - adj
laplacian <- as(laplacian, "dgCMatrix")

# Fit with and without graph regularization
baseline_g <- nmf(hawaiibirds, k = 6, seed = 42, maxit = 50)
smooth_g <- nmf(hawaiibirds, k = 6, graph_H = laplacian,
                graph_lambda = c(0, 0.5), seed = 42, maxit = 50)

# Measure smoothness: mean squared difference between connected H values
measure_smoothness <- function(model, adj_mat) {
  H <- model@h
  ij <- Matrix::which(adj_mat != 0, arr.ind = TRUE)
  ij <- ij[ij[, 1] < ij[, 2], , drop = FALSE]  # upper triangle only
  total <- 0
  for (f in seq_len(nrow(H))) {
    diffs <- H[f, ij[, 1]] - H[f, ij[, 2]]
    total <- total + mean(diffs^2)
  }
  total / nrow(H)
}

smooth_table <- data.frame(
  Method = c("Baseline (no graph)", "Graph Laplacian (λ = 0.5)"),
  `Reconstruction MSE` = c(evaluate(baseline_g, hawaiibirds),
                           evaluate(smooth_g, hawaiibirds)),
  `H Smoothness (mean Δ²)` = c(measure_smoothness(baseline_g, adj),
                                measure_smoothness(smooth_g, adj)),
  check.names = FALSE
)
knitr::kable(smooth_table, digits = 5,
             caption = "Graph-regularized NMF produces spatially smoother factors")
```

```{r graph-spatial-plot, fig.width = 7, fig.height = 5}
# Spatial visualization: dominant factor at each grid cell
assign_base <- apply(baseline_g@h, 2, which.max)
assign_smooth <- apply(smooth_g@h, 2, which.max)

spatial_df <- rbind(
  data.frame(lng = meta$lng, lat = meta$lat, Factor = factor(assign_base),
             Method = "Baseline (no graph)"),
  data.frame(lng = meta$lng, lat = meta$lat, Factor = factor(assign_smooth),
             Method = "Graph Laplacian (λ = 0.5)")
)

ggplot(spatial_df, aes(x = lng, y = lat, color = Factor)) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~ Method) +
  coord_fixed() +
  labs(title = "Dominant Factor by Grid Cell: Baseline vs. Graph-Regularized",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

Graph-regularized NMF produces spatially coherent community patterns — nearby grid cells load similarly on each factor, reflecting the geographic structure of bird communities. The spatial plot shows that graph regularization produces smoother factor assignments across the Hawaiian landscape, with the smoothness metric confirming the quantitative improvement.

## What's Next

- *See the [NMF Fundamentals](nmf-fundamentals.html) vignette for the core NMF API.*
- *See the [Image Decomposition](image-decomposition.html) vignette for L1 regularization applied to visual parts learning.*
- *See the [Factor Graphs](factor-graphs.html) vignette for regularization within composable graph architectures.*
