---
title: "Discovering Cell Types from Single-Cell RNA-Seq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discovering Cell Types from Single-Cell RNA-Seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
HAS_PBMC3K <- "pbmc3k" %in% data(package = "RcppML")$results[, "Item"]
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  eval = NOT_CRAN && HAS_PBMC3K
)
```

## Why NMF for Single-Cell Genomics?

Single-cell RNA sequencing (scRNA-seq) measures the expression of thousands of genes
in individual cells, producing a high-dimensional matrix where rows are genes and
columns are cells. A central task is to identify **cell types** — groups of cells that
share a common transcriptional program.

Non-negative matrix factorization (NMF) is particularly well-suited for this task:

- **Non-negativity** matches the biology: gene expression counts are non-negative, and
  NMF factors represent additive gene programs.
- **Parts-based decomposition**: each factor captures a distinct transcriptional program
  (metagene), and each cell is described as a mixture of these programs.
- **No pre-selection of genes**: NMF operates on the full expression matrix, letting the
  data determine which genes are informative.

In contrast to clustering algorithms (like k-means or Louvain), NMF simultaneously
identifies both the cell groupings (in $H$) and the gene signatures that define them
(in $W$). The factorization $A \approx W \cdot \text{diag}(d) \cdot H$ yields:

- $W$ (genes × k): **metagene signatures** — which genes define each program
- $d$ (length k): **factor weights** — relative importance of each program
- $H$ (k × cells): **cell loadings** — how strongly each cell expresses each program

## The PBMC 3K Dataset

The `pbmc3k` dataset is a classic benchmark from 10x Genomics containing 2,700
peripheral blood mononuclear cells (PBMCs) from a healthy donor. Expert-curated
cell type labels are stored as metadata:

```{r load-data, purl = FALSE}
library(RcppML)
library(ggplot2)
library(Matrix)

data(pbmc3k)
cat("Raw dimensions:", nrow(pbmc3k), "genes ×", ncol(pbmc3k), "cells\n")
cat("Sparsity:", round(1 - nnzero(pbmc3k) / prod(dim(pbmc3k)), 3), "\n")

cell_types <- attr(pbmc3k, "cell_type")
table(cell_types)
```

The data are highly sparse — most genes are not expressed in most cells, consistent
with the "dropout" phenomenon in scRNA-seq.

### Preprocessing

Before running NMF, we apply standard scRNA-seq preprocessing: library-size
normalization (so each cell contributes equally regardless of sequencing depth)
followed by log-transformation (to stabilize variance and reduce the influence of
highly-expressed housekeeping genes):

```{r preprocess, purl = FALSE}
# Library-size normalize to median total counts, then log-transform
lib_sizes <- colSums(pbmc3k)
scale_factors <- median(lib_sizes) / lib_sizes
pbmc_norm <- pbmc3k %*% Diagonal(x = scale_factors)
pbmc_norm@x <- log1p(pbmc_norm@x)

cat("After normalization:\n")
cat("  Value range:", round(range(pbmc_norm@x), 2), "\n")
cat("  Sparsity:", round(1 - nnzero(pbmc_norm) / prod(dim(pbmc_norm)), 3), "\n")
```

This preprocessing is critical: without it, NMF factors tend to capture library-size
effects and housekeeping gene programs rather than cell-type-specific signatures.

## Choosing the Right Number of Factors

Before fitting NMF, we need to determine the number of factors $k$ — analogous to the
number of latent transcriptional programs. Cross-validation provides a principled,
data-driven approach: we hold out a random subset of matrix entries, fit NMF on the
remaining entries, and measure prediction error on the held-out set.

### Multi-Rank Cross-Validation

We sweep across a range of ranks and use multiple random seeds to assess variability:

```{r cross-validation, fig.width=7, fig.height=4, purl = FALSE}
cv <- nmf(pbmc_norm, k = 2:12, test_fraction = 0.2, cv_seed = 1:3,
          tol = 1e-5, maxit = 100, verbose = FALSE)
plot(cv)
```

For highly sparse scRNA-seq count data, the cross-validation landscape is often
**noisy** — different random seeds and test splits produce variable results. Rather
than seeking a sharp "elbow," we look for the region where test error generally
stabilizes and use domain knowledge to guide the final choice.

### Automatic Rank Selection

RcppML can also search for the optimal rank automatically using binary search:

```{r auto-rank, purl = FALSE}
cv_auto <- nmf(pbmc_norm, k = "auto", cv_k_range = c(2, 15), test_fraction = 0.2,
               cv_seed = 42, tol = 1e-5, maxit = 100, verbose = FALSE)
cat("Automatically selected rank:", cv_auto$rank_search$k_optimal, "\n")
```

For this dataset, the automatic search may select a high rank (even the upper bound),
indicating that test error continues to decrease as more factors are added. This is
common with sparse count data where the effective dimensionality is high.

## Fitting the NMF Model

Based on cross-validation and the known biology (7 annotated cell types in this
dataset), we use $k = 7$. We apply L1 regularization on $H$ to encourage sparse
cell assignments (each cell loads primarily on a few factors):

```{r fit-nmf, purl = FALSE}
model <- nmf(pbmc_norm, k = 7, L1 = c(0, 0.5), loss = "mse",
             tol = 1e-5, maxit = 200, seed = 42, verbose = FALSE)
model
```

### Training Diagnostics

The training history shows how the loss decreased over iterations:

```{r plot-loss, fig.width=7, fig.height=4, purl = FALSE}
plot(model, type = "loss")
```

### Factor Scaling

The diagonal scaling vector $d$ tells us the relative importance of each factor:

```{r diag-scaling, purl = FALSE}
model_sorted <- sort(model)
barplot(model_sorted@d, names.arg = paste0("F", seq_along(model_sorted@d)),
        main = "Factor Importance (diagonal scaling, sorted)",
        ylab = "d", col = "steelblue", border = NA)
```

Note: after `sort()`, factors are reordered by decreasing importance. The labels
F1, F2, ... refer to the sorted order, not the original factor indices.

## Interpreting Factors as Cell Type Programs

### Cell Type Composition by Factor

The `summary()` method computes how each factor maps to known cell type labels,
producing a factor-by-group representation table:

```{r summary-groups, fig.width=7, fig.height=5, purl = FALSE}
nmf_summary <- summary(model, group_by = cell_types)
plot(nmf_summary) +
  ggtitle("Factor Representation Across Cell Types") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

If NMF is working well, each factor should be dominated by one or two cell types,
demonstrating that NMF independently discovers biologically meaningful programs.

### Biplot: Visualizing Cell Relationships

A biplot projects cells into a 2D space defined by two factors, revealing how cell
populations separate:

```{r biplot, fig.width=7, fig.height=5, purl = FALSE}
biplot(model, factors = c(1, 2), matrix = "h", group_by = cell_types) +
  ggtitle("Cell Biplot: Factor 1 vs Factor 2")
```

## Factor Sparsity

L1 regularization encourages sparsity in $H$, meaning each cell loads on only a few
factors. We can quantify this:

```{r sparsity, purl = FALSE}
sp <- sparsity(model)
print(sp)
```

Higher sparsity in $H$ means cleaner cell-to-factor assignments. Higher sparsity in $W$
means each factor is defined by fewer genes, producing more interpretable signatures.

## Model Evaluation

We can quantify reconstruction quality using multiple metrics:

```{r evaluate, purl = FALSE}
mse_val <- evaluate(model, pbmc_norm, loss = "mse")
mae_val <- evaluate(model, pbmc_norm, loss = "mae")
cat("MSE:", round(mse_val, 4), "\n")
cat("MAE:", round(mae_val, 4), "\n")
cat("Per-element RMSE:", round(sqrt(mse_val), 4), "\n")
```

These are mean losses averaged over all matrix entries (including zeros). For context,
the mean non-zero value in the normalized matrix is:

```{r context, purl = FALSE}
cat("Mean non-zero value:", round(mean(pbmc_norm@x), 4), "\n")
cat("RMSE / mean non-zero:", round(sqrt(mse_val) / mean(pbmc_norm@x), 2), "\n")
```

## Projecting New Cells onto Existing Factors

A powerful application of NMF is projecting new data onto a learned model. Given a
matrix $W$ of gene programs, we can classify new cells by solving for their loadings:

```{r predict, purl = FALSE}
# Simulate "new cells" by holding out 100 cells
test_cells <- pbmc_norm[, 1:100]
train <- pbmc_norm[, 101:ncol(pbmc_norm)]

# Fit model on training data
model_train <- nmf(train, k = 7, L1 = c(0, 0.5), tol = 1e-5,
                   maxit = 200, seed = 42, verbose = FALSE)

# Project held-out cells
h_new <- predict(model_train, test_cells)
cat("Projected loadings dimensions:", nrow(h_new), "×", ncol(h_new), "\n")
```

Each column of `h_new` gives the factor loadings for a new cell, enabling
classification without re-running the full factorization.

## Divisive Clustering with dclust

RcppML also provides `dclust()`, which performs hierarchical clustering via recursive
rank-2 bipartitioning. This is useful as an alternative to graph-based clustering:

```{r dclust, purl = FALSE}
clust <- dclust(pbmc_norm, min_samples = 20, seed = 42)
if (length(clust$id) > 0) {
  cat("Number of clusters found:", length(clust$id), "\n")
  cat("Cluster sizes:", sapply(clust$samples, length), "\n")
} else {
  cat("dclust did not find stable subclusters with these parameters.\n")
  cat("This can happen when the bipartition splitting criterion\n")
  cat("is not met — the data may not have a strong hierarchical structure\n")
  cat("at this resolution.\n")
}
```

## Extracting Top Marker Genes

The $W$ matrix reveals which genes drive each factor. By examining the highest-loading
genes per factor, we can identify **marker genes** — genes that distinguish each cell
population:

```{r marker-genes, purl = FALSE}
# Get top 5 genes per factor
for (i in 1:7) {
  idx <- order(model@w[, i], decreasing = TRUE)[1:5]
  genes <- rownames(model@w)[idx]
  cat(sprintf("Factor %d: %s\n", i, paste(genes, collapse = ", ")))
}
```

Some factors yield clear cell-type markers (e.g., LYZ/S100A9 for monocytes,
CD74/HLA-DRA for B cells, NKG7 for NK cells). Others may be dominated by highly-
expressed “housekeeping” genes (MALAT1, RPL/RPS ribosomal proteins) that are
ubiquitously expressed. This is a known limitation — additional gene filtering
(e.g., removing ribosomal/mitochondrial genes) can sharpen the biological
interpretability of NMF factors.

## Key Takeaways

1. **NMF naturally discovers cell types** by decomposing the gene expression matrix into
   additive transcriptional programs — without prior knowledge of cell labels.

2. **Cross-validation** provides a principled way to select the number of programs,
   avoiding overfitting while capturing biological complexity.

3. **L1 regularization** produces sparse cell loadings, making factor-to-cell-type
   assignments cleaner and more interpretable.

4. **Factor inspection** reveals marker genes — the `summary()` and biplot show how
   factors map to biological cell populations.

5. **Projection** via `predict()` enables classification of new cells against a
   reference model, critical for atlas-based cell type annotation.

6. **Preprocessing matters**: Library-size normalization and log-transformation
   are essential before NMF on scRNA-seq data. Without them, factors capture
   technical variation (sequencing depth) rather than biological programs.

7. **Divisive clustering** via `dclust()` provides complementary hierarchical structure
   that can reveal rare cell populations missed by fixed-rank NMF.

## Further Reading

- [Getting started with RcppML](getting-started.html) — Core NMF concepts
- [Cross-validation for rank selection](cross-validation.html) — Detailed CV guide
- [Cancer subtype discovery](cancer-subtype-discovery.html) — Consensus NMF for robust clustering
- [Image decomposition](image-decomposition.html) — Parts-based visual representations
