---
title: "SparsePress: High-Performance Sparse Matrix I/O"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SparsePress: High-Performance Sparse Matrix I/O}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = NOT_CRAN
)
```

## Introduction

Large sparse matrices are the backbone of many computational workflows — single-cell genomics, recommendation systems, text mining, and more.
Yet standard formats like `.rds`, `.mtx`, and `.h5ad` are either slow to read, poorly compressed, or both. **SparsePress** (`.spz`) is a column-oriented compressed format built directly into RcppML that solves this with:
 
- **10–20× compression** on typical sparse data (e.g., scRNA-seq count matrices)
- **Fast decompression**, often faster than reading an uncompressed `.rds` file
- **Self-describing header**: inspect dimensions, density, and compression ratio without decompressing
- **Lossless fidelity**: exact round-trip for double-precision values (with optional lossy precision modes)
- **Pre-stored transpose**: eliminate runtime transpose costs for NMF and other iterative algorithms
- **Streaming NMF**: run out-of-core NMF directly from a `.spz` file, without loading the full matrix into RAM
- **GPU zero-copy path**: decode a `.spz` file directly into GPU device memory, bypassing the CPU-to-GPU transfer

This vignette covers the full SparsePress API and explains how `.spz` files accelerate both CPU and GPU compute pipelines.

```{r libraries}
library(RcppML)
library(Matrix)
```

## 1. The SparsePress Format

SparsePress stores CSC (Compressed Sparse Column) matrices using a combination of techniques tuned for the statistics of real-world sparse data:

| Technique | What it does |
|-----------|-------------|
| **rANS entropy coding** | Near-optimal compression of row-index gaps and values |
| **Delta prediction** | Exploits row-density structure to shrink residuals before encoding |
| **Value prediction** | For integer data, predicts values from column context |
| **Chunked v2 layout** | Enables partial column reads and GPU-friendly decode |
| **Row permutation** | Sorts rows by density for better compression (optional) |
| **Precision modes** | `fp64` (lossless), `fp32`, `fp16`, `quant8` for size/accuracy trade-offs |

The format preserves `dgCMatrix` structure (column pointers, row indices, values) and round-trips dimension names.

## 2. Writing: `sp_write()`

`sp_write()` compresses a sparse matrix to a `.spz` file:

```{r write-basic}
# Simulate a sparse matrix (5000 genes × 1000 cells, ~5% density)
set.seed(42)
A <- rsparsematrix(5000, 1000, density = 0.05)

# Write to .spz
f <- tempfile(fileext = ".spz")
stats <- sp_write(A, f, verbose = TRUE)
```

The return value contains compression statistics:

```{r write-stats}
cat(sprintf("  Raw size:        %.2f MB\n", stats$raw_bytes / 1e6))
cat(sprintf("  Compressed size: %.2f MB\n", stats$compressed_bytes / 1e6))
cat(sprintf("  Ratio:           %.1fx\n", stats$ratio))
cat(sprintf("  Time:            %.1f ms\n", stats$compress_ms))
```

### Key Parameters

| Parameter | Default | Effect |
|-----------|---------|--------|
| `delta` | `TRUE` | Density-based delta prediction; improves compression for structured sparsity |
| `value_pred` | `FALSE` | Value prediction; effective only for non-negative integer data |
| `precision` | `"auto"` | Value precision: `"fp64"`, `"fp32"`, `"fp16"`, `"quant8"` |
| `row_sort` | `FALSE` | Sort rows by nnz for better compression |
| `include_transpose` | `FALSE` | Also store CSC(A^T) — roughly doubles file size |

### Including the Pre-Stored Transpose

Many iterative algorithms (NMF, ALS) need both $A$ and $A^\top$ in CSC format. Computing the transpose at runtime is expensive for large matrices. With `include_transpose = TRUE`, both orientations are stored in the `.spz` file:

```{r write-transpose}
f_with_t <- tempfile(fileext = ".spz")
sp_write(A, f_with_t, include_transpose = TRUE, verbose = TRUE)
```

This approximately doubles the file size but eliminates the runtime transpose entirely — a significant win for iterative algorithms.

### Precision Modes

For count data, half-precision or 8-bit quantization can dramatically reduce file size with negligible impact on downstream analysis:

```{r precision-modes}
# Integer count matrix (common in scRNA-seq)
counts <- abs(A)
counts@x <- round(counts@x * 10)  # simulate integer counts

f_fp32 <- tempfile(fileext = ".spz")
f_fp16 <- tempfile(fileext = ".spz")
f_q8   <- tempfile(fileext = ".spz")

s_auto <- sp_write(counts, tempfile(fileext = ".spz"))
s_fp32 <- sp_write(counts, f_fp32, precision = "fp32")
s_fp16 <- sp_write(counts, f_fp16, precision = "fp16")
s_q8   <- sp_write(counts, f_q8,   precision = "quant8")

cat(sprintf("  auto  : %.2f MB (%.1fx)\n", s_auto$compressed_bytes/1e6, s_auto$ratio))
cat(sprintf("  fp32  : %.2f MB (%.1fx)\n", s_fp32$compressed_bytes/1e6, s_fp32$ratio))
cat(sprintf("  fp16  : %.2f MB (%.1fx)\n", s_fp16$compressed_bytes/1e6, s_fp16$ratio))
cat(sprintf("  quant8: %.2f MB (%.1fx)\n", s_q8$compressed_bytes/1e6,   s_q8$ratio))
```

## 3. Reading: `sp_read()`

`sp_read()` decompresses a `.spz` file back into a `dgCMatrix`:

```{r read-basic}
B <- sp_read(f)
cat(sprintf("Round-trip match: %s\n", all.equal(A, B)))
dim(B)
```

### Partial Column Reads

For v2 files, you can read a subset of columns without decompressing the entire file:

```{r partial-read}
# Read only columns 1 through 100
B_partial <- sp_read(f, cols = 1:100)
dim(B_partial)
```

### Reading the Pre-Stored Transpose

If the file was written with `include_transpose = TRUE`, read the transpose directly:

```{r read-transpose}
At <- sp_read_transpose(f_with_t)
dim(At)  # transposed dimensions
```

This is equivalent to `t(sp_read(f_with_t))` but avoids the transpose computation.

## 4. Inspecting: `sp_info()`

`sp_info()` reads only the 72-byte file header — no decompression occurs. This makes it instant even for multi-gigabyte files:

```{r inspect}
info <- sp_info(f)
cat(sprintf("Matrix:      %d × %d\n", info$rows, info$cols))
cat(sprintf("Non-zeros:   %s\n", format(info$nnz, big.mark = ",")))
cat(sprintf("Density:     %.2f%%\n", info$density_pct))
cat(sprintf("Compression: %.1fx\n", info$ratio))
cat(sprintf("Format:      v%d\n", info$version))
```

This is particularly useful for cataloging large collections of `.spz` files or building data inventories without loading any data into memory.

## 5. In-Memory Compression: `sp_compress()` / `sp_decompress()`

For network transfer, caching, or database storage, you can compress a matrix to a raw byte vector without touching disk:

```{r in-memory}
blob <- sp_compress(A)
cat(sprintf("Compressed to %s (raw vector of length %d)\n",
            format(object.size(blob), units = "MB"), length(blob)))

A_restored <- sp_decompress(blob)
cat(sprintf("Round-trip match: %s\n", all.equal(A, A_restored)))
```

This is useful for:

- Sending matrices over network connections (e.g., Shiny apps, APIs)
- Storing compressed matrices in databases
- In-memory caching with reduced footprint

## 6. Format Conversion: `sp_convert()`

`sp_convert()` is a one-stop migration tool that reads from any supported format and writes a `.spz` file:

```{r convert, eval = FALSE}
# Convert from various formats
sp_convert("counts.h5ad", "counts.spz")            # AnnData
sp_convert("matrix.mtx",  "matrix.spz")            # MatrixMarket
sp_convert("data.rds",    "data.spz")              # R serialized
sp_convert("table.csv",   "table.spz")             # CSV

# Convert with options
sp_convert("big_matrix.h5ad", "big_matrix.spz",
           precision = "fp16",              # lossy half-precision
           include_transpose = TRUE,        # pre-store transpose
           row_sort = TRUE)                 # sort rows for better compression
```

For `.h5ad` files, `sp_convert()` automatically extracts AnnData metadata (obs, var, uns) and saves it as a JSON sidecar file (`<output>.meta`) so cell/gene annotations are preserved alongside the compressed matrix.

## 7. Comparison with Other Formats

To understand where SparsePress fits, consider how it compares with common alternatives:

| Format | Compression | Read Speed | Partial Reads | Transpose | Metadata |
|--------|:-----------:|:----------:|:-------------:|:---------:|:--------:|
| `.rds` (R binary) | ~2–4× | Moderate | No | No | No |
| `.mtx` (MatrixMarket) | None | Slow | No | No | No |
| `.h5ad` (AnnData) | ~2–5× | Moderate | Yes | No | Yes |
| `.h5` (HDF5) | ~2–5× | Moderate | Yes | No | Flexible |
| **`.spz` (SparsePress)** | **10–20×** | **Fast** | **Yes** | **Yes** | **Sidecar** |

The key advantages of `.spz` are compression ratio and decompression speed. The rANS entropy coding combined with density-aware delta prediction is specifically tuned for the statistics of sparse matrices, which gives it a significant edge over general-purpose compressors.

## 8. Accelerating CPU Compute

SparsePress is not just a storage format — it integrates deeply with RcppML's compute pipeline to improve CPU performance in several ways.

### 8.1 Faster Data Loading

For large matrices, I/O is often the bottleneck. A 10× smaller file means 10× less data to read from disk or transfer over a network. Combined with fast decompression, `.spz` typically loads faster than uncompressed formats:

```{r cpu-loading, eval = FALSE}
# Traditional workflow: slow read from .rds
system.time(A <- readRDS("big_matrix.rds"))        # e.g., 45 seconds

# SparsePress: faster even though it includes decompression
system.time(A <- sp_read("big_matrix.spz"))         # e.g., 8 seconds
```

### 8.2 Eliminating the Transpose Cost

NMF alternates between updating $W$ (which requires $A$ in CSC) and updating $H$ (which requires $A^\top$ in CSC). For a matrix with $m$ rows, $n$ columns, and $\text{nnz}$ non-zeros, computing the transpose costs $O(\text{nnz})$ time and doubles peak memory. With `include_transpose = TRUE`, both orientations are pre-stored:

```{r transpose-benefit, eval = FALSE}
# Without pre-stored transpose: transpose computed at runtime
A <- sp_read("data.spz")
system.time(At <- t(A))  # O(nnz) time + doubles memory

# With pre-stored transpose: instant
A  <- sp_read("data.spz")
At <- sp_read_transpose("data.spz")  # already stored, just decompress
```

For a 30,000 × 50,000 matrix with 100M non-zeros, this can save 5–10 seconds per NMF run.

### 8.3 Streaming (Out-of-Core) NMF

For matrices too large to fit in RAM, RcppML can run NMF directly from a `.spz` file using **streaming mode**. Instead of loading the entire matrix into memory, the algorithm reads columns in panels and processes them incrementally:

```{r streaming-nmf, eval = FALSE}
# Write a large matrix with transpose (required for streaming)
sp_write(huge_matrix, "huge.spz", include_transpose = TRUE)

# Enable streaming mode
options(RcppML.streaming = TRUE)

# NMF runs out-of-core — only one panel in memory at a time
result <- nmf("huge.spz", k = 20, maxit = 100, seed = 42)

options(RcppML.streaming = FALSE)
```

Streaming NMF works by:

1. Reading a panel of columns from the `.spz` file
2. Solving the NNLS subproblem for those columns
3. Discarding the panel and moving to the next
4. After processing all columns, reading panels from the pre-stored transpose for the $W$ update

This means peak memory is proportional to the panel size, not the full matrix — enabling NMF on datasets that are 10–100× larger than available RAM.

**Requirements**: The `.spz` file must be written with `include_transpose = TRUE` (v2 format). Streaming NMF supports all loss functions including GP, NB, and other non-MSE losses via full-weight IRLS.

**Cross-validation** also works in streaming mode — pass multiple values of `k` to automatically sweep ranks:

```{r streaming-cv, eval = FALSE}
options(RcppML.streaming = TRUE)
cv_results <- nmf("huge.spz", k = 2:20, test_fraction = 0.1, seed = 42)
plot(cv_results)
options(RcppML.streaming = FALSE)
```

## 9. Accelerating GPU Compute

SparsePress provides a direct path from compressed file to GPU device memory, bypassing the usual bottleneck of loading to CPU first and then transferring to the GPU.

### 9.1 The GPU Data Transfer Problem

The traditional GPU NMF workflow looks like this:

```
Disk → [decompress] → CPU RAM → [copy to GPU] → GPU memory → NMF
```

For a large matrix, the CPU-to-GPU transfer over PCIe is the bottleneck:

- A 20,000 × 50,000 matrix with 50M non-zeros requires ~600 MB in CSC format
- PCIe 3.0 ×16 bandwidth: ~12 GB/s → ~50 ms transfer time
- PCIe 4.0 ×16 bandwidth: ~25 GB/s → ~24 ms transfer time

But the real cost is holding the full uncompressed matrix in CPU RAM *in addition to* the GPU copy. For very large matrices, this can exceed available memory.

### 9.2 Zero-Copy GPU Decode: `sp_read_gpu()`

`sp_read_gpu()` reads a `.spz` file, decodes it, and places the resulting CSC arrays directly in GPU device memory. The matrix never exists in full in CPU RAM:

```{r gpu-workflow, eval = FALSE}
# Check GPU availability
if (gpu_available()) {
  print(gpu_info())
  
  # Read .spz directly to GPU memory
  gpu_mat <- sp_read_gpu("data.spz")
  print(gpu_mat)
  
  # Run NMF on GPU-resident data (zero-copy)
  result <- nmf(gpu_mat, k = 10, maxit = 100, verbose = TRUE)
  
  # Clean up GPU memory (optional — GC handles this)
  sp_free_gpu(gpu_mat)
}
```

The `gpu_sparse_matrix` object is an opaque handle to device memory with standard R methods:

```{r gpu-methods, eval = FALSE}
dim(gpu_mat)      # c(rows, cols)
nrow(gpu_mat)     # number of rows
ncol(gpu_mat)     # number of columns
print(gpu_mat)    # summary with pointer info
```

### 9.3 GPU Memory Management

GPU memory allocated by `sp_read_gpu()` is automatically freed when the R object is garbage-collected. For explicit control:

```{r gpu-memory, eval = FALSE}
# Explicit free
sp_free_gpu(gpu_mat)

# Verify — pointers are zeroed after free
gpu_mat$.col_ptr  # 0
```

This is important when working with multiple large matrices on GPU, where device memory is limited.

### 9.4 When GPU + SparsePress Helps Most

The GPU pipeline benefits from `.spz` most in these scenarios:

| Scenario | Benefit |
|----------|---------|
| Matrix close to GPU memory limit | `.spz` avoids the CPU copy, freeing RAM for other uses |
| Repeated NMF runs (e.g., CV sweep) | Load once to GPU, run many factorizations |
| Network/cluster storage | 10–20× smaller files = faster transfer to compute nodes |
| Repeated NMF runs on GPU | Direct GPU loading avoids repeated CPU→GPU transfers |

### 9.5 Requirements

- `.spz` v2 format (written by `sp_write()` or `sp_convert()`)
- NVIDIA GPU with compute capability ≥ 7.0
- CUDA Toolkit ≥ 12.x
- RcppML GPU library built and available (see `vignette("gpu-acceleration")`)

## 10. Complete Workflow Example

Here is a complete example showing the SparsePress workflow from data creation through NMF:

```{r full-example}
library(RcppML)
library(Matrix)

# 1. Create or load a sparse matrix
data(pbmc3k)
A <- pbmc3k
cat(sprintf("Matrix: %d genes × %d cells, %s non-zeros (%.1f%% dense)\n",
            nrow(A), ncol(A), format(length(A@x), big.mark = ","),
            length(A@x) / (as.double(nrow(A)) * ncol(A)) * 100))

# 2. Write to .spz with pre-stored transpose
spz_path <- tempfile(fileext = ".spz")
stats <- sp_write(A, spz_path, include_transpose = TRUE)
cat(sprintf("Compressed: %.1fx (%.2f MB → %.2f MB)\n",
            stats$ratio, stats$raw_bytes/1e6, stats$compressed_bytes/1e6))

# 3. Inspect without loading
info <- sp_info(spz_path)
cat(sprintf("Header: %d × %d, nnz=%s, v%d\n",
            info$rows, info$cols,
            format(info$nnz, big.mark = ","),
            info$version))

# 4. Read back
B <- sp_read(spz_path)
cat(sprintf("Lossless: %s\n", all.equal(A, B)))

# 5. Read pre-stored transpose
At <- sp_read_transpose(spz_path)
cat(sprintf("Transpose: %d × %d\n", nrow(At), ncol(At)))

# 6. Run NMF (using the original matrix)
model <- nmf(A, k = 8, seed = 42, verbose = FALSE)
cat(sprintf("NMF: k=%d, %d iterations, loss=%.4e\n",
            ncol(model@w), model@misc$iter, model@misc$tol))

# 7. Clean up
unlink(spz_path)
```

## 11. API Reference Summary

| Function | Description |
|----------|-------------|
| `sp_write(x, path, ...)` | Write a `dgCMatrix` to a `.spz` file |
| `sp_read(path, cols, reorder)` | Read a `.spz` file into a `dgCMatrix` |
| `sp_info(path)` | Read the 72-byte header (no decompression) |
| `sp_read_gpu(path, device)` | Decode `.spz` directly into GPU device memory |
| `sp_free_gpu(x)` | Free GPU-resident sparse matrix |
| `sp_read_transpose(path)` | *(internal)* Read the pre-stored CSC(A^T) from a v2 file |
| `sp_compress(x)` | *(internal)* Compress to an in-memory raw vector |
| `sp_decompress(x)` | *(internal)* Decompress a raw vector to a `dgCMatrix` |
| `sp_convert(input, output, ...)` | *(internal)* Convert any supported format to `.spz` |

## 12. Best Practices

1. **Always use `include_transpose = TRUE`** if you plan to run NMF or other iterative algorithms. The ~2× file size increase is well worth the eliminated transpose cost.

2. **Use `row_sort = TRUE`** for archival storage. Row sorting improves compression by grouping similar-density rows, but the original row order is restored on read.

3. **Choose precision wisely**:
   - `"auto"` (default): lossless for floats, auto-detects integers
   - `"fp32"`: ~50% smaller values, suitable when double precision is not needed
   - `"fp16"`: aggressive compression for count data or when 3–4 significant digits suffice
   - `"quant8"`: maximum compression, suitable only for rough approximations

4. **Use `sp_info()` before loading** to check dimensions and estimate memory requirements.

5. **Enable streaming for large data**: Set `options(RcppML.streaming = TRUE)` when working with `.spz` files larger than available RAM.

6. **For GPU workflows**, prefer `sp_read_gpu()` over `sp_read()` + GPU transfer to minimize peak memory usage and eliminate the transfer step.

```{r cleanup, include = FALSE}
# Clean up any temp files
unlink(c(f, f_with_t, f_fp32, f_fp16, f_q8))
```
