---
title: "StreamPress: High-Performance Sparse Matrix I/O"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{StreamPress: High-Performance Sparse Matrix I/O}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
library(RcppML)
library(ggplot2)
```

## Motivation

R's built-in serialization formats (`.rds`, `.rda`) are general-purpose but
inefficient for sparse matrices. They serialize the entire CSC structure including
all slot metadata, which inflates file sizes and slows I/O for large datasets.

StreamPress (`.spz`) is a columnar compressed format designed specifically for
sparse matrices. It uses rANS entropy coding with delta-encoded indices to achieve
5–10× compression over raw CSC binary on typical scRNA-seq count data, and larger
gains over `.rds` files.  Beyond storage savings, SPZ is also faster to read
than a raw float32 CSC binary — even single-threaded — because the dominant
cost in loading a large sparse matrix is constructing the R/Eigen sparse object
from raw index and value arrays, not transferring bytes off disk.  SPZ parallelises
that reconstruction step across chunks, while raw CSC must be assembled
sequentially.  For dense matrices, StreamPress v3 provides optional FP16, QUANT8,
and rANS compression codecs.

## API Reference

### Writing Sparse Matrices

`st_write()` compresses a sparse `dgCMatrix` to `.spz` format:

```r
stats <- st_write(x, path,
  precision = "auto",         # "fp64", "fp32", "fp16", "quant8", or "auto"
  include_transpose = TRUE,   # store CSC(A^T) for fast row access
  delta = TRUE,               # delta prediction for row indices
  value_pred = FALSE,         # value prediction for integer data
  chunk_cols = NULL,          # columns per chunk (NULL = auto from chunk_bytes)
  chunk_bytes = 8e6,          # target bytes per chunk (8 MB default)
  verbose = FALSE             # print compression statistics
)
```

The default `chunk_bytes = 8e6` sets chunk size to ~50 columns for a typical
38 k-row scRNA-seq matrix, yielding ~60 chunks per sample file.  This is the
sweet spot for parallel decompression: enough chunks for real multi-core
speedup, while each chunk is large enough that per-chunk overhead is negligible.
Use `chunk_bytes = 64e6` (or larger) only if you are writing very small matrices
where chunk overhead would dominate.

Returns an invisible list with compression statistics: `raw_bytes`,
`compressed_bytes`, `ratio`, `compress_ms`.

### Reading Sparse Matrices

`st_read()` decompresses a `.spz` file to a `dgCMatrix`:

```r
mat <- st_read(path,
  cols = NULL,       # column indices to read (NULL = all)
  reorder = TRUE,    # undo any row permutation
  threads = NULL     # NULL = auto (1 thread for <50 MB, all cores for ≥50 MB)
)
```

`threads = NULL` (the default) automatically selects the thread count based on
file size.  For individual sample files (typically < 10 MB compressed),
multi-threading adds overhead without measurable benefit — decompression
throughput is ~21 MB/s per thread and thread fork/join costs dominate.  For
large concatenated files (≥ 50 MB), all available cores are used.

The `cols` parameter enables partial reads — load only the columns you need
without decompressing the entire file.

### Inspecting Files

`st_info()` reads only the header, with no decompression:

```r
info <- st_info(path)
# Returns: rows, cols, nnz, density_pct, file_bytes, raw_bytes, ratio,
#          version, value_type, chunk_cols, num_chunks, ...
```

### Dense Matrix Support

StreamPress v3 handles dense matrices with optional compression codecs:

```r
st_write_dense(x, path,
  codec = "raw",              # "raw", "fp16", "quant8", "fp16_rans", "fp32_rans"
  include_transpose = FALSE,  # store transposed panels
  chunk_cols = 2048L          # columns per chunk
)
mat <- st_read_dense(path)
```

## Theory

### Columnar Storage

StreamPress stores each column independently in a CSC-like layout. This design
enables:

- **Random column access**: Read column $j$ without decompressing columns 1 through
  $j-1$.
- **Streaming I/O**: Process columns in chunks for out-of-core NMF on datasets
  larger than RAM.
- **Parallel decompression**: Independent columns can be decoded concurrently.

### Compression Pipeline

For each column, StreamPress applies:

1. **Delta encoding** of row indices: Store differences between consecutive indices
   rather than absolute values. For dense rows, deltas are mostly 1 — tiny integers
   that compress extremely well.
2. **rANS entropy coding**: An asymmetric numeral system that approaches the
   Shannon entropy bound, yielding near-optimal compression.
3. **Precision reduction** (optional): Convert 64-bit doubles to 32-bit floats,
   16-bit half-precision, or 8-bit quantized values before entropy coding.

### Precision Hierarchy

| Precision | Bytes/value | Exact for integers up to | Typical use case |
|:----------|:------------|:-------------------------|:-----------------|
| `fp64` | 8 | $2^{53}$ | Lossless (default) |
| `fp32` | 4 | $2^{24}$ (16.7 million) | Count data, most use cases |
| `fp16` | 2 | $2^{11}$ (2,048) | Small counts, ratings |
| `quant8` | 1 | 256 buckets | Maximum compression |

For integer count data (UMI counts, ratings), `fp32` or `fp16` is lossless up to
the limits shown. `quant8` maps values to 256 equally-spaced buckets, introducing
quantization error proportional to the data range.

## Worked Examples

### Example 1: Sparse Matrix Round-Trip

We write the MovieLens ratings matrix to StreamPress and verify lossless
recovery.

```{r sparse-roundtrip}
data(movielens)

# Write to StreamPress
spz_file <- tempfile(fileext = ".spz")
stats <- st_write(movielens, spz_file, include_transpose = FALSE)

# Inspect the file
info <- st_info(spz_file)

# Read back
ml_back <- st_read(spz_file)

# Compare file sizes
rds_file <- tempfile(fileext = ".rds")
saveRDS(movielens, rds_file)

size_df <- data.frame(
  Format = c("R object (in memory)", ".rds file", ".spz file"),
  `Size (KB)` = c(
    round(as.numeric(object.size(movielens)) / 1024, 1),
    round(file.info(rds_file)$size / 1024, 1),
    round(file.info(spz_file)$size / 1024, 1)
  ),
  check.names = FALSE
)

knitr::kable(size_df, align = "lr",
             caption = "MovieLens storage: R object vs. .rds vs. .spz")
```

```{r sparse-size-plot, fig.height = 3}
size_df$Format <- factor(size_df$Format, levels = rev(size_df$Format))
ggplot(size_df, aes(x = Format, y = `Size (KB)`, fill = Format)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("steelblue", "grey60", "grey80"), guide = "none") +
  coord_flip() +
  labs(title = "MovieLens Storage Comparison", x = "", y = "Size (KB)") +
  theme_minimal()
```

```{r verify-sparse}
# Verify values are identical (dimnames may differ since SPZ doesn't store them)
cat("Values identical:", identical(movielens@x, ml_back@x), "\n")
cat("Structure identical:", identical(movielens@i, ml_back@i) &&
      identical(movielens@p, ml_back@p), "\n")
```

The `.spz` format preserves all numeric values and the CSC structure exactly.
Dimension names (row/column labels) are not stored in the `.spz` file itself;
use R attributes or sidecar metadata for these.

### Example 2: Dense Matrix Round-Trip

StreamPress v3 handles dense matrices with optional compression codecs.

```{r dense-roundtrip}
data(aml)

# Write dense matrix
dense_file <- tempfile(fileext = ".spz")
st_write_dense(aml, dense_file)

# Read back
aml_back <- st_read_dense(dense_file)

# Compare sizes
rds_dense <- tempfile(fileext = ".rds")
saveRDS(aml, rds_dense)

dense_df <- data.frame(
  Format = c("R object (in memory)", ".rds file", ".spz v3 file"),
  `Size (KB)` = c(
    round(as.numeric(object.size(aml)) / 1024, 1),
    round(file.info(rds_dense)$size / 1024, 1),
    round(file.info(dense_file)$size / 1024, 1)
  ),
  check.names = FALSE
)
knitr::kable(dense_df, align = "lr",
             caption = "AML dense matrix storage comparison")
```

```{r verify-dense}
cat("Max absolute difference:", max(abs(aml - aml_back)), "\n")
```

### Example 3: Precision Mode Comparison

Different precision modes trade file size for numerical accuracy. We write
the MovieLens data at each precision level and measure the error introduced.

```{r precision-compare}
precisions <- c("fp64", "fp32", "fp16", "quant8")
prec_results <- data.frame(
  Precision = precisions,
  `File Size (KB)` = NA_real_,
  `Max Abs Error` = NA_real_,
  Lossless = NA_character_,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

for (i in seq_along(precisions)) {
  f <- tempfile(fileext = ".spz")
  st_write(movielens, f, precision = precisions[i], include_transpose = FALSE)
  back <- st_read(f)
  err <- max(abs(movielens@x - back@x))
  prec_results$`File Size (KB)`[i] <- round(file.info(f)$size / 1024, 1)
  prec_results$`Max Abs Error`[i] <- round(err, 6)
  prec_results$Lossless[i] <- if (err == 0) "Yes" else "No"
}

knitr::kable(prec_results, align = "lrrr",
             caption = "Precision vs. compression tradeoff (MovieLens ratings)")
```

For integer rating data (values 1–5), `fp64`, `fp32`, and `fp16` are all lossless
because the values fit within the exact representation range of each format.
Only `quant8` introduces quantization error, mapping the continuous value range
into 256 discrete buckets.

### Example 4: Real-World Dataset — pbmc3k Single-Cell RNA-seq

The `pbmc3k` dataset ships as StreamPress-compressed raw bytes — a representative
subset (8,000 genes × 500 cells) of the 10x Genomics PBMC 3k dataset, demonstrating
how SPZ enables shipping sparse datasets within CRAN's tarball size limits.

```{r pbmc3k-demo}
# Load compressed bytes (not a matrix — raw bytes)
data(pbmc3k, package = "RcppML")

# Write to temp file and inspect
pbmc3k_file <- tempfile(fileext = ".spz")
writeBin(pbmc3k, pbmc3k_file)
pbmc_info <- st_info(pbmc3k_file)

pbmc_df <- data.frame(
  Property = c("Rows (genes)", "Columns (cells)", "Non-zeros",
               "Density", "SPZ file size", "Compression ratio"),
  Value = c(
    format(pbmc_info$rows, big.mark = ","),
    format(pbmc_info$cols, big.mark = ","),
    format(pbmc_info$nnz, big.mark = ","),
    paste0(round(pbmc_info$density_pct, 1), "%"),
    paste0(round(pbmc_info$file_bytes / 1024, 0), " KB"),
    paste0(round(pbmc_info$ratio, 1), "x")
  ),
  stringsAsFactors = FALSE
)
knitr::kable(pbmc_df, align = "lr", caption = "pbmc3k StreamPress file summary")
```

```{r pbmc3k-nmf}
# Load into R and run a quick NMF on a subset
counts <- st_read(pbmc3k_file)

# Subset for speed: 500 most variable genes × all cells
# Compute row variance efficiently: Var = E[x^2] - E[x]^2
n <- ncol(counts)
row_means <- Matrix::rowMeans(counts)
row_sq_means <- Matrix::rowMeans(counts^2)
gene_var <- row_sq_means - row_means^2
top_genes <- order(gene_var, decreasing = TRUE)[1:500]
counts_sub <- counts[top_genes, ]

model <- nmf(counts_sub, k = 5, seed = 42, tol = 1e-4, maxit = 50,
             verbose = FALSE)
cat("NMF complete: k =", ncol(model@w), ", loss =",
    round(model@misc$loss, 2), "\n")
```

### Example 5: Out-of-Core Streaming NMF (Conceptual)

For datasets too large to fit in memory, StreamPress enables streaming NMF
that processes the matrix in chunks directly from disk:

```{r streaming, eval = FALSE}
# Stream NMF from a large .spz file
# The matrix is never fully loaded into RAM
model <- nmf("/path/to/huge_matrix.spz", k = 20, streaming = TRUE,
             maxit = 50, seed = 42)
```

Streaming reads chunks of columns from the `.spz` file, updates the factor
matrices incrementally, and discards each chunk before loading the next.
This enables NMF on datasets that are 10–100× larger than available RAM.

## When StreamPress Wins

StreamPress is **always** a compression win — roughly 5–10× smaller on disk
than a raw float32 CSC binary on typical scRNA-seq count data (5% density,
38 k genes × 278 k cells benchmark: **5.36× compression**, 828 MB vs 4,434 MB).
More importantly, SPZ is also **faster to read** than raw CSC binary, even
single-threaded.

### Why SPZ Beats Raw CSC Even at 1 Thread

Reading a large sparse matrix has three costs:

| Step | Raw float32 CSC | SPZ |
|:-----|----------------:|----:|
| 1. Bytes off disk | 1.0 s (4,434 MB) | 0.2 s (828 MB) |
| 2. Decompress / decode | — | 92.9 s @ 1 thread |
| 3. Sparse object construction | 90.8 s | included in step 2 |
| **Total @ 1 thread** | **116.6 s** | **93.1 s** |

Benchmarked on a 38,606 × 278,676 matrix (554 M NNZ) using HPC local storage
at ~4,000 MB/s bandwidth. Raw CSC spends 78% of its time in step 3 —
`sparseMatrix()` must sort indices, validate structure, and allocate R objects
for 554 M non-zeros.  SPZ's parallel chunk decoder does the equivalent work
faster because chunks are independent and require no global sort.

### Parallel Scaling

With 5,465 independent chunks (51 cols/chunk at 38 k rows), SPZ scales
efficently up to ~32 threads:

| Threads | SPZ total | vs raw CSC (116.6 s) |
|--------:|----------:|---------------------:|
| 1 | 93.1 s | **1.25×** |
| 4 | 38.4 s | **3.04×** |
| 8 | 31.0 s | **3.76×** |
| 16 | 25.5 s | **4.57×** |
| 32 | 22.4 s | **5.20×** |
| 40 | 26.9 s | 4.33× *(OMP overhead)* |

The optimal thread count is around 32 for this matrix size; beyond that,
thread creation and cache contention on a shared node reduce efficiency.
The 5,465 chunks provide far more parallel work than any reasonable core
count, so SPZ scaling is limited purely by OMP overhead, not by chunk
starvation.

### The I/O Bandwidth Picture

For understanding the pure IO vs decode trade-off (relevant when comparing
to formats with cheap or no reconstruction cost, e.g., a pre-built mmap'd
binary):

| Component | Measured rate |
|:----------|:-------------|
| Single-thread SPZ decode | ~9 MB/s compressed input |
| IO bandwidth (hot cache / local) | ~4,000 MB/s |
| IO bandwidth (cold NFS) | ~500 MB/s |

On cold NFS (500 MB/s), even 4-thread SPZ comfortably beats the equivalent
raw binary read in total wall-clock time.  On hot-cache local storage (
4,000 MB/s), raw IO is bounded by `sparseMatrix()` construction, which SPZ
also avoids.

### Summary: When to Use SPZ

| Scenario | Recommendation |
|:---------|:--------------|
| Any large matrix (≥ 100 k columns) | **Always use SPZ** — wins at ≥ 1 thread |
| Many small sample files (< 20 MB each) | Parallelise *across* files with `mclapply` |
| Cold NFS, streaming NMF | **Use SPZ** — 5× less data to transfer |
| Long-term archive / CRAN distribution | **Use SPZ** — 5–10× smaller files |

### Thread Guidance

```r
# Default: auto-selects 1 thread for files < 50 MB, all cores for larger files
mat <- st_read("sample.spz")

# Large file: explicitly use 32 threads for best throughput
mat <- st_read("large_atlas.spz", threads = 32L)

# Batch-load 100 small samples: parallelise across files, 1 thread each
# This avoids OMP thread contention and saturates the CPU more efficiently
library(parallel)
mats <- mclapply(spz_paths, st_read, threads = 1L, mc.cores = 32L)
```

For batch loading of many small samples, parallelising **across** files
(one thread per file, many files concurrently) is more efficient than
parallelising within each file, because intra-file parallelism offers
limited speedup when files have few chunks.

## Key Takeaways

1. **StreamPress achieves 5–10× compression** over raw float32 CSC binary on
   typical scRNA-seq count data (5.36× measured on a 38 k × 279 k matrix).
2. **SPZ is faster to read than raw CSC binary, even single-threaded.**  The
   dominant cost of loading a large sparse matrix is not disk I/O but sparse
   object construction (sorting, validation, R/Eigen object allocation for
   hundreds of millions of non-zeros).  SPZ's parallel chunk decoder performs
   the equivalent construction in parallel without a global sort step.
3. **Parallel scaling is real and substantial.** With `chunk_bytes = 8e6`
   (the default), a 38 k-row matrix produces ~5,400 chunks across 279 k
   columns — far more than any available core count.  Speedup is 5.2× vs raw
   CSC at 32 threads on the benchmark hardware.
4. **Default `chunk_bytes = 8e6`** (8 MB) produces ~50 columns per chunk for
   typical scRNA-seq matrices.  Do not increase this beyond 64 MB: too few
   chunks concentrates work and kills OMP parallelism.
5. **`threads = NULL`** (the default) auto-selects 1 thread for files < 50 MB
   and all available threads for larger files.  For batch loading of many small
   files, parallelise across files with `mclapply(..., mc.cores = N)` and pass
   `threads = 1L` to each `st_read` call.
6. **Precision modes** offer a clear tradeoff: `fp64` for lossless preservation,
   `fp32`/`fp16` for lossless on integer data, `quant8` for maximum compression
   with controlled error.
7. **Streaming NMF** from `.spz` files enables analysis of datasets larger than
   available RAM. The streaming path reads chunks of columns, performs NMF updates
   on each chunk, and discards it before loading the next — peak memory usage is
   proportional to chunk size, not total matrix size.

*See the [GPU Acceleration](gpu-acceleration.html) vignette for GPU-direct
reads from `.spz`, and the [NMF Fundamentals](nmf-fundamentals.html) vignette for
fitting NMF models on loaded data.*

